---
title: "Individual Report"
author: "Evelyn Rodriguez"
editor:
    mode: source
format:
    html:
        toc: true
        code-fold: true
        code-summary: "Display Code"
        code-overflow: wrap
        code-block-bg: true
        code-block-border-left: "#6610f2"
        highlight-style: "a11y"
---

```{r setup, include=FALSE}
# ---- Libraries ----
library(tidyverse)
library(arrow)
library(lubridate)
library(scales)
library(gt)
library(plotly)
library(leaflet)
library(sf)
library(patchwork)
library(glue)
library(broom)
```

## Introduction

<!-- 
WRITING NOTES:
- Introduce the research question: does adverse weather in NYC affect 311 
  service request resolution times?
- Briefly describe the two datasets being used
- State the time period: January 1, 2022 to December 31, 2025
- Mention the five boroughs as the geographic scope
- End with a brief roadmap of what the report covers
-->

## Data Sources

<!-- 
WRITING NOTES:
- Describe NYC 311 Open Data (NYC Open Data / Socrata API)
  * What it is, who maintains it, what it contains
  * Link to the dataset: https://data.cityofnewyork.us/resource/erm2-nwe9.json
  * Note that it covers all 311 service requests from 2020 to present
  * Explain we filtered to Closed requests between 2022-2025
  
- Describe Open-Meteo Historical Archive API
  * What it is, that it is free and requires no API key
  * That it uses ERA5 reanalysis model combining satellite, station, and atmospheric data
  * Link: https://archive-api.open-meteo.com/v1/archive
  * Note the variables we pulled: temp max/min/avg, precipitation, snowfall, wind speed
  * Note American units: °F, inches, mph
  
- Describe the adverse/extreme weather classification system
  * Table of thresholds with sources (NWS, NYC OEM)
  * Explain the Normal/Adverse/Extreme categorization logic
-->

## Data Acquisition & Preparation

<!-- 
WRITING NOTES:
- Explain the fetch_data.R script approach and why data acquisition 
  lives in a separate script rather than inside the report
- Describe the Socrata API pagination approach (50,000 rows per page)
  and why this was necessary given the dataset size
- Explain the parquet caching strategy and why it was chosen over CSV
- Document key decisions made during cleaning:
  * Filtering to Closed status only
  * Resolution time calculation (closed_date - created_date in hours)
  * Resolution time bounds (0 to 17,520 hours / 2 years)
  * Borough standardization to title case
  * NA assignment for Unspecified borough records
- Explain the borough-level weather approach:
  * Why one coordinate per borough rather than a single city-wide point
  * The representative locations chosen for each borough
  * The join strategy (borough + date)
  * Future nearest-neighbor upgrade path
- Note the columns dropped from the original dataset and why
  (additional_details and location_type not available via JSON API)
  
- The Socrata API and pagination approach
- The parquet caching strategy
- Filtering decisions (Closed status, date range)
- Resolution time calculation and bounds
- Borough standardization and NA handling
- Borough-level weather approach and join strategy
-->

```{r data-acquisition}

## Read the pre-processed parquet files produced by fetch_data.R

# Weather data: 23,376 rows (16 points across 5 boroughs x 1,461 days)
weather <- read_parquet("data/individual/weather/weather_clean.parquet")

# Clean 311 data: ~13M rows, may take a moment to load
nyc_311 <- read_parquet("data/individual/nyc_311_clean.parquet")

# Joined dataset: 311 requests with weather data
nyc_311_with_weather <- read_parquet("data/individual/nyc_311_with_weather.parquet")

```

```{r data-checks, include = FALSE}

# Validate that data loaded correctly
glimpse(nyc_311)
glimpse(weather)
glimpse(nyc_311_with_weather)

# ---- Date ranges ----
cat("311 date range:", format(min(nyc_311$date)), "to", format(max(nyc_311$date)), "\n")
cat("Weather date range:", format(min(weather$date)), "to", format(max(weather$date)), "\n")
cat("Weather points:", n_distinct(weather$point_id), "\n")
cat("311 records:", scales::comma(nrow(nyc_311)), "\n")
cat("Matched records:", scales::comma(sum(!is.na(nyc_311_with_weather$weather_category))), "\n")

# ---- Borough coverage ----
nyc_311 |> count(borough, sort = TRUE)
weather |> count(borough)

# ---- Weather points ----
# All 16 weather points present with equal day counts
weather |> count(point_id)

# ---- Match rate ----
nyc_311_with_weather |>
  summarise(
    matched     = sum(!is.na(weather_category)),
    unmatched   = sum(is.na(weather_category)),
    total       = n(),
    pct_matched = mean(!is.na(weather_category)) * 100
  )

# ---- Resolution time bounds ----
# Confirm resolution times are within expected range (0 to 17,520 hours)
nyc_311 |>
  summarise(
    min_hrs    = min(resolution_hrs),
    max_hrs    = max(resolution_hrs),
    median_hrs = median(resolution_hrs)
  )

# ---- Missing coordinates ----
nyc_311 |>
  summarise(
    has_coords     = sum(!is.na(latitude) & !is.na(longitude)),
    missing_coords = sum(is.na(latitude) | is.na(longitude)),
    pct_missing    = mean(is.na(latitude) | is.na(longitude)) * 100
  )

n_total <- nrow(nyc_311)
n_missing_coords <- sum(is.na(nyc_311$latitude) | is.na(nyc_311$longitude))
pct_missing_coords <- round(mean(is.na(nyc_311$latitude) | is.na(nyc_311$longitude)) * 100, 2)
n_matched_weather <- sum(!is.na(nyc_311_with_weather$weather_category))
n_unmatched_weather <- sum(is.na(nyc_311_with_weather$weather_category))
pct_matched_weather <- round(mean(!is.na(nyc_311_with_weather$weather_category)) * 100, 2)
date_start_311     <- format(min(nyc_311$date))
date_end_311       <- format(max(nyc_311$date))
median_res_hrs     <- round(median(nyc_311$resolution_hrs), 2)
n_weather_points   <- n_distinct(weather$point_id)
n_weather_rows     <- nrow(weather)

```

**Snapshot of Data Acquired**

* Of the `r comma(n_total)` records, `r comma(n_missing_coords)` (`r pct_missing_coords`%) lacked geographic coordinates and could not be matched to a weather point.

* The 311 dataset spans `r date_start_311` to `r date_end_311`, containing `r comma(n_total)` closed service requests.

* Of all records, `r comma(n_matched_weather)` (`r pct_matched_weather`%) were successfully matched to a weather point.

* The median resolution time across all requests was `r median_res_hrs` hours.

**Data Acquisition**

Created a separate script that is run once manually to download and cache all data into parquet files. Due to high volume of 311 data (~13 million rows), the initial fetch takes around 40 minutes and caching helps prevent running and downloading this data from square one every time a change is made. The parquet files produced sit on disk and this report will just read them in order to perform analysis. 

I chose to use parquet over CSV because CSV files would process slowly and be memory-inefficient as it offers no compression. Parquet, a binary file format designed specifically for large tabular data sets, is a better choice for this particular project. Parquet compression makes file sizes significantly smaller and it preserves data types directly provided by the source API, which means the 13 million row data set loads much faster and uses significantly less memory.

Every time fetch_data.R runs, before making any API call it checks whether the output file already exists on disk.

If the file is already there, the function exits immediately and skips the download entirely. If it isn't there, it proceeds with the fetch.
This applies at two levels:

Page level — each of the 264+ individual API pages is saved immediately as its own parquet file the moment it downloads. If the session crashes on page 200, re-running picks up from page 201 rather than starting over
File level — once the final clean parquet files exist, the entire fetch is skipped on every subsequent run

the report can be rendered repeatedly and quickly — it just reads the finished parquet files from disk. The expensive data acquisition work only ever happens once, and the results are preserved indefinitely until you deliberately delete the cached files to force a refresh.  persistent cache decouples data acquisition from report rendering

Note on Excluded Records after Join: Joining on if there is a borough and lat/long coords 

193,283 records are missing both lat and lon — these are the records with no coordinates at all. Lat and lon are always missing together, never one without the other, which makes sense — if a request has no location it has neither coordinate.
196,803 unmatched in the joined dataset — the difference of 3,520 between this and the 193,283 missing coordinates accounts for records that have coordinates but whose borough was NA, so they were excluded from the spatial lookup in Step 1.
1.46% of all records have no coordinates — a small but meaningful fraction. These are typically requests submitted by phone where the caller didn't provide a specific address, or agency-generated requests without a precise location.

13,002,701 records (98.54%) have valid coordinates and can be spatially matched
193,283 records (1.46%) have no coordinates and will receive NA weather fields
13,195,984 total — all records accounted for

phone-submitted requests often don't have precise location data



```{r data-prep-summary}
# Summary statistics of the cleaned datasets to validate acquisition
# This chunk will display a brief summary table confirming row counts,
# date ranges, and borough coverage for both datasets
```

## EDA: NYC 311 Requests & Weather in NYC

### Limitations in Analysis

**Data Limitations**
* 193,283 records (1.46%) missing coordinates could not be spatially matched to weather data and receive NA weather fields

* 16 weather coordinate points represent borough-level conditions but cannot capture hyper-local microclimate variation within boroughs

* Weather data is daily resolution — hourly weather events like afternoon thunderstorms or overnight cold snaps are averaged into daily summaries, potentially obscuring short-duration but high-impact weather events

* Open-Meteo uses ERA5 reanalysis model rather than direct station readings — modeled estimates rather than observed measurements

**Methodological Limitations**
* Resolution time is used as a proxy for service quality — a closed request does not necessarily mean the issue was resolved satisfactorily

* Sub-1-minute closures (316,472 records / 2.4%) likely represent automated or duplicate closures rather than genuine resolutions — their inclusion may affect agency-level averages

* temp_min_f used for cold weather classification captures overnight lows which drive infrastructure stress but may not reflect daytime conditions experienced by residents

* One weather coordinate per borough subregion — future nearest-neighbor upgrade with denser coordinates would improve spatial precision

* Analysis is correlational — weather conditions are associated with resolution time changes but causation cannot be established from observational data alone

**Scope Limitations**
* Four year window (2022-2025) may not capture longer-term seasonal patterns or anomalous weather years

* Borough is the finest geographic unit for weather matching — community district and neighborhood level variation in both weather and service delivery is not captured



### Weather Patterns in NYC (2022–2025)

<!-- 
WRITING NOTES:
- This is the EDA section for the weather dataset
- Guide the reader through what NYC weather actually looked like 
  over this four-year period before connecting it to 311 data
-->

#### Adverse and Extreme Weather Days by Borough

<!-- 
WRITING NOTES:
- How many Normal, Adverse, and Extreme days occurred per borough?
- Are there meaningful differences between boroughs in adverse day counts?
- Note: Bronx max wind = exactly 31mph (our threshold) vs coastal boroughs
  which saw higher gusts — worth mentioning
-->

```{r weather-category-by-borough}
# Bar chart or table: count of Normal/Adverse/Extreme days per borough
# Use plotly for interactivity

# Count of days per weather category per borough
# Aggregated to borough level using the most common weather category
# across all points within the borough on each day
weather_category_by_borough <- weather |>
  group_by(borough, date) |>
  summarise(
    # Most common category across all points in the borough that day
    weather_category = names(which.max(table(weather_category))),
    .groups = "drop"
  ) |>
  mutate(
    weather_category = factor(weather_category,
                              levels = c("Normal", "Adverse", "Extreme"),
                              ordered = TRUE)
  ) |>
  count(borough, weather_category)

weather_category_by_borough

# Stacked bar chart: count of Normal/Adverse/Extreme days by borough
weather_conditions_by_borough <- ggplot(weather_category_by_borough,
       aes(x = borough, y = n, fill = weather_category)) +
  geom_col(width = 0.6) +
  # Labels for Normal and Adverse — centered inside their segments
  geom_text(
    data     = weather_category_by_borough |> 
                 filter(weather_category != "Extreme"),
    aes(label = comma(n)),
    position = position_stack(vjust = 0.5),
    size     = 3,
    color    = "white",
    fontface = "bold"
  ) +
  # Label for Extreme — positioned above the full bar
  geom_text(
    data     = weather_category_by_borough |> 
                 filter(weather_category == "Extreme"),
    aes(label = paste0("Extreme: ", comma(n)),
        y     = 0),
    vjust    = 1.0,
    size     = 2.5,
    color    = "#e74c3c",
    fontface = "bold"
  ) +
  scale_fill_manual(
    values = c(Normal  = "#2ecc71",
               Adverse = "#f39c12",
               Extreme = "#e74c3c"),
    name = "Weather Category"
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Weather Conditions by Borough (2022-2025)",
    subtitle = "Number of days classified as Normal, Adverse, or Extreme",
    x        = NULL,
    y        = "Number of Days"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "gray40")
  )

weather_conditions_by_borough


```

Normal weather days dominate across all 5 boroughs as expected. The Bronx and Manhattan experienced the most Adverse days and Staten Island experienced the fewest. Extreme days are rare. 

#### Temperature Trends

<!-- 
WRITING NOTES:
- Show temperature patterns over the four-year period
- Highlight notable heat waves and cold snaps
- Reference the max temperatures observed (103-106°F) and 
  min temperatures (1-4.5°F)
-->

```{r temperature-trends}
# Time series: daily avg temperature by borough over 2022-2025
# Annotate notable extreme heat and cold periods


# Aggregate daily temperature to borough level by averaging across all
# weather points within each borough, then compute weekly averages
weather_temp_weekly <- weather |>
  # Step 1: average across points within each borough for each day
  group_by(borough, date) |>
  summarise(
    daily_avg_temp = mean(temp_avg_f, na.rm = TRUE),
    .groups = "drop"
  ) |>
  # Step 2: aggregate daily values to weekly averages
  # floor_date() rounds down to the nearest Sunday for each date
  mutate(week = floor_date(date, "week")) |>
  group_by(borough, week) |>
  summarise(
    weekly_avg_temp = mean(daily_avg_temp, na.rm = TRUE),
    .groups = "drop"
  ) |>
  # Step 3: extract year from week for faceting by year
  # filter removes the 2021-12-26 week that floor_date() creates
  # when rounding down dates in the first week of January 2022
  mutate(year = year(week)) |>
  filter(year >= 2022)

# Line chart: weekly average temperature by borough faceted by year
# Each panel shows one year allowing clear comparison of seasonal patterns
weather_temp_trends <- ggplot(
    weather_temp_weekly,
    aes(x = week, y = weekly_avg_temp, color = borough)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  scale_color_brewer(palette = "Set1", name = "Borough") +
  # Show month abbreviations on x axis — free_x allows each facet
  # to have its own x axis scaled to that year's date range
  scale_x_date(
    date_breaks = "2 months",
    date_labels = "%b"
  ) +
  scale_y_continuous(labels = \(x) paste0(x, "°F")) +
  # Arrange four years in a 2x2 grid
  facet_wrap(~ year, scales = "free_x", ncol = 2) +
  labs(
    title    = "Weekly Average Temperature by Borough (2022-2025)",
    subtitle = "Average of daily mean temperatures across all weather points per borough",
    x        = NULL,
    y        = "Average Temperature (°F)"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "gray40"),
    axis.text.x     = element_text(angle = 45, hjust = 1)
  )

weather_temp_trends


```

The near-identical lines show that temperature is consistent across boroughs with no major variability across boroughs as far as average temperature goes. 


#### Precipitation and Snowfall

<!-- 
WRITING NOTES:
- Show precipitation and snowfall patterns over the period
- Total annual precipitation ~50 inches/year — consistent with NYC historical avg
- Note seasonal patterns and any notable storms
-->

```{r precipitation-trends}
# Time series or bar chart: monthly precipitation and snowfall totals
# Consider showing by borough to highlight coastal vs inland differences

# Check variability of precipitation and snowfall across boroughs
weather |>
  group_by(borough) |>
  summarise(
    mean_precip  = mean(precip_in,   na.rm = TRUE),
    mean_snow    = mean(snowfall_in, na.rm = TRUE),
    total_precip = sum(precip_in,    na.rm = TRUE),
    total_snow   = sum(snowfall_in,  na.rm = TRUE)
  )

# Aggregate daily precipitation and snowfall to NYC-wide monthly totals
# Step 1: average across all 16 weather points for each day to get a single city-wide daily value, correcting for Queens having 4 points vs 3 for other boroughs
# Step 2: sum daily values to monthly totals

weather_precip_monthly <- weather |>
  # Step 1: average across all points for each day
  group_by(date) |>
  summarise(
    daily_precip = mean(precip_in,   na.rm = TRUE),
    daily_snow   = mean(snowfall_in, na.rm = TRUE),
    .groups      = "drop"
  ) |>
  # Step 2: aggregate to monthly totals
  mutate(month = floor_date(date, "month")) |>
  group_by(month) |>
  summarise(
    total_precip = sum(daily_precip, na.rm = TRUE),
    total_snow   = sum(daily_snow,   na.rm = TRUE),
    .groups      = "drop"
  )


# Pivot to long format for faceted visualization
# This allows ggplot to facet by variable (precipitation vs snowfall)
# while sharing the same x axis
weather_precip_long <- weather_precip_monthly |>
  pivot_longer(
    cols      = c(total_precip, total_snow),
    names_to  = "variable",
    values_to = "inches"
  ) |>
  mutate(
    # Clean up variable labels for facet titles
    variable = recode(variable,
                      total_precip = "Precipitation (in)",
                      total_snow   = "Snowfall (in)")
  )


# Stacked panel chart: monthly precipitation and snowfall (2022-2025)
# Two panels share the same x axis — precipitation on top, snowfall on bottom
weather_precip_snow <- ggplot(weather_precip_long,
       aes(x = month, y = inches, fill = variable)) +
  geom_col(width = 25, show.legend = FALSE) +
  scale_fill_manual(
    values = c(
      "Precipitation (in)" = "#3498db",
      "Snowfall (in)"      = "#85c1e9"
    )
  ) +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b %Y"
  ) +
  scale_y_continuous(labels = \(x) paste0(x, '"')) +
  facet_wrap(~ variable, ncol = 1, scales = "free_y") +
  labs(
    title    = "Monthly Precipitation and Snowfall in NYC (2022-2025)",
    subtitle = "NYC-wide average across all 16 weather points",
    x        = NULL,
    y        = "Inches"
  ) +
  theme_bw() +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40"),
    axis.text.x   = element_text(angle = 45, hjust = 1),
    strip.text    = element_text(face = "bold")
  )

weather_precip_snow

```

**Precipitation:**

* October 2022 stands out as an unusually wet month - storm/hurricane?

* 2023 was the wettest year across all boroughs

* 2025 appears to be the driest (though it may be incomplete depending on when the data ends)

* Borough differences are minimal — roughly 1-2 inches variation in any given year

**Snowfall:**

* 2023 had dramatically less snow than other years across all boroughs

* Staten Island consistently gets less snow than the other boroughs

* 2022 and 2025 were the snowiest years (January 2022 had the highest snowfall of the entire period and Late 2025 shows a notable snowfall spike earlier in year)

#### Weather Event Calendar

<!-- 
WRITING NOTES:
- A visual calendar or heatmap showing weather category by day
- Helps reader see clustering of adverse/extreme events across the year
- Could be a useful reference for the analysis section
-->

```{r weather-calendar}
# Heatmap: weather category by day of year, faceted by year
# Shows seasonality of adverse/extreme weather events

# Aggregate daily weather categories to monthly proportions per borough
# For each borough-month combination, calculate the proportion of days
# that were classified as Adverse or Extreme
weather_heatmap <- weather |>
  # Step 1: get one category per borough per day using most common method
  group_by(borough, date) |>
  summarise(
    weather_category = names(which.max(table(weather_category))),
    .groups = "drop"
  ) |>
  # Step 2: aggregate to monthly proportions per borough
  mutate(month = floor_date(date, "month")) |>
  group_by(borough, month) |>
  summarise(
    n_days          = n(),
    n_adverse       = sum(weather_category == "Adverse"),
    n_extreme       = sum(weather_category == "Extreme"),
    pct_adverse     = mean(weather_category %in% c("Adverse", "Extreme")) * 100,
    .groups         = "drop"
  )

# # Monthly heatmap: proportion of adverse/extreme days by borough
# weather_adverse_heatmap <- ggplot(weather_heatmap,
#        aes(x = month, y = borough, fill = pct_adverse)) +
#   geom_tile(color = "white", linewidth = 0.3) +
#   scale_fill_gradient(
#     low      = "#ffffff",
#     high     = "#e74c3c",
#     name     = "% Adverse/\nExtreme Days"
#   ) +
#   scale_x_date(
#     date_breaks = "3 months",
#     date_labels = "%b %Y"
#   ) +
#   labs(
#     title    = "Monthly Adverse & Extreme Weather Days by Borough (2022-2025)",
#     subtitle = "Percentage of days per month classified as Adverse or Extreme",
#     x        = NULL,
#     y        = NULL
#   ) +
#   theme_bw() +
#   theme(
#     plot.title    = element_text(face = "bold"),
#     plot.subtitle = element_text(color = "gray40"),
#     axis.text.x   = element_text(angle = 45, hjust = 1),
#     legend.position = "right"
#   )
# 
# weather_adverse_heatmap

# Monthly heatmap: proportion of adverse/extreme days by borough
weather_adverse_heatmap <- ggplot(weather_heatmap,
       aes(x = month, y = borough, fill = pct_adverse,
           text = paste0(
             "Borough: ", borough, "\n",
             "Month: ", format(month, "%B %Y"), "\n",
             "Adverse Days: ", n_adverse, "\n",
             "Extreme Days: ", n_extreme, "\n",
             "Total Days: ", n_days, "\n",
             "% Adverse/Extreme: ", round(pct_adverse, 1), "%"
           ))) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient(
    low  = "#ffffff",
    high = "#e74c3c",
    name = "% Adverse/\nExtreme Days"
  ) +
  scale_x_date(
    date_breaks = "3 months",
    date_labels = "%b %Y"
  ) +
  labs(
    title    = "Monthly Adverse & Extreme Weather Days by Borough (2022-2025)",
    subtitle = "Percentage of days per month classified as Adverse or Extreme",
    x        = NULL,
    y        = NULL
  ) +
  theme_bw() +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "gray40"),
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

# Convert to interactive plotly chart with custom tooltip
ggplotly(weather_adverse_heatmap, tooltip = "text")


```

Winter months (Jan/Feb) show the darkest reds across all boroughs — high proportion of adverse/extreme days driven by cold temperatures

Summer months show lighter colors with some mid-range reds in July/August — heat events

January 2022 Bronx shows 77% adverse/extreme days — a very cold January which matches our earlier findings

July and August 2022 show 32% adverse days — likely heat events

The pattern is consistent across boroughs which reinforces our earlier finding that weather conditions don't vary dramatically between boroughs


### 311 Service Request Overview

<!-- 
WRITING NOTES:
- This is the EDA section for the 311 dataset
- Give the reader a full picture of the 311 data before the analysis
-->

#### Request Volume

<!-- 
WRITING NOTES:
- Total requests by year and borough
- Are there year-over-year trends in volume?
- Borough distribution: Brooklyn > Queens > Bronx > Manhattan > Staten Island
-->

```{r request-volume}
# Bar chart: total requests by borough and year
# Use plotly for interactivity

# Count of 311 requests by year
requests_by_year <- nyc_311 |>
  mutate(year = year(date)) |>
  count(year)

# Line chart: total 311 service requests by year
requests_by_year_chart <- ggplot(requests_by_year,
       aes(x = year, y = n)) +
  geom_line(color = "#3498db", linewidth = 1) +
  geom_point(color = "#3498db", size = 3) +
  geom_text(
    aes(label = comma(n),
            hjust = ifelse(year == 2022, -0.1, 0.5)),
    vjust    = -1,
    size     = 3.5,
    fontface = "bold"
  ) +
  scale_x_continuous(breaks = c(2022, 2023, 2024, 2025)) +
  scale_y_continuous(
    labels = comma,
    expand = expansion(mult = c(0.05, 0.1))
  ) +
  labs(
    title    = "NYC 311 Service Requests by Year (2022-2025)",
    subtitle = "Total closed service requests per year",
    x        = NULL,
    y        = "Number of Requests"
  ) +
  theme_bw() +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40")
  )

requests_by_year_chart

requests_by_year_borough <- nyc_311 |>
  mutate(year = year(date)) |>
  count(year, borough) |>
  filter(!is.na(borough))


# Net growth of 311 requests by borough from 2022 to 2025
requests_growth <- requests_by_year_borough |>
  filter(year %in% c(2022, 2025)) |>
  pivot_wider(names_from = year, values_from = n, names_prefix = "year_") |>
  mutate(
    net_growth     = year_2025 - year_2022,
    pct_growth     = round((year_2025 - year_2022) / year_2022 * 100, 1)
  ) |>
  arrange(desc(pct_growth))

# Horizontal bar chart: net growth of 311 requests by borough (2022-2025)
requests_growth_chart <- ggplot(requests_growth,
       aes(x = pct_growth, y = reorder(borough, pct_growth))) +
  geom_col(fill = "#3498db", width = 0.6) +
  geom_text(
    aes(label = paste0(pct_growth, "%")),
    hjust    = -0.2,
    size     = 3.5,
    fontface = "bold"
  ) +
  scale_x_continuous(
    labels = \(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title    = "Net Growth in 311 Service Requests by Borough (2022-2025)",
    subtitle = "Percentage increase in closed service requests from 2022 to 2025",
    x        = "Percentage Growth",
    y        = NULL
  ) +
  theme_bw() +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40")
  )

requests_growth_chart

```

Brooklyn consistently has the highest volume of requests and has shown consistent year over year growth

Bronx dipped in 2023 and rebounded in 2024 and 2025.

Staten Island the lowest (and relatively flat across all four years)

Manhattan Shows steady but modest growth from 618,835 in 2022 to 703,589 in 2025 — about 14% increase over the period (2022 was notably lower)

Queens Shows the strongest and most consistent growth of any borough — from 703,600 in 2022 to 850,735 in 2025, a 21% increase over the period. Queens surpassed Manhattan in every year despite having a comparable population, which is interesting given that Manhattan is the most densely populated borough. The gap between Queens and Manhattan widens each year suggesting Queens residents are increasingly engaging with 311

Brooklyn, Queens, and the Bronx are all growing faster than Manhattan and Staten Island


#### Top Complaint Types and Agencies

<!-- 
WRITING NOTES:
- Top 20 complaint types by volume
- Top 10 agencies by volume
- Note NYPD handles 46% of all requests (noise, parking)
- Note HPD handles 22% (housing: HEAT/HOT WATER, plumbing)
- Flag weather-sensitive complaint types for the analysis section:
  HEAT/HOT WATER, Street Condition, Water System, Missed Collection
-->

**Top Complaints**

```{r complaint-types}
# Horizontal bar chart: top 20 complaint types
# Highlight weather-sensitive types in a different color

# Normalize complaint types to uppercase to catch all casing duplicates
# then aggregate counts and find top 2 agencies for each type
top_complaint_types <- nyc_311 |>
  mutate(complaint_type = str_to_upper(str_trim(complaint_type))) |>
  count(complaint_type, sort = TRUE) |>
  slice_head(n = 15)

# For each of the top 15 complaint types, find the top 2 responding agencies
top_complaint_agencies <- nyc_311 |>
  mutate(complaint_type = str_to_upper(str_trim(complaint_type))) |>
  filter(complaint_type %in% top_complaint_types$complaint_type) |>
  count(complaint_type, agency, sort = TRUE) |>
  group_by(complaint_type) |>
  slice_head(n = 2) |>
  summarise(
    top_agencies = paste(agency, collapse = "/"),
    .groups = "drop"
  )

# Join complaint counts with top agency labels
top_complaints_with_agencies <- top_complaint_types |>
  left_join(top_complaint_agencies, by = "complaint_type")

# Extract primary agency for color fill
top_complaints_with_agencies <- top_complaints_with_agencies |>
  mutate(
    primary_agency = str_extract(top_agencies, "^[^/]+"),
    # Reorder by count for horizontal bar chart
    complaint_type = fct_reorder(complaint_type, n)
  )

# Define color palette for agencies
agency_colors <- c(
  "NYPD" = "#2980b9",
  "HPD"  = "#e74c3c",
  "DOT"  = "#27ae60",
  "DEP"  = "#8e44ad",
  "DSNY" = "#f39c12",
  "DOB"  = "#16a085"
)

# Horizontal bar chart: top 15 complaint types colored by primary agency
# with top 2 agencies labeled on each bar
top_complaints_chart <- ggplot(top_complaints_with_agencies,
       aes(x = n, y = complaint_type, fill = primary_agency)) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = paste0(top_agencies, "  ")),
    hjust    = 1,
    size     = 3,
    color    = "white",
    fontface = "bold"
  ) +
  scale_fill_manual(values = agency_colors, name = "Primary Agency") +
  scale_x_continuous(labels = comma) +
  labs(
    title    = "Top 15 NYC 311 Complaint Types (2022-2025)",
    subtitle = "And their primary responding agency",
    x        = "Number of Requests",
    y        = NULL
  ) +
  theme_bw() +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "gray40"),
    legend.position = "bottom"
  )

top_complaints_chart

```

**Top Responding Agencies**

```{r agencies}
# Bar chart: top 10 agencies by request volume

# Count of 311 requests by agency, normalized to uppercase to catch
# any casing inconsistencies, top 10 by volume
n_total_requests <- nrow(nyc_311)

top_agencies <- nyc_311 |>
  mutate(agency_name = str_to_upper(str_trim(agency_name))) |>
  count(agency_name, sort = TRUE) |>
  slice_head(n = 10) |>
  mutate(
    # Percentage of all 13.2M requests handled by each agency
    pct_total   = round(n / n_total_requests * 100, 1),
    agency_name = fct_reorder(agency_name, n)
  )

# GT table: top 10 responding agencies by request volume
top_agencies_table <- top_agencies |>
  arrange(desc(n)) |>
  select(agency_name, n, pct_total) |>
  mutate(
    # Convert agency names from uppercase to title case for readability
    agency_name = str_to_title(str_to_lower(agency_name))
  ) |>
  gt() |>
  tab_header(
    title    = "Top 10 Responding Agencies by 311 Request Volume (2022-2025)",
    subtitle = "Total closed service requests per agency"
  ) |>
  cols_label(
    agency_name = "Agency",
    n           = "Requests",
    pct_total   = "% of Total"
  ) |>
  fmt_number(columns = n, decimals = 0) |>
  fmt_number(columns = pct_total, decimals = 1, suffix = "%") |>
  cols_align(align = "left",  columns = agency_name) |>
  cols_align(align = "right", columns = c(n, pct_total)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_fill(color = "#f2f2f2"),
    locations = cells_body(rows = seq(1, 10, by = 2))
  )

top_agencies_table

```

#### Resolution Time Distribution

<!-- 
WRITING NOTES:
- Show the overall distribution of resolution times
- Note the heavy right skew: median 7.36 hrs vs mean 323 hrs
- Explain why median is more appropriate than mean for this data
- Introduce the resolution time categories:
  * Instant: < 1 minute (< 0.017 hrs) — 316,472 records (2.4%)
  * Fast: 1 minute to 1 hour
  * Standard: 1 hour to 24 hours
  * Extended: 1 day to 30 days
  * Long-term: over 30 days (8.2% of records)
- Discuss the instant closure finding:
  * What complaint types and agencies drive them?
  * Are these automated closures or data entry patterns?
-->

```{r resolution-distribution}
# Histogram or density plot of resolution times (log scale)
# Annotate the median and key percentiles

# Resolution time distribution data
# Version 1: all records excluding exact zeros (log scale requires > 0)
resolution_dist_all <- nyc_311 |>
  filter(resolution_hrs > 0) |>
  select(resolution_hrs)

# Resolution time categories
# Applied to full dataset including instant closures
resolution_categories <- nyc_311 |>
  mutate(
    resolution_category = case_when(
      resolution_hrs <  0.017             ~ "Instant",
      resolution_hrs <  1                 ~ "Fast",
      resolution_hrs <  24                ~ "Standard",
      resolution_hrs <  720               ~ "Extended",
      TRUE                                ~ "Long-term"
    ) |> factor(levels = c("Instant", "Fast", "Standard",
                            "Extended", "Long-term"),
                ordered = TRUE)
  ) |>
  count(resolution_category) |>
  mutate(pct = round(n / sum(n) * 100, 1))


# Density plot: resolution time distribution on log scale
resolution_density <- ggplot(resolution_dist_all,
       aes(x = resolution_hrs)) +
  geom_density(
    color     = "#3498db",
    linewidth = 0.8
  ) +
  scale_x_log10(
    labels = \(x) case_when(
      x < 1    ~ paste0(round(x * 60), "m"),
      x < 24   ~ paste0(round(x), "h"),
      x < 720  ~ paste0(round(x / 24), "d"),
      TRUE     ~ paste0(round(x / 720), "mo")
    ),
    breaks = c(0.017, 0.25, 1, 6, 24, 168, 720, 4380, 17520)
  ) +
  labs(
    title   = "Distribution of 311 Service Request Resolution Times (2022-2025)",
    subtitle = "Log scale — two peaks visible at ~1 hour and ~1-2 days",
    x       = "Resolution Time",
    y       = "Density",
    caption = "Source: NYC Open Data — 311 Service Requests (2022-2025)"
  ) +
  theme_bw() +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40"),
    plot.caption  = element_text(color = "gray40")
  )

resolution_density

# GT table: percentage of requests by resolution time category
resolution_categories_table <- resolution_categories |>
  mutate(resolution_category = as.character(resolution_category)) |>
  gt() |>
  tab_header(
    title    = "311 Service Requests by Resolution Time Category (2022-2025)",
    subtitle = "Distribution of closed service requests across resolution time categories"
  ) |>
  cols_label(
    resolution_category = "Resolution Category",
    n                   = "Requests",
    pct                 = "% of Total"
  ) |>
  fmt_number(columns = n,   decimals = 0) |>
  fmt_number(columns = pct, decimals = 1, suffix = "%") |>
  cols_align(align = "left",  columns = resolution_category) |>
  cols_align(align = "right", columns = c(n, pct)) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_fill(color = "#f2f2f2"),
    locations = cells_body(rows = seq(1, 5, by = 2))
  ) |>
  tab_source_note(
    source_note = "Source: NYC Open Data — 311 Service Requests (2022-2025)"
  )

resolution_categories_table


```

**Resolution Time Categories**

* Instant: < 0.017 hrs (under 1 minute)
* Fast: 0.017 to 1 hr
* Standard: 1 to 24 hrs
* Extended: 24 to 720 hrs (1 month)
* Long-term: over 720 hrs

nearly a quarter of all requests resolved within an hour. Fast + Standard together account for 57.8% of all requests resolved within 24 hours.

two peaks, one around 1 hour and another around 24-48 hours, with a long tail extending to months. This bimodal pattern suggests two distinct types of resolution behavior in the data

Long tail: 8.2% of requests take over 30 days


```{r instant-closures}
# Table: top complaint types and agencies for sub-1-minute resolutions
# This sets up the later analysis of whether weather affects instant closure rates
```

#### Seasonal Patterns

<!-- 
WRITING NOTES:
- Do 311 request volumes or resolution times vary by season?
- Are certain complaint types more seasonal?
- This provides context for the weather analysis that follows
-->

```{r seasonal-patterns}
# Line chart: monthly request volume over 2022-2025
# Facet or color by complaint type category

# Identify the 28 complaint types that account for 80% of all requests
top_complaint_types_80 <- nyc_311 |>
  mutate(complaint_type = str_to_upper(str_trim(complaint_type))) |>
  count(complaint_type, sort = TRUE) |>
  mutate(
    pct            = n / sum(n) * 100,
    cumulative_pct = cumsum(pct)
  ) |>
  filter(cumulative_pct <= 80) |>
  pull(complaint_type)

# Compute average monthly request volume per complaint type
# averaged across all 4 years
seasonal_patterns <- nyc_311 |>
  mutate(
    complaint_type = str_to_upper(str_trim(complaint_type)),
    month_num      = month(date),
    month_name     = month(date, label = TRUE, abbr = TRUE)
  ) |>
  filter(complaint_type %in% top_complaint_types_80) |>
  count(complaint_type, month_num, month_name) |>
  # Divide by 4 to get average monthly volume across the 4 year period
  mutate(avg_monthly = n / 4)

# Heatmap: average monthly request volume by complaint type (2022-2025)
# Rows = complaint types, columns = months, color = avg monthly volume
seasonal_heatmap <- ggplot(seasonal_patterns,
       aes(x = month_name, 
           y = reorder(complaint_type, avg_monthly),
           fill = avg_monthly)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient(
    low    = "#ffffff",
    high   = "#2980b9",
    limits = c(0, 40000),
    oob    = scales::squish,
    name   = "Avg Monthly\nRequests\n(capped at 30k)"
  ) +
  scale_x_discrete(position = "top") +
  labs(
    title   = "Seasonal Patterns in NYC 311 Requests by Complaint Type",
    subtitle = "Average monthly request volume across 2022-2025",
    x        = NULL,
    y        = NULL,
    caption  = "Source: NYC Open Dat: 311 Service Requests (2022-2025)"
  ) +
  theme_bw() +
  theme(
    plot.title    = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(color = "gray40"),
    plot.caption  = element_text(color = "gray40"),
    axis.text.y   = element_text(size = 8),
    legend.position = "right"
  )

seasonal_heatmap

```

Clear seasonal patterns visible:

HEAT/HOT WATER — darkest in winter months (Jan-Mar, Nov-Dec), nearly white in summer — directly driven by cold weather heating failures
NOISE - RESIDENTIAL — slightly darker in summer months — people home more with windows open
ABANDONED VEHICLE — relatively consistent year-round with slight summer uptick
WATER SYSTEM — slight winter peak — freeze/thaw stress on pipes
MISSED COLLECTION — slight winter peak — disrupted by snow and ice

ILLEGAL PARKING
Consistently high year-round — unlike most complaint types that show clear seasonal peaks and troughs, Illegal Parking maintains a relatively uniform dark color across all 12 months. This suggests it's driven more by persistent urban conditions than seasonal factors.
Slight summer peak — there's a subtle darkening in the summer months (June-August) which could reflect increased street activity, outdoor events, or construction-related parking displacement.
No winter dip — most complaint types show some reduction in winter, likely because people are less active outdoors. Illegal Parking doesn't follow this pattern, suggesting people continue reporting parking violations regardless of weather conditions.
Why this matters for your analysis — if Illegal Parking (handled by NYPD) shows little seasonal variation, it may also show little sensitivity to adverse weather conditions. This connects directly to your working hypothesis that agencies whose work isn't directly affected by physical conditions will show less weather sensitivity. NYPD handling parking complaints is a good candidate for that hypothesis.

### Weather × Resolution Time Analysis

<!-- 
WRITING NOTES:
- This is the core analytical section of the report
- Build from simple to complex findings
-->

#### Resolution Times by Weather Category

<!-- 
WRITING NOTES:
- The central finding: do resolution times differ across Normal/Adverse/Extreme days?
- Present median resolution times with IQR by weather category
- Reference the overall pattern before breaking down by borough
-->

```{r resolution-by-weather}
# Box plot or violin plot: resolution time distribution by weather category
# Use log scale on y-axis given the skewed distribution

# Working dataset for Section 6 analysis
# Excludes 196,803 records with missing weather data
nyc_311_weather_analysis <- nyc_311_with_weather |>
  filter(!is.na(weather_category))

# Median resolution times by weather category
# Using median rather than mean due to heavy right skew in resolution times
resolution_by_weather <- nyc_311_weather_analysis |>
  group_by(weather_category) |>
  summarise(
    n_requests  = n(),
    median_hrs  = median(resolution_hrs, na.rm = TRUE),
    mean_hrs    = mean(resolution_hrs,   na.rm = TRUE),
    p25_hrs     = quantile(resolution_hrs, 0.25, na.rm = TRUE),
    p75_hrs     = quantile(resolution_hrs, 0.75, na.rm = TRUE),
    .groups     = "drop"
  )

# Bar chart: median resolution time by weather category
resolution_by_weather_chart <- ggplot(resolution_by_weather,
       aes(x = weather_category, y = median_hrs, fill = weather_category)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = paste0(round(median_hrs, 1), " hrs")),
    vjust    = -0.5,
    size     = 3.5,
    fontface = "bold"
  ) +
  scale_fill_manual(
    values = c(Normal  = "#2ecc71",
               Adverse = "#f39c12",
               Extreme = "#e74c3c"),
    name = "Weather Category"
  ) +
  scale_y_continuous(
    labels = \(x) paste0(x, " hrs"),
    expand = expansion(mult = c(0, 0.1))
  ) +
  labs(
    title    = "Median 311 Resolution Time by Weather Category (2022-2025)",
    subtitle = "Across All Complaint Types",
    x        = NULL,
    y        = "Median Resolution Time (hrs)",
    caption  = "Sources: NYC Open Data — 311 Service Requests; Open-Meteo Historical Archive API (2022-2025)"
  ) +
  theme_bw() +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "gray40"),
    plot.caption    = element_text(color = "gray40"),
    legend.position = "none"
  )

resolution_by_weather_chart

# Median resolution times by dominant weather condition
# Excludes "Clear" since that corresponds to Normal weather days
resolution_by_condition <- nyc_311_weather_analysis |>
  filter(dominant_condition != "Clear") |>
  group_by(dominant_condition, weather_category) |>
  summarise(
    n_requests = n(),
    median_hrs = median(resolution_hrs, na.rm = TRUE),
    p25_hrs    = quantile(resolution_hrs, 0.25, na.rm = TRUE),
    p75_hrs    = quantile(resolution_hrs, 0.75, na.rm = TRUE),
    .groups    = "drop"
  )

# Bar chart: median resolution time by dominant weather condition
# with reference line showing Normal day median for comparison
resolution_by_condition_chart <- ggplot(resolution_by_condition,
       aes(x = reorder(dominant_condition, median_hrs),
           y = median_hrs,
           fill = weather_category)) +
  geom_col(position = "dodge", width = 0.6) +
  # Reference line showing Normal day median for comparison
  geom_hline(
    yintercept = 6.02,
    color      = "#2ecc71",
    linewidth  = 0.8,
    linetype   = "dashed"
  ) +
  # Label for reference line
  annotate(
    "text",
    x     = 0.5,
    y     = 7,
    label = "Normal: 6.02 hrs",
    color = "#2ecc71",
    size  = 3,
    hjust = 0,
    fontface = "bold"
  ) +
  geom_text(
    aes(label = paste0(round(median_hrs, 1), " hrs")),
    position = position_dodge(width = 0.6),
    vjust    = -0.5,
    size     = 3,
    fontface = "bold"
  ) +
  scale_fill_manual(
    values = c(Adverse = "#f39c12",
               Extreme = "#e74c3c"),
    name   = "Weather Category"
  ) +
  scale_y_continuous(
    labels = \(x) paste0(x, " hrs"),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title    = "Median 311 Resolution Time by Weather Condition (2022-2025)",
    subtitle = "Dashed line shows Normal day median (6.02 hrs) as reference",
    x        = NULL,
    y        = "Median Resolution Time (hrs)",
    caption  = "Sources: NYC Open Data — 311 Service Requests; Open-Meteo Historical Archive API (2022-2025)"
  ) +
  theme_bw() +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "gray40"),
    plot.caption    = element_text(color = "gray40"),
    legend.position = "bottom"
  )

resolution_by_condition_chart

```

Mean resolution time is lower on Adverse days (301 hr) than Normal days (329 hrs) - mean is heavily influenced by long term unresolved requests and masks the real pattern. Can't rely on mean

Median resolution time increases with weather severity:

Normal: 6.02 hours
Adverse: 15.3 hours — 2.5x longer than Normal
Extreme: 20.6 hours — 3.4x longer than Normal

IQR (p25 to p75) widens with severity:

Normal: 0.87 to 82.6 hrs
Adverse: 1.07 to 94.1 hrs
Extreme: 1.30 to 89.5 hrs

Not only does the typical resolution time increase during adverse weather, but the spread of resolution times also widens — some requests get resolved faster while others take much longer, suggesting uneven agency responses during bad weather.

**Across Adverse Weather Events**
what kind of adverse weather matters most

Extreme Cold drives the longest resolution times:

Adverse cold: 24.6 hrs
Extreme cold: 25.3 hrs
This is the highest median of any condition — cold weather clearly has the biggest impact on resolution times, likely driven by HEAT/HOT WATER complaints and infrastructure failures

Extreme Heat is the only condition that resolves faster than Normal days — both Adverse (3.7 hrs) and Extreme (4.3 hrs) are well below the 6.02 hr reference line. It may reflect that heat complaints get prioritized or that NYPD closes heat-related complaints quickly

Heavy Rain shows a split:

Adverse rain: 10.2 hrs — moderate impact
Extreme rain (combined with another hazard): 22.5 hrs — much longer, suggesting compounding effects

High Wind has relatively low impact: 10.1 hrs, just under 2x Normal. High Wind only shows one bar (Adverse) since it never appears as the sole Extreme condition in the data.

Snow: 19.7 hrs — significant impact, only appears as Extreme. 

Snow and Extreme Cold are the most disruptive — both above 19 hrs, with Extreme Cold reaching 25.3 hrs — over 4x the Normal median

```{r resolution-by-weather-agency}

# Median resolution time by agency and weather category
# Includes all agencies — we can filter down later if needed
resolution_by_agency_weather <- nyc_311_weather_analysis |>
  mutate(agency = str_to_upper(str_trim(agency))) |>
  group_by(agency, weather_category) |>
  summarise(
    n_requests = n(),
    median_hrs = median(resolution_hrs, na.rm = TRUE),
    .groups    = "drop"
  )

# Split agencies into high and low volume groups
# High volume: > 100,000 requests on Normal days
# Low volume: < 100,000 requests on Normal days
# OOS and OTI excluded due to insufficient sample size

high_volume_agencies <- c("NYPD", "HPD", "DSNY", "DOT", "DEP", 
                           "DPR", "DOB", "DOHMH")

low_volume_agencies  <- c("TLC", "DCWP", "DHS", "EDC", "DOE")

# Create two filtered datasets for visualization
resolution_agency_high <- resolution_by_agency_weather |>
  filter(agency %in% high_volume_agencies)

resolution_agency_low <- resolution_by_agency_weather |>
  filter(agency %in% low_volume_agencies)

# Bar chart: median resolution time by agency and weather category
# High volume agencies
resolution_agency_high_chart <- ggplot(resolution_agency_high,
       aes(x = agency, y = median_hrs, fill = weather_category)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(
    aes(label = round(median_hrs, 1)),
    position = position_dodge(width = 0.7),
    vjust    = -0.5,
    size     = 2.5,
    fontface = "bold"
  ) +
  scale_fill_manual(
    values = c(Normal  = "#2ecc71",
               Adverse = "#f39c12",
               Extreme = "#e74c3c"),
    name   = "Weather Category"
  ) +
  scale_y_continuous(
    labels = \(x) paste0(x, " hrs"),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title    = "Median Resolution Time by Agency — High Volume (2022-2025)",
    subtitle = "Across Normal, Adverse and Extreme weather days",
    x        = NULL,
    y        = "Median Resolution Time (hrs)",
    caption  = "Sources: NYC Open Data — 311 Service Requests; Open-Meteo Historical Archive API (2022-2025)"
  ) +
  theme_bw() +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "gray40"),
    plot.caption    = element_text(color = "gray40"),
    legend.position = "bottom"
  )

# Bar chart: median resolution time by agency and weather category
# Low volume agencies
resolution_agency_low_chart <- ggplot(resolution_agency_low,
       aes(x = agency, y = median_hrs, fill = weather_category)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(
    aes(label = round(median_hrs, 1)),
    position = position_dodge(width = 0.7),
    vjust    = -0.5,
    size     = 2.5,
    fontface = "bold"
  ) +
  scale_fill_manual(
    values = c(Normal  = "#2ecc71",
               Adverse = "#f39c12",
               Extreme = "#e74c3c"),
    name   = "Weather Category"
  ) +
  scale_y_continuous(
    labels = \(x) paste0(x, " hrs"),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title    = "Median Resolution Time by Agency — Lower Volume (2022-2025)",
    subtitle = "Across Normal, Adverse and Extreme weather days",
    x        = NULL,
    y        = "Median Resolution Time (hrs)",
    caption  = "Sources: NYC Open Data — 311 Service Requests; Open-Meteo Historical Archive API (2022-2025)"
  ) +
  theme_bw() +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "gray40"),
    plot.caption    = element_text(color = "gray40"),
    legend.position = "bottom"
  )

resolution_agency_high_chart
resolution_agency_low_chart


```

#### Borough Breakdown

<!-- 
WRITING NOTES:
- How does the weather effect vary across boroughs?
- Key findings from data checks:
  * Queens baseline: 4.22 hrs (fastest) vs Staten Island: 18.4 hrs (slowest)
  * Queens Extreme: 8.58 hrs — faster than most boroughs on Normal days
  * Staten Island Adverse (22.3 hrs) > Extreme (14.5 hrs) — counterintuitive reversal
- These borough differences are as interesting as the weather effect itself
-->

```{r resolution-by-borough-weather}
# Grouped bar chart or heatmap: median resolution time by borough x weather category
# Use plotly for interactivity
```

#### Resolution Time Categories × Weather

<!-- 
WRITING NOTES:
- Cross-tabulate resolution time categories against weather category
- Key question: does adverse weather shift the distribution of resolution types?
- In particular: does extreme weather suppress instant closures while 
  increasing long-term resolutions?
- This is the instant closure analysis flagged during data checks
-->

```{r resolution-categories-weather}
# Stacked bar chart: proportion of each resolution category by weather category
# Shows whether the distribution of resolution types shifts with weather
```

#### Weather-Sensitive Complaint Types

<!-- 
WRITING NOTES:
- Focus on complaint types directly linked to weather:
  HEAT/HOT WATER, Street Condition, Water System, Missed Collection
- Do these types show stronger weather effects than the overall dataset?
- Compare resolution times for these types on Normal vs Adverse vs Extreme days
-->

```{r weather-sensitive-complaints}
# Faceted bar chart: median resolution time by weather category
# for weather-sensitive complaint types only
```

### Statistical Analysis

<!-- 
WRITING NOTES:
- Log-transform resolution times to address right skew
- Run ANOVA testing whether weather category significantly predicts 
  log-transformed resolution time
- Follow up with Tukey's HSD post-hoc test to identify which 
  category pairs differ significantly
- Report effect sizes alongside p-values
- Note: results are back-transformed for interpretation in the text
- Revisit whether ANOVA is the right approach here
-->

```{r statistical-analysis}
# Step 1: Log-transform resolution times
# Step 2: One-way ANOVA: log(resolution_hrs) ~ weather_category
# Step 3: Tukey's HSD post-hoc test
# Step 4: Display results in a gt() table
```

## Interactive Map

<!-- 
WRITING NOTES:
- Leaflet map showing median resolution times across NYC
- Color polygons by median resolution time (green -> amber -> red scale)
- Layer toggle: Borough / Community District
- Popup on click: borough name, request count, median resolution time,
  % of days that were adverse/extreme, top complaint type
- Explain what the map shows and how to interact with it
-->

```{r interactive-map}
# Step 1: Download borough and community district boundary shapefiles
# Step 2: Aggregate median resolution time and weather stats to each geography
# Step 3: Build leaflet map with layer toggle
```

## Conclusions & Limitations

<!-- 
WRITING NOTES:
- Summarize the key findings:
  * Does adverse weather meaningfully affect resolution times?
  * Which boroughs are most affected?
  * Which complaint types are most weather-sensitive?
  * What does the instant closure analysis reveal?
  
- Discuss limitations:
  * One representative weather coordinate per borough — misses within-borough variation
  * Future nearest-neighbor upgrade path using actual request coordinates
  * temp_min_f used for cold classification — captures overnight lows 
    which drive infrastructure stress (pipes, boilers) but may not reflect 
    daytime conditions
  * 14,170 records (0.1%) excluded from weather analysis due to missing borough
  * Median used instead of mean due to right-skewed distribution
  * ANOVA assumes approximate normality — log transformation applied but 
    distribution remains imperfect
    
- Suggest directions for future analysis:
  * Nearest-neighbor spatial join for more precise weather matching
  * Hourly weather data for more granular analysis
  * Adding more weather coordinates per borough
-->



::: {.callout-note title="AI Usage Statement" collapse="true"}

Claude was used to help write the code in this project and brainstorm final deliverable ideas, but all non-code text was generated without the use of any Generative AI tools.\

:::

---

This work ©2025 by Evelyn Rodriguez was initially prepared as a Course Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[Final Project](https://michael-weylandt.com/STA9750/project.html)





