---
title: "Mini-Project 01: Assessing the Impact of SFFA on Campus Diversity One-Year Later"
author: "Evelyn Rodriguez"
editor:
    mode: source
format:
    html:
        toc: true
        code-fold: true
        code-summary: "Display Code"
        code-overflow: wrap
        code-tools: true
        code-block-bg: true
        code-block-border-left: "#6610f2"
        highlight-style: "a11y"
---
```{r setup}
library(gt)
library(skimr)
library(scales)
```

## Introduction

A suitable introduction goes here. In this mini-project, I intend to demonstrate
my skills at Data Summarization/Exploratory Analysis/Table Formatting in an analysis of Official Statistics/Demography data.

## Data Acquisition and Preparation

**Data Acquisition from IPEDS**

```{r data-acquisition}
#| warning: false 
#' IPEDS Data for MP#01 - Data starts in 2010
#' This function will acquire and standardize all data for MP#01
#' from IPEDS (https://nces.ed.gov/ipeds/datacenter/DataFiles.aspx)

acquire_ipeds_data <- function(start_year=2010, end_year=2024){
    library(tidyverse)
    library(glue)
    
    data_dir <- file.path("data", "mp01")
    
    if(!dir.exists(data_dir)){
        dir.create(data_dir, showWarnings=FALSE, recursive=TRUE)
    }
    
    YEARS <- seq(start_year, end_year)
    
    EFA_ALL <- map(YEARS, function(yy){
        if(yy <= 2022){
            ef_url <- glue("https://nces.ed.gov/ipeds/datacenter/data/EF{yy}A.zip")
            
        } else {
            ef_url <- glue("https://nces.ed.gov/ipeds/data-generator?year={yy}&tableName=EF{yy}A&HasRV=0&type=csv")
        }
        
        ef_file <- file.path(data_dir, glue("ef{yy}a.csv.zip"))
        
        if(!file.exists(ef_file)){
            message(glue("Downloading Enrollment Data for {yy} from {ef_url}"))
            download.file(ef_url, destfile = ef_file, quiet=TRUE, mode="wb")    
        }
        
        read_csv(ef_file, 
                 show_col_types=FALSE) |>
            mutate(year = yy, 
                   # American Indian or Alaskan Native
                   enrollment_m_aian = EFAIANM, 
                   enrollment_f_aian = EFAIANW, 
                   # Asian
                   enrollment_m_asia = EFASIAM, 
                   enrollment_f_asia = EFASIAW, 
                   # Black or African-American, 
                   enrollment_m_bkaa = EFBKAAM, 
                   enrollment_f_bkaa = EFBKAAW, 
                   # Hispanic 
                   enrollment_m_hisp = EFHISPM, 
                   enrollment_f_hisp = EFHISPW, 
                   # Native Hawaiian or Other Pacific Islander 
                   enrollment_m_nhpi = EFNHPIM, 
                   enrollment_f_nhpi = EFNHPIW, 
                   # White
                   enrollment_m_whit = EFWHITM, 
                   enrollment_f_whit = EFWHITW, 
                   # Two or More Races
                   enrollment_m_2mor = EF2MORM, 
                   enrollment_f_2mor = EF2MORW, 
                   # Unknown / Undisclosed Race
                   enrollment_m_unkn = EFUNKNM, 
                   enrollment_f_unkn = EFUNKNW, 
                   # US Non-Resident
                   enrollment_m_nral = EFNRALM, 
                   enrollment_f_nral = EFNRALW, 
            ) |> filter(
                (EFALEVEL %in% c(2, 12)) | (LINE %in% c(1, 15))
                # Per 2024 Data Dictionary, 
                # - EFALEVEL 2 = undergrad 
                # - EFALELVE 12 = grad
                # - Line 1 = first year first time full-time undergrad
                # - Line 15 = first year first time part-time undergrad
            ) |> mutate(level = case_when(
                       EFALEVEL == 2 ~ "all undergrad", 
                       EFALEVEL == 12 ~ "all graduate",
                       LINE %in% c(1, 15) ~ "first year undergrad"
                   )
            ) |>
            select(institution_id = UNITID, 
                   year, 
                   level,
                   starts_with("enrollment_")) |>
            group_by(institution_id, 
                     year, 
                     level) |>
            summarize(across(starts_with("enrollment_"), sum), 
                      .groups = "drop")
        
    }) |> bind_rows()
    
    DESC_ALL <- map(YEARS, function(yy){
        if(yy <= 2022){
            hd_url <- glue("https://nces.ed.gov/ipeds/datacenter/data/HD{yy}.zip")
            
        } else {
            hd_url <- glue("https://nces.ed.gov/ipeds/data-generator?year={yy}&tableName=HD{yy}&HasRV=0&type=csv")
        }
        
        hd_file <- file.path(data_dir, glue("hd{yy}.csv.zip"))
        
        if(!file.exists(hd_file)){
            message(glue("Downloading Institutional Descriptions for {yy} from {hd_url}"))
            download.file(hd_url, destfile = hd_file, quiet=TRUE, mode="wb")    
        }
        
        suppressWarnings(
            read_csv(hd_file, 
                 show_col_types=FALSE, 
                 locale=locale(encoding=if_else(yy==2024, "utf-8", "windows-1252"))) |>
            mutate(year = yy, 
                   INSTNM) |> 
            select(institution_id = UNITID, 
                   institution_name = INSTNM, 
                   state = STABBR, 
                   year)
        )
        
    }) |> bind_rows()
    
    inner_join(EFA_ALL, 
               DESC_ALL, 
               join_by(institution_id == institution_id, 
                       year == year))
    
}

IPEDS <- acquire_ipeds_data()

```

### Identify CUNY Schools

**New column identifying CUNY schools by institution name**
```{r identify-cuny-schools}
# Adding is_cuny column

IPEDS <- IPEDS |> 
  mutate(is_cuny = str_detect(institution_name, "CUNY"))

```

### Identify UC and Cal State Schools

**New column identifying California public schools**

In order to correctly identify California public schools, it was necessary to include the '-' at the end of University of California string to prevent non-public universities with 'University of California' in their name from showing up. For example, "Bethesda University of California" is included in the data set but isn't apart of either the UC or CSU systems. 

```{r identify-calpublic-schools}

# Identifying Cali. public schools and creating is_calpublic column
IPEDS <- IPEDS |>
  mutate(is_calpublic = str_detect(institution_name, "California State University") | str_detect(institution_name, "University of California-"))

```

**California Public Schools**
```{r calpublic-preview}
# list of cali schools

IPEDS_california_preview <- IPEDS |>
  select(institution_name, is_calpublic) |>
  filter(is_calpublic == TRUE) |>
  group_by(institution_name) |> 
  summarize(n_programs = n()) |> pull(institution_name)

IPEDS_california_preview
```


## Exploratory Analysis

### Initial EDA Pass
Below are an initial glance at the IPEDS data utilizing the `glimpse` and `skim` functions.

```{r initial-EDA-checks}
glimpse(IPEDS)

skim(IPEDS)

```

### Exploratory Questions

#### Inline Values
**1. How many distinct institutions appear in this data set?**

```{r distinct-institutions}
# Q1: distinct institutions

distinct_institutions <- IPEDS |> 
  summarize(n_institutions = n_distinct(institution_name))

```

This data set contains `r distinct_institutions$n_institutions` distinct institutions.

**2. How many graduate students were enrolled at Baruch in 2024?**

```{r baruch-graduate-students-2024}
# Q2: graduate students at Baruch (2024)
baruch_graduates_2024 <- IPEDS |>
  filter(
    str_detect(institution_name, "Baruch"), 
    level == "all graduate", 
    year == 2024
  ) |>
  mutate(total_graduates = rowSums(across(starts_with("enrollment_")), na.rm = TRUE)) |>
  pull(total_graduates)

```

**3. How many total students were enrolled at Baruch in 2024?**

```{r baruch-total-students-2024}
# Q3: Total students enrolled in Baruch (2024)
baruch_total_2024 <- IPEDS |>
  filter(
    str_detect(institution_name, "Baruch"),
    level %in% c("all undergrad", "all graduate"),
    year == 2024
  ) |>
  mutate(total_students = rowSums(across(starts_with("enrollment_")), na.rm = TRUE)) |>
  summarize(total_enrollment = sum(total_students)) |>
  pull(total_enrollment)

```

**4. Which institution had the highest number of enrolled female students in 2019?**

```{r highest-female-students-2019}
# Q4: institution w/ highest # enrolled f students (2019)

female_enrollment_2019 <- IPEDS |>
  filter(
    year == 2019, 
    level %in% c("all undergrad", "all graduate")
  ) |>
  mutate(total_female_students = rowSums(across(starts_with("enrollment_f_")), na.rm = TRUE)) |>
  group_by(institution_id, institution_name) |>
  summarize(total_female_enrollment = sum(total_female_students), .groups = "drop") |>
  slice_max(total_female_enrollment, n = 1) |> 
  select(institution_name, total_female_enrollment)


```

`r female_enrollment_2019$institution_name` had the highest female enrollment with `r comma(female_enrollment_2019$total_female_enrollment)` female students.

**5. Which institution with over 1000 total students admitted the highest proportion of Native Hawaiian or Pacific Islander (nhpi) first-year undergraduates in 2024? **

```{r highest-nhpi}
nhpi_2024 <- IPEDS |>
  filter(
    year == 2024, 
    level == "first year undergrad"
  ) |>
  mutate(
    total_enrollment = rowSums(across(starts_with("enrollment_")), na.rm = TRUE),
    total_nhpi = rowSums(across(starts_with("enrollment_") & ends_with("nhpi")), na.rm = TRUE), 
    percent_nhpi = total_nhpi / total_enrollment
         ) |>
  filter(total_enrollment >= 1000) |>
  slice_max(percent_nhpi, n = 1) |>
  select(institution_name, total_enrollment, total_nhpi, percent_nhpi)

```

`r nhpi_2024$institution_name` has the highest proportion of Native Hawaiian or Pacific Islander first year undergraduates (`r percent(nhpi_2024$percent_nhpi, accuracy=0.001)`) out of all universities with over 1000 total students admitted. 

---

#### Tables

**1. Which 5 states had the highest number of graduate students across all institutions located in that state?**

```{r highest-graduate-states}
high_graduate_states <- IPEDS |>
  filter(level == "all graduate") |>
  mutate(total_graduate_students = rowSums(across(starts_with("enrollment_")), na.rm = TRUE)) |>
  group_by(state) |>
  summarise(total_state_graduate_students = sum(total_graduate_students), .groups = "drop") |>
  slice_max(total_state_graduate_students, n = 5) |>
  select(state, total_state_graduate_students) |>
  gt() |>
  cols_label(
    state = md("**State**"), 
    total_state_graduate_students = md("**Total Graduate Students**")
  ) |>
  fmt_number(
    columns = total_state_graduate_students, 
    decimals = 0
  ) |>
  tab_header(
    title = md("**Top 5 States by Graduate Enrollment**"), 
    subtitle = "Enrollment combined for all years across institutions reporting to IPEDS"
  )

high_graduate_states
```

**2. In 2024, how many first year undergraduate students were enrolled at CUNY colleges and which colleges did they attend? Report both absolute enrollment numbers and percent of total first-year undergraduates?**

```{r cuny-first-year-enrollment}

cuny_first_year_enrollment_2024 <- IPEDS |> 
  filter(
    year == 2024,
    is_cuny == TRUE, 
    level == "first year undergrad"
  ) |>
  mutate(first_year_total_enrollment = rowSums(across(starts_with("enrollment_")), na.rm = TRUE)) |>
  select(institution_name, first_year_total_enrollment) |>
  mutate(percent_total = first_year_total_enrollment/ sum(first_year_total_enrollment)) |>
  gt() |>
  cols_label(
    institution_name = md("**CUNY College**"), 
    first_year_total_enrollment = md("**First-Year<br>Enrollment at College**"),
    percent_total = md("**% of Total<br>First-Year Enrollment**")
  ) |>
  fmt_percent(
    columns = percent_total, 
    decimals = 2
  ) |>
  grand_summary_rows(
    columns = first_year_total_enrollment, 
    fns = list(Total = ~ sum(.)),
    fmt = ~ fmt_number(., decimals = 0)
  ) |>
  cols_align(
    align = "center",
    columns = c(first_year_total_enrollment, percent_total)
  ) |>
  cols_width(institution_name ~ px(150)) |>
  tab_header(
    title = md("**CUNY First-Year Undergraduate Enrollment (2024)**"),
    subtitle = "Enrollment by College"
  )

cuny_first_year_enrollment_2024

```


**3. How has Baruch’s total undergraduate enrollment changed over the study period? Report both enrollment numbers and percent change year-over-year.**

```{r undergrad-change-YoY}

baruch_undergrad_change_yoy <- IPEDS |>
  filter(
    is_cuny == TRUE, 
    str_detect(institution_name, "Baruch"), 
    level == "all undergrad"
  ) |> 
  mutate(total_undergrad_enrollment = rowSums(across(starts_with("enrollment_")), na.rm = TRUE)) |>
  select(year, total_undergrad_enrollment) |>
  arrange(year) |>
  mutate(
    change_yoy = total_undergrad_enrollment - lag(total_undergrad_enrollment), 
    percent_change_yoy = change_yoy / lag(total_undergrad_enrollment)
  ) |>
  gt() |>
  cols_label(
    year = md("**Year**"), 
    total_undergrad_enrollment = md("**Undergraduate<br>Enrollment**"),
    change_yoy = md("**Change<br>Year-over-Year**"), 
    percent_change_yoy = md("**% Change<br>Year-over-Year**")
  ) |>
  fmt_percent(
    columns = percent_change_yoy, 
    decimals = 2
  ) |>
  tab_style(
    style = cell_text(color = "darkgreen"),
    locations = cells_body(columns = c(change_yoy, percent_change_yoy), 
                           rows = change_yoy > 0)
  ) |>
  tab_style(
    style = cell_text(color = "darkred"),
    locations = cells_body(columns = c(change_yoy, percent_change_yoy),
                           rows = change_yoy < 0)
  ) |>
  sub_missing(missing_text = "—") |>
  cols_align(align = "center") |>
  tab_header(
    title = "CUNY Baruch College Undergraduate Enrollment (2010-2024)",
    subtitle = "Total enrollment change year-over-year"
  )
  
baruch_undergrad_change_yoy
```


**4. At what 5 institutions did the fraction of white students decrease the most over the period from 2010 to 2020?**

```{r white-decline}

white_decline <- IPEDS |>
  filter(
    level %in% c("all undergrad", "all graduate"), 
    year %in% c(2010, 2020)
  ) |>
  mutate(
    total_enrollment = rowSums(across(starts_with("enrollment_")), na.rm = TRUE),
    total_white_enrollment = rowSums(across(starts_with("enrollment_") & ends_with("whit")), na.rm = TRUE)
  ) |>
  group_by(institution_id, institution_name, state, year) |>
  summarize(
    school_total_enrollment = sum(total_enrollment),
    school_total_white_enrollment = sum(total_white_enrollment),
    .groups = "drop"
  ) |>
  mutate(percent_white = school_total_white_enrollment / school_total_enrollment) |>
  filter(school_total_enrollment >= 1000) |>
  select(institution_id, institution_name, state, year, percent_white) |>
  pivot_wider(
    names_from = year,
    values_from = percent_white,
    names_prefix = "percent_white_"
  ) |> 
  filter(!is.na(percent_white_2010), !is.na(percent_white_2020)) |>
  mutate(percent_change = percent_white_2020 - percent_white_2010) |>
  slice_min(percent_change, n=5) |>
  select(institution_name, state, percent_white_2010, percent_white_2020, percent_change) |>
  gt() |>
  cols_label(
    institution_name = md("**School**"),
    state = md("**State**"),
    percent_white_2010 = md("**% White<br>(2010)**"),
    percent_white_2020 = md("**% White<br>(2020)**"),
    percent_change = md("**Change**")
  ) |>
  cols_align(
    align = "center",
    columns = c(state, percent_white_2010, percent_white_2020, percent_change)
  ) |>
  cols_width(institution_name ~ px(100)) |>
  fmt_percent(
    columns = c(percent_white_2010, percent_white_2020, percent_change),
    decimals = 2
  ) |>
  tab_style(
    style = cell_text(color = "darkred"),
    locations = cells_body(columns = percent_change)
  ) |>
  tab_header(
    title = md("**Largest Decreases in White Student Enrollment (2010-2020)**"),
    subtitle = "For schools with 1,000+ total enrollment"
  )

white_decline

```


**5. In which 3 states did the fraction of female undergraduates increase the most over the period from 2010 to 2024?**

```{r female-undergrad-increase}

female_undergrad_increase <- IPEDS |>
  filter(
    level == "all undergrad", 
    year %in% c(2010, 2024)
  ) |>
  mutate(
    total_enrollment = rowSums(across(starts_with("enrollment_")), na.rm = TRUE),
    total_female_enrollment = rowSums(across(starts_with("enrollment_f_")), na.rm = TRUE)
  ) |>
  group_by(state, year) |>
  summarize(
    state_total_enrollment = sum(total_enrollment),
    state_total_female_enrollment = sum(total_female_enrollment),
    .groups = "drop"
  ) |>
  mutate(percent_female = state_total_female_enrollment / state_total_enrollment) |>
  select(state, year, state_total_enrollment, percent_female) |>
  pivot_wider(
    names_from = year, 
    values_from = c(percent_female, state_total_enrollment),
    names_sep = "_"
  ) |>
  filter(
    !is.na(percent_female_2010),
    !is.na(percent_female_2024)
  ) |>
  mutate(percent_change = percent_female_2024 - percent_female_2010) |>
  slice_max(percent_change, n=3) |>
  select(state, percent_female_2010, percent_female_2024, percent_change, state_total_enrollment_2010, state_total_enrollment_2024) |>
  gt() |>
  cols_label(
    state = md("**State**"),
    percent_female_2010 = md("**% Female<br>(2010)**"),
    percent_female_2024 = md("**% Female<br>(2024)**"),
    percent_change = md("**Change**"),
    state_total_enrollment_2010 = md("**Total Enrollment<br>(2010)**"),
    state_total_enrollment_2024 = md("**Total Enrollment<br>(2024)**")
  ) |>
  cols_align(
    align = "center",
  ) |>
  fmt_percent(
    columns = c(percent_female_2010, percent_female_2024, percent_change),
    decimals = 2
  ) |>
  tab_style(
    style = cell_text(color = "darkgreen"),
    locations = cells_body(columns = percent_change)
  ) |>
  tab_header(
    title = md("**Largest Increases in Female Undergraduate Enrollment (2010-2024)**"),
    subtitle = "All states and territories with data in both 2010 and 2024"
  )

female_undergrad_increase

```


## Final Insights and Deliverable

Code related to the final deliverable of the assignment goes here. 


::: {.callout-note title="AI Usage Statement" collapse="true"}

Claude was used to look up how to use functions that are new to me (such as sumRows), pull up documentation on gt, r, and tidyverse packages, and to check over code blocks for efficiency. 

:::

------------------------------------------------------------------------

This work ©2026 by Evelyn Rodriguez was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #01](https://michael-weylandt.com/STA9750/mini/mini01.html)
