---
title: "Mini-Project 02: How Do You Do 'You Do You'?"
author: "Evelyn Rodriguez"
editor:
    mode: source
format:
    html:
        toc: true
        code-fold: true
        code-summary: "Display Code"
        code-overflow: wrap
        code-block-bg: true
        code-block-border-left: "#6610f2"
        highlight-style: "a11y"
---

## Introduction

A suitable introduction goes here. In this mini-project, I intend to demonstrate
my skills at Combining Data/Data Visualization in an analysis of Survey Date/Social Sciences data.

## Data Acquisition and Preparation

### Data Acquisition from ATUS

```{r data-acquisitions}
#| warning: false
# Your code goes in chunks like this

library(tidyverse)
library(glue)
library(readxl)
library(httr2)
library(skimr)
library(dplyr)

load_atus_data <- function(file_base = c("resp", "rost", "act")){
    if(!dir.exists(file.path("data", "mp02"))){
        dir.create(file.path("data", "mp02"), showWarnings=FALSE, recursive=TRUE)
    }
    
    file_base <- match.arg(file_base)
    
    file_name_out <- file.path("data", 
                               "mp02", 
                               glue("atus{file_base}_0324.dat"))
        
    if(!file.exists(file_name_out)){
        
        url_end <- glue("atus{file_base}-0324.zip")
        
        temp_zip <- tempfile(fileext=".zip")
        temp_dir <- tools::file_path_sans_ext(temp_zip)
        
        request("https://www.bls.gov") |>
            req_url_path("tus", "datafiles", url_end) |>
            req_headers(`User-Agent` = "Mozilla/5.0 (Macintosh; Intel Mac OS X 10.15; rv:149.0) Gecko/20100101 Firefox/149.0") |>
            req_perform(temp_zip)
    
        unzip(temp_zip, exdir = temp_dir)
        
        temp_file <- file.path(temp_dir, basename(file_name_out))
        
        file.copy(temp_file, file_name_out)
    }
    
    file <- read_csv(file_name_out, 
                     show_col_types = FALSE)
    
    switch(file_base, 
           resp=file |> 
                  rename(survey_weight = TUFNWGTP, 
                         time_alone = TRTALONE, 
                         survey_year = TUYEAR) |>
                  #adding a column that counts number of children from roster file
                  left_join(
                    load_atus_data("rost") |>
                      filter(age < 18) |>
                      count(participant_id, name = "n_children_under18"),
                    join_by(TUCASEID == participant_id)
                  ) |>
                  #coalesce will turn NA values into 0
                  mutate(n_children_under18 = coalesce(n_children_under18, 0L),
                         labor_force_status = case_match(TELFS,
                            1 ~ "Employed - at work",
                            2 ~ "Employed - absent",
                            3 ~ "Unemployed - on layoff",
                            4 ~ "Unemployed - looking",
                            5 ~ "Not in labor force"),
                         school_enrollment = case_match(TESCHENR,
                           1 ~ "Enrolled",
                           2 ~ "Not Enrolled",
                           -1 ~ "Blank"),
                         #checks if enrolled and teschlvl matches college
                         college_enrolled = TESCHENR == 1 & TESCHLVL == 2)
           ,
           rost=file |>
                  mutate(sex = case_match(TESEX, 1 ~ "M", 2 ~ "F"),
                         #adding relation to respondent
                         relation = case_match(TERRP,
                           c(18,19) ~ "Self",
                           20 ~ "Spouse",
                           21 ~ "Unmarried partner",
                           22 ~ "Own household child",
                           23 ~ "Grandchild",
                           24 ~ "Parent",
                           25 ~ "Brother/sister",
                           26 ~ "Other related person",
                           27 ~ "Foster child",
                           28 ~ "Housemate/roommate",
                           29 ~ "Roomer/boarder",
                           30 ~ "Other nonrelative",
                           40 ~ "Own nonhousehold child < 18")) |>
                  rename(age = TEAGE), 
           act =file |>
                mutate(activity_n_of_day = TUACTIVITY_N, 
                       time_spent_min = TUACTDUR24, 
                       start_time = TUSTARTTIM,
                       stop_time = TUSTOPTIME, 
                       level1_code = paste0(TRTIER1P, "0000"), 
                       level2_code = paste0(TRTIER2P, "00"),
                       level3_code = TRCODEP,
                       #location of activity
                       location = case_match(TEWHERE,
                         1 ~ "Home",
                         2 ~ "Workplace",
                         3 ~ "Someone else's home",
                         4 ~ "Restaurant bar",
                         5 ~ "Place of worship",
                         6 ~ "Grocery store",
                         7 ~ "Other store/mall",
                         8 ~ "School",
                         9 ~ "Outdoors",
                         10 ~ "Library",
                         11 ~ "Other place",
                         12 ~ "Car, truck, or motorcycle (driver)",
                         13 ~ "Car, truck, or motorcycle (passenger)",
                         14 ~ "Walking",
                         15 ~ "Bus",
                         16 ~ "Subway/train",
                         17 ~ "Bicycle",
                         18 ~ "Boat/ferry",
                         19 ~ "Taxi/limousine service",
                         20 ~ "Airplane",
                         21 ~ "Other transportation",
                         30 ~ "Bank",
                         31 ~ "Gym/health club",
                         32 ~ "Post office",
                         89 ~ "Unspecificed place",
                         99 ~ "Unspecified transportation",
                         (-1) ~ "Blank"
                       )) 
           ) |> 
                rename(participant_id = TUCASEID) |>
                select(matches("[:lower:]", 
                               ignore.case=FALSE))
}

load_atus_activities <- function(){
    if(!dir.exists(file.path("data", "mp02"))){
        dir.create(file.path("data", "mp02"), showWarnings=FALSE, recursive=TRUE)
    }
    
    dest_file <- file.path("data",
                           "mp02",
                           "atus_activity_codes.csv")
    
    if(!file.exists(dest_file)){
        download.file("https://michael-weylandt.com/STA9750/mini/atus_activity_codes.csv", 
                      quiet=TRUE, 
                      destfile=dest_file, 
                      mode="wb")
    }
    
    read_csv(dest_file, show_col_types = FALSE) 
}

respondents  <- load_atus_data("resp")
roster       <- load_atus_data("rost")
activities   <- load_atus_data("act")
activity_codes <- load_atus_activities()

glimpse(respondents)

glimpse(activities)

glimpse(activity_codes)

glimpse(roster)


```

and you can then print variables from chunks: 

```{r}

```

or inline, like this: $x$ is .

## Exploratory Analysis

Code related to Exploratory Data Analysis goes here. This may include
exploratory graphics, instructor-provided exploratory questions, or other
similar elements. 

## Final Insights and Deliverable

Code related to the final deliverable of the assignment goes here. 



------------------------------------------------------------------------

This work ©2025 by er0driguez was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #02](https://michael-weylandt.com/STA9750/mini/mini02.html)
