---
title: "Mini-Project 02: How Do You Do 'You Do You'?"
author: "Evelyn Rodriguez"
editor:
    mode: source
format:
    html:
        toc: true
        code-fold: true
        code-summary: "Display Code"
        code-overflow: wrap
        code-block-bg: true
        code-block-border-left: "#6610f2"
        highlight-style: "a11y"
---

## Introduction

A suitable introduction goes here. In this mini-project, I intend to demonstrate
my skills at Combining Data/Data Visualization in an analysis of Survey Date/Social Sciences data.

## Data Acquisition and Preparation

### Data Acquisition from ATUS

```{r data-acquisitions}
#| warning: false
# Your code goes in chunks like this

library(tidyverse)
library(glue)
library(readxl)
library(httr2)
library(skimr)
library(dplyr)
library(scales)
library(gt)
library(Hmisc)
library(patchwork)

load_atus_data <- function(file_base = c("resp", "rost", "act", "cps")){
    if(!dir.exists(file.path("data", "mp02"))){
        dir.create(file.path("data", "mp02"), showWarnings=FALSE, recursive=TRUE)
    }
    
    file_base <- match.arg(file_base)
    
    file_name_out <- file.path("data", 
                               "mp02", 
                               glue("atus{file_base}_0324.dat"))
        
    if(!file.exists(file_name_out)){
        
        url_end <- glue("atus{file_base}-0324.zip")
        
        temp_zip <- tempfile(fileext=".zip")
        temp_dir <- tools::file_path_sans_ext(temp_zip)
        
        request("https://www.bls.gov") |>
            req_url_path("tus", "datafiles", url_end) |>
            req_headers(`User-Agent` = "Mozilla/5.0 (Macintosh; Intel Mac OS X 10.15; rv:149.0) Gecko/20100101 Firefox/149.0") |>
            req_perform(temp_zip)
    
        unzip(temp_zip, exdir = temp_dir)
        
        temp_file <- file.path(temp_dir, basename(file_name_out))
        
        file.copy(temp_file, file_name_out)
    }
    
    file <- read_csv(file_name_out, 
                     show_col_types = FALSE)
    
    switch(file_base, 
           #add cps data to determine if respondent is married
           cps=file |>
             filter(TULINENO == 1) |>
             mutate(marital_status = case_match(PEMARITL,
               1 ~ "Married - spouse present",
               2 ~ "Married - spouse absent",
               3 ~ "Widowed",
               4 ~ "Divorced",
               5 ~ "Separated",
               6 ~ "Never married"),
             #check if respondents are married
               has_spouse = marital_status %in% c("Married - spouse present", "Married - spouse absent"),
               is_retired = PEMLR == 5,
             #check if respondent works in education by main job occupation code
               in_education = PRDTOCC1 == 8)
             ,
           resp=file |> 
                  rename(survey_weight = TUFNWGTP, 
                         time_alone = TRTALONE, 
                         survey_year = TUYEAR,
                         survey_month = TUMONTH,
                         weekly_earnings = TRERNWA) |>
                  #add a column that counts number of children from roster file
                  left_join(
                    load_atus_data("rost") |>
                      filter(relation == "Self") |>
                      select(participant_id, age, sex), 
                    join_by(TUCASEID == participant_id)
                  ) |>
                  left_join(
                    load_atus_data("rost") |>
                      filter(age < 18) |>
                      count(participant_id, name = "n_children_under18"),
                    join_by(TUCASEID == participant_id)
                  ) |>
                  # pull in CPS data to add marital_status and has_spouse columns
                  left_join(
                    load_atus_data("cps"),
                    join_by(TUCASEID == participant_id)
                  ) |>
                  #coalesce will turn NA values into 0
                  mutate(n_children_under18 = coalesce(n_children_under18, 0L),
                         labor_force_status = case_match(TELFS,
                            1 ~ "Employed - at work",
                            2 ~ "Employed - absent",
                            3 ~ "Unemployed - on layoff",
                            4 ~ "Unemployed - looking",
                            5 ~ "Not in labor force"),
                         #creates column indicating enrollment
                         school_enrollment = case_match(TESCHENR,
                           1 ~ "Enrolled",
                           2 ~ "Not Enrolled",
                           -1 ~ "Blank"),
                         #create demographic variable that checks if enrolled and teschlvl matches college/university
                         college_enrolled = TESCHENR == 1 & TESCHLVL == 2)
           ,
           rost=file |>
                  mutate(sex = case_match(TESEX, 1 ~ "M", 2 ~ "F"),
                         #adding relation to respondent
                         relation = case_match(TERRP,
                           c(18,19) ~ "Self",
                           20 ~ "Spouse",
                           21 ~ "Unmarried partner",
                           22 ~ "Own household child",
                           23 ~ "Grandchild",
                           24 ~ "Parent",
                           25 ~ "Brother/sister",
                           26 ~ "Other related person",
                           27 ~ "Foster child",
                           28 ~ "Housemate/roommate",
                           29 ~ "Roomer/boarder",
                           30 ~ "Other nonrelative",
                           40 ~ "Own nonhousehold child < 18")) |>
                  rename(age = TEAGE), 
           act =file |>
                mutate(activity_n_of_day = TUACTIVITY_N, 
                       time_spent_min = TUACTDUR24, 
                       start_time = TUSTARTTIM,
                       stop_time = TUSTOPTIME, 
                       level1_code = paste0(TRTIER1P, "0000"), 
                       level2_code = paste0(TRTIER2P, "00"),
                       level3_code = TRCODEP,
                       #location of activity
                       location = case_match(TEWHERE,
                         1 ~ "Home",
                         2 ~ "Workplace",
                         3 ~ "Someone else's home",
                         4 ~ "Restaurant bar",
                         5 ~ "Place of worship",
                         6 ~ "Grocery store",
                         7 ~ "Other store/mall",
                         8 ~ "School",
                         9 ~ "Outdoors",
                         10 ~ "Library",
                         11 ~ "Other place",
                         12 ~ "Car, truck, or motorcycle (driver)",
                         13 ~ "Car, truck, or motorcycle (passenger)",
                         14 ~ "Walking",
                         15 ~ "Bus",
                         16 ~ "Subway/train",
                         17 ~ "Bicycle",
                         18 ~ "Boat/ferry",
                         19 ~ "Taxi/limousine service",
                         20 ~ "Airplane",
                         21 ~ "Other transportation",
                         30 ~ "Bank",
                         31 ~ "Gym/health club",
                         32 ~ "Post office",
                         89 ~ "Unspecificed place",
                         99 ~ "Unspecified transportation",
                         (-1) ~ "Blank"
                       )) 
           ) |> 
                rename(participant_id = TUCASEID) |>
                select(matches("[:lower:]", 
                               ignore.case=FALSE))
}

load_atus_activities <- function(){
    if(!dir.exists(file.path("data", "mp02"))){
        dir.create(file.path("data", "mp02"), showWarnings=FALSE, recursive=TRUE)
    }
    
    dest_file <- file.path("data",
                           "mp02",
                           "atus_activity_codes.csv")
    
    if(!file.exists(dest_file)){
        download.file("https://michael-weylandt.com/STA9750/mini/atus_activity_codes.csv", 
                      quiet=TRUE, 
                      destfile=dest_file, 
                      mode="wb")
    }
    
    read_csv(dest_file, show_col_types = FALSE) 
}

respondents  <- load_atus_data("resp")
cps <- load_atus_data("cps") 
roster       <- load_atus_data("rost")
activities   <- load_atus_data("act")
activity_codes <- load_atus_activities()

glimpse(respondents)

glimpse(cps)

glimpse(activities)

glimpse(activity_codes)

glimpse(roster)


```

and you can then print variables from chunks: 

```{r}

```

or inline, like this: $x$ is .

## Exploratory Analysis

Code related to Exploratory Data Analysis goes here. This may include
exploratory graphics, instructor-provided exploratory questions, or other
similar elements. 

### Inline Values

**1. How many different respondents have answered an ATUS survey since 2003?**

```{r unique-respondents}

unique_respondents <- respondents |> 
  distinct(participant_id) |>
  nrow()

```

There are `r comma(unique_respondents)` respondents that answered the ATUS survey since 2003.


**2. Americans enjoy watching a variety of sports. How many different sports does ATUS ask about watching as a “Level 3” task?**

```{r number-of-sports-watched}

n_sports_watched <- activity_codes |>
  filter(code_level2 == 130200) |>
  summarise(n = n()) |>
  pull(n)

```

ATUS asks about watching `r n_sports_watched` different sports as a Level 3 task. 

**3. Approximately what percent of Americans are retired?**

```{r percent-retired}

percent_retired <- respondents |>
  summarise(pct_retired = weighted.mean(is_retired, survey_weight)) |>
  pull(pct_retired)

```

Approximately `r percent(percent_retired)` of Americans are retired.


**4. On average, how many hours do Americans sleep per night?**

```{r average-sleep-hours}

avg_sleep_hours <- activities |>
  #filtering only on level 3 code 10101 since we don't know if Other Sleep includes actually sleeping
  filter(level3_code == "010101") |>
  group_by(participant_id) |>
  summarise(total_sleep_min = sum(time_spent_min)) |>
  left_join(
    respondents |> select(participant_id, survey_weight),
    join_by(participant_id)
  ) |>
  summarise(avg_sleep_hours = weighted.mean(total_sleep_min, survey_weight) / 60) |>
  pull(avg_sleep_hours)

```

On average, Americans sleep `r round(avg_sleep_hours, 1)` hours per night.

**5. On average, how many hours do Americans with one or more children in the home spend taking care of their children (all activities including healthcare and involvement in their children’s education)?**

```{r average-childcare-hours}

avg_childcare_hours <- activities |>
  filter(
    level1_code == "030000",
    level2_code %in% c("030100", "030200", "030300")) |>
  group_by(participant_id) |>
  summarise(total_childcare_min = sum(time_spent_min)) |>
  #right join ensures we keep all respondents with children even if they recorded zero childcare activities
  #this will prevent an upward bias in the average
  right_join(
    respondents |>
      filter(n_children_under18 >= 1) |>
      select(participant_id, survey_weight),
    join_by(participant_id)
  ) |>
  mutate(total_childcare_min = coalesce(total_childcare_min, 0L)) |>
  summarise(avg_childcare_hours = weighted.mean(total_childcare_min, survey_weight) / 60) |>
  pull(avg_childcare_hours)

```

On average, American with one or more children spend `r round(avg_childcare_hours, 2)` hours taking care of their children. 


### Tables

**1. Do retired persons spend more time on lawn-care than people who have full-time employment?**

```{r lawn-care-time}

lawn_care <- activities |>
  filter(
    level2_code == "020500",
    #filter only one level 3 activity code that specifically talks about lawns since we don't know what is included in Other category
    level3_code == "020501"
  ) |>
  group_by(participant_id) |>
  summarise(total_lawn_min = sum(time_spent_min)) |>
  right_join(
    respondents |>
      filter(is_retired | labor_force_status == "Employed - at work") |>
      select(participant_id, survey_weight, is_retired, labor_force_status), 
    join_by(participant_id)
  ) |>
  mutate(total_lawn_min = coalesce(total_lawn_min, 0L)) |>
  group_by(is_retired) |>
  summarise(avg_lawn_hours = weighted.mean(total_lawn_min, survey_weight) / 60)

#table
lawn_care_table <- lawn_care |>
  mutate(is_retired = if_else(is_retired, "Retired", "Employed Full-Time")) |>
  gt() |>
  cols_label(
    is_retired = md("**Employment Status**"),
    avg_lawn_hours = md("**Average Hours Spent<br>on Lawn Care**")
  ) |>
  cols_align(align = "center", columns = avg_lawn_hours) |>
  fmt_number(avg_lawn_hours, decimals = 2) |>
  tab_header(
    title = md("**Lawn Care Hours by Employment Status**"), 
    subtitle = "Average Time Spent on Lawn-Care for Retired vs. Full-Time Employed Persons"
  ) |>
  tab_source_note(
    source_note = md("*Source: [American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm), 2003-2024, U.S. Bureau of Labor Statistics*")
  )

lawn_care_table

```

On average, retired people spend more time on lawn-care than full-time employed people.


**2. How does the amount of time spent on social and recreational activities typically compare for married and unmarried people?**

```{r social-and-recreational-activities-married-vs-unmarried}

social_rec <- activities |>
  filter(level1_code %in% c("120000", "130000")) |>
  group_by(participant_id) |>
  summarise(total_social_rec_min = sum(time_spent_min)) |>
  right_join(
    respondents |>
      select(participant_id, survey_weight, has_spouse),
    join_by(participant_id)
  ) |>
  mutate(total_social_rec_min = coalesce(total_social_rec_min, 0L)) |>
  group_by(has_spouse) |>
  summarise(avg_social_rec_hours = weighted.mean(total_social_rec_min, survey_weight) / 60)

social_rec_table <- social_rec |>
  mutate(has_spouse = if_else(has_spouse, "Married", "Unmarried")) |>
  gt() |>
  cols_label(
    has_spouse = md("**Marital Status**"),
    avg_social_rec_hours = md("**Avg. Hours Spent on<br>Social/Recreational Activities**")
  ) |>
  cols_align(align = "center", columns = avg_social_rec_hours) |>
  fmt_number(avg_social_rec_hours, decimals = 2) |>
  tab_header(
    title = md("**Social and Recreational Activity Time by Marital Status**"), 
    subtitle = md("Average Time Spent on Social and Recreational Activities for Married vs. Non-Married People")
  ) |>
  tab_source_note(
    source_note = md("*Source: [American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm), 2003-2024, U.S. Bureau of Labor Statistics*")
  )

social_rec_table
  

```


**3. Where do high-earning Americans spend the majority of their time?**

```{r high-earner-time}

high_earner_threshold <- respondents |>
  filter(weekly_earnings > 0) |>
  #define high earner threshold as someone earning above the 75th percentile
  summarise(threshold = quantile(weekly_earnings, 0.75)) |>
  pull(threshold)

high_earner_location <- activities |>
  right_join(
    respondents |>
      filter(weekly_earnings >= high_earner_threshold) |>
      select(participant_id, survey_weight),
    join_by(participant_id)
  ) |>
  #remove rows where location is missing or blank
  filter(!is.na(location), location != "Blank") |>
  group_by(location) |>
  summarise(avg_hours = weighted.mean(time_spent_min, survey_weight) / 60) |>
  arrange(desc(avg_hours))

high_earner_location_table <- high_earner_location |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(
    location = md("**Location**"),
    avg_hours = md("**Average Hours per Day**")
  ) |>
  cols_align(align = "center", columns = avg_hours) |>
  fmt_number(avg_hours, decimals = 2) |>
  tab_header(
    title = md("**Where Do High-Earning Americans Spend Their Time**"), 
    subtitle = "Locations where respondents above the 75th percentile of weekly earnings spend the majority of their time"
  ) |>
  tab_source_note(
    source_note = md("*Source: [American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm), 2003-2024, U.S. Bureau of Labor Statistics*")
  )

high_earner_location_table  
  

```


*High earners are defined as those earning above the 75th percentile for weekly earnings*


**4. What are the most common activities reported by Americans when on a bus, train, subway, boat, or ferry?**

```{r common-transit-activities}

transit_activities <- activities |>
  filter(location %in% c("Bus", "Subway/train", "Boat/ferry")) |>
  left_join(
    activity_codes |> select(code_level3, task_level3) |> distinct(), 
    join_by(level3_code == code_level3)
  ) |>
  filter(!is.na(task_level3)) |>
  group_by(task_level3) |>
  summarise(n_occurrences = n()) |>
  arrange(desc(n_occurrences)) |>
  slice_head(n = 10)

transit_activities_table <- transit_activities |>
  gt() |>
  cols_label(
    task_level3 = md("**Activity**"),
    n_occurrences = md("**Number of Occurrences**")
  ) |>
  cols_align(align = "center", columns = n_occurrences) |>
  tab_header(
    title = md("**Most Common Activities on Public Transportation**"),
    subtitle = "Activities on Bus, Subway, Train, Boat or Ferry"
  ) |>
  tab_source_note(
    source_note = md("*Source: [American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm), 2003-2024, U.S. Bureau of Labor Statistics*")
  )

transit_activities_table

```

*Most common, in this case, means most frequently reported activity as opposed to activity with most reported time spent on activity*


**5. How much time per day do Americans spend on your four favorite activities and how does this vary by age group?**

**Evelyn's Favorite Activities:**

* Reading (ATUS Level 3 Task Name: Reading for personal interest)

* Playing Video and Board Games (ATUS Level 3 Task Name: Playing games)

* Lifting at the Gym (ATUS Level 3 Task Name: Weightlifting or strength training)

* Scuba Diving (ATUS Level 3 Task Name: Participating in water sports)

***Note:** Analysis includes those who reported 0 time spent on these activities so we can see prevalance in the general population*

```{r favorite-activities-by-age-group-one-table}

c("Reading for personal interest", "Playing games", "Weightlifting or strength training", "Participating in water sports") |>
    map(\(activity_name) {
        code <- activity_codes |>
            filter(task_level3 == activity_name) |>
            pull(code_level3)
        
        activities |>
            filter(level3_code == code) |>
            right_join(
                respondents |> select(participant_id, survey_weight),
                join_by(participant_id)
            ) |>
            mutate(time_spent_min = coalesce(time_spent_min, 0L)) |>
            left_join(
                roster |>
                    filter(relation == "Self") |>
                    select(participant_id, age),
                join_by(participant_id)
            ) |>
            mutate(age_group = cut(age, breaks = c(0, 24, 34, 44, 54, 64, Inf),
                                   labels = c("Under 25", "25-34", "35-44",
                                              "45-54", "55-64", "65 and over")),
                   activity = activity_name) |>
            group_by(activity, age_group) |>
            summarise(avg_min = weighted.mean(time_spent_min, survey_weight),
                      .groups = "drop")
    }) |>
    list_rbind() |>
    pivot_wider(names_from = activity, values_from = avg_min) |>
    gt() |>
    cols_label(
        age_group = md("**Age Group**"), 
        `Reading for personal interest` = md("**Reading**"),
        `Playing games` = md("**Playing Games**"),
        `Weightlifting or strength training` = md("**Weightlifting or<br>Strength Training**"), 
        `Participating in water sports` = md("**Participating in<br>Water Sports**")
    ) |>
    fmt_number(c("Reading for personal interest", "Playing games", "Weightlifting or strength training", "Participating in water sports"), 
               decimals = 0) |>
    cols_align(align = "center", 
               columns = c("Reading for personal interest", "Playing games", "Weightlifting or strength training", "Participating in water sports")) |>
    tab_header(
        title = md("**Time Spent on Evelyn's Favorite Activities by Age Group**"), 
        subtitle = "Average Minutes per Day for Each Activity"
    ) |>
    tab_source_note(
        source_note = md("*Source: [American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm), 2003-2024, U.S. Bureau of Labor Statistics*")
    )


c("Reading for personal interest", "Playing games", "Weightlifting or strength training", "Participating in water sports") |>
    map(\(activity_name) {
        code <- activity_codes |>
            filter(task_level3 == activity_name) |>
            pull(code_level3)
        
        activities |>
            filter(level3_code == code) |>
            inner_join(
                respondents |> select(participant_id, survey_weight),
                join_by(participant_id)
            ) |>
            left_join(
                roster |>
                    filter(relation == "Self") |>
                    select(participant_id, age),
                join_by(participant_id)
            ) |>
            mutate(age_group = cut(age, breaks = c(0, 24, 34, 44, 54, 64, Inf),
                                   labels = c("Under 25", "25-34", "35-44",
                                              "45-54", "55-64", "65 and over")),
                   activity = activity_name) |>
            group_by(activity, age_group) |>
            summarise(avg_min = weighted.mean(time_spent_min, survey_weight),
                      .groups = "drop")
    }) |>
    list_rbind() |>
    pivot_wider(names_from = activity, values_from = avg_min) |>
    gt() |>
    cols_label(
        age_group = md("**Age Group**"),
        `Reading for personal interest` = md("**Reading for Personal Interest**"),
        `Playing games` = md("**Playing Games**"),
        `Weightlifting or strength training` = md("**Weightlifting or Strength Training**"),
        `Participating in water sports` = md("**Participating in Water Sports**")
    ) |>
    fmt_number(c("Reading for personal interest", "Playing games", 
                 "Weightlifting or strength training", "Participating in water sports"),
               decimals = 0) |>
    cols_align(align = "center",
               columns = c("Reading for personal interest", "Playing games",
                           "Weightlifting or strength training", "Participating in water sports")) |>
    tab_header(
        title = md("**Time Spent on Selected Activities by Age Group (Among Participants Only)**"), 
        subtitle = "Of those who participated in activity"
    ) |>
    tab_source_note(
        source_note = md("*Source: [American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm), 2003-2024, U.S. Bureau of Labor Statistics*")
    )

```


```{r favorite-activities-by-age-group}

# time_by_age_group_table <- function(activity_name){
#   act_code <- activity_codes |>
#     filter(task_level3 == activity_name) |>
#     pull(code_level3)
#   
#   activities |> 
#     filter(level3_code == act_code) |>
#     right_join(
#       respondents |> select(participant_id, survey_weight),
#       join_by(participant_id)
#     ) |>
#     mutate(time_spent_min = coalesce(time_spent_min, 0L)) |>
#     left_join(
#       roster |> 
#         filter(relation == "Self") |>
#         select(participant_id, age),
#       join_by(participant_id)
#     ) |>
#     mutate(age_group = cut(age, breaks = c(0, 24, 34, 44, 54, 64, Inf),
#                             labels = c("Under 25", "25-34", "35-44", "45-54", "55-64", "65 and Over"))) |>
#     group_by(age_group) |>
#     summarise(avg_min = weighted.mean(time_spent_min, survey_weight)) |>
#     arrange(age_group) |>
#     gt() |>
#     cols_label(
#       age_group = md("**Age Group**"), 
#       avg_min = md("**Average Minutes per Day**")
#     ) |>
#     cols_align(align = "center", columns = c(age_group, avg_min)) |>
#     fmt_number(avg_min, decimals = 0) |>
#     tab_header(
#       title = md(glue("**Time Spent on {activity_name} by Age Group**")),
#       subtitle = glue("Average Minutes per Day {activity_name}")
#     ) |>
#     tab_source_note(
#       source_note = md("*Source: [American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm), 2003-2024, U.S. Bureau of Labor Statistics*")
#     )
# }
# 
# time_by_age_group_table("Reading for personal interest")
# time_by_age_group_table("Playing games")
# time_by_age_group_table("Weightlifting or strength training")
# time_by_age_group_table("Participating in water sports")

```


### Plots

**1. How does the amount of time spent on household activities change as Americans age?**

```{r household-activities-as-Americans-age}

household_by_age <- activities |>
  filter(level1_code == "020000") |>
  right_join(
    respondents |> select(participant_id, survey_weight),
    join_by(participant_id)
  ) |>
  mutate(time_spent_min = coalesce(time_spent_min, 0L)) |>
  left_join(
    roster |>
      filter(relation == "Self") |>
      select(participant_id, age),
    join_by(participant_id)
  ) |>
  group_by(age) |>
  summarise(avg_min = weighted.mean(time_spent_min, survey_weight, na.rm = TRUE))

household_act_by_age_table <- household_by_age |>
  ggplot(aes(x = age, y = avg_min)) +
  geom_line() +
  geom_smooth(method = "loess", formula = y ~ x) +
  scale_x_continuous(breaks = seq(0, 90, by = 10)) +
  labs(
    title = "Time Spent on Household Activities by Age",
    x = "Age",
    y = "Average Minutes per Day",
    caption = "Source: American Time Use Survey (ATUS), 2003-2024, U.S. Bureau of Labor Statistics"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

household_act_by_age_table


# household_by_age <- activities |>
#     filter(level1_code == "020000") |>
#     right_join(
#         respondents |> select(participant_id, survey_weight),
#         join_by(participant_id)
#     ) |>
#     mutate(time_spent_min = coalesce(time_spent_min, 0L)) |>
#     left_join(
#         roster |>
#             filter(relation == "Self") |>
#             select(participant_id, age),
#         join_by(participant_id)
#     ) |>
#     mutate(age_group = cut(age, breaks = c(0, 24, 34, 44, 54, 64, Inf),
#                            labels = c("Under 25", "25-34", "35-44",
#                                       "45-54", "55-64", "65 and over"))) |>
#     filter(!is.na(age_group)) |>
#     group_by(age_group) |>
#     summarise(avg_min = weighted.mean(time_spent_min, survey_weight, na.rm = TRUE))
# 
# household_by_age |>
#     ggplot(aes(x = age_group, y = avg_min)) +
#     geom_col() +
#     labs(
#         title = "Time Spent on Household Activities by Age Group",
#         x = "Age Group",
#         y = "Average Minutes per Day",
#         caption = "Source: American Time Use Survey (ATUS), 2003-2024, U.S. Bureau of Labor Statistics"
#     ) +
#     theme_minimal() +
#     theme(plot.title = element_text(hjust = 0.5))

```

**2. A major perk of working in education is the freedom to spend summers how you wish. What are the most popular activities for Americans employed in education during the summer months (June, July, and August)?**

***Note:** Most popular measured by most frequently reported activity.*

```{r education-employees-summer-activities}

education_summer_activities <- activities |>
    inner_join(
        respondents |>
            filter(in_education,
                   labor_force_status %in% c("Employed - at work", "Employed - absent"),
                   survey_month %in% c(6, 7, 8)) |>
            select(participant_id, survey_weight),
        join_by(participant_id)
    ) |>
    inner_join(
        activity_codes,
        join_by(level3_code == code_level3)
    ) |>
    group_by(task_level3) |>
    summarise(n_occurrences = n()) |>
    arrange(desc(n_occurrences)) |>
    slice_head(n = 10)

education_summer_activities_table <- education_summer_activities |>
    ggplot(aes(x = reorder(task_level3, n_occurrences), y = n_occurrences)) +
    geom_col() +
    coord_flip() +
    labs(
        title = "Most Popular Summer Activities for Education Workers",
        x = "Activity",
        y = "Number of Occurrences",
        caption = "Source: American Time Use Survey (ATUS), 2003-2024, U.S. Bureau of Labor Statistics"
    ) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5))

education_summer_activities_table

```


**3. It is often remarked that “education is a privilege wasted on the young”. Among respondents who report being enrolled in a higher-education program, how does the amount of time allocated to various educational activities (attending class, doing homework, studying, etc.) vary by age? Does it seem to be true that older students dedicate more time to their education?**

```{r educational-activity-time-by-age}

education_time_by_age <- activities |>
    filter(level1_code == "060000") |>
    right_join(
        respondents |>
            filter(college_enrolled) |>
            select(participant_id, survey_weight, age),
        join_by(participant_id)
    ) |>
    mutate(time_spent_min = coalesce(time_spent_min, 0L)) |>
    mutate(age_group = cut(age, breaks = c(0, 24, 34, 44, 54, 64, Inf),
                           labels = c("Under 25", "25-34", "35-44",
                                      "45-54", "55-64", "65 and over"))) |>
    filter(!is.na(age_group)) |>
    group_by(age_group) |>
    summarise(avg_min = weighted.mean(time_spent_min, survey_weight)) |>
    arrange(age_group)
# education_time_by_age <- activities |>
#   filter(level1_code == "060000") |>
#   right_join(
#     respondents |>
#       filter(college_enrolled) |>
#       select(participant_id, survey_weight, age),
#     join_by(participant_id)
#   ) |>
#   mutate(time_spent_min = coalesce(time_spent_min, 0L)) |>
#   group_by(age) |>
#   summarise(avg_min = weighted.mean(time_spent_min, survey_weight)) |>
#   arrange(age)

education_time_by_age_table <- education_time_by_age |>
    ggplot(aes(x = age_group, y = avg_min, fill = age_group)) +
    geom_col() +
    scale_fill_brewer(palette = "Set2") +
    labs(
        title = "Time Spent on Educational Activities by Age Group",
        subtitle = "Among respondents enrolled in higher education",
        x = "Age Group",
        y = "Average Minutes per Day",
        caption = "Source: American Time Use Survey (ATUS), 2003-2024, U.S. Bureau of Labor Statistics"
    ) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5))

education_time_by_age_table

```

Among respondents who report being enrolled in a higher education program, students who are younger on average spend more time on educational activities than older students. Students under 25 report the highest average time spent on educational activities, followed by students 25-34. Students between 35-44 report the lowest average time spent on educational activities. 

**4. A “bump plot” is an effective way to display ranking (ordinal) data and to show how it changes over time. Use a bump plot to show how the relative interest in different sports changes as a function of either age or income (your choice).**

```{r sports-by-income}

sports_by_income <- activities |>
    #filter by participation in sports
    filter(level2_code == "130100") |>
    inner_join(
        respondents |>
            filter(weekly_earnings > 0) |>
            select(participant_id, survey_weight, weekly_earnings),
        join_by(participant_id)
    ) |>
    inner_join(
        activity_codes |> select(code_level3, task_level3) |> distinct(),
        join_by(level3_code == code_level3)
    ) |>
    mutate(income_group = cut((weekly_earnings / 100) * 52,
                              breaks = c(0, 25000, 50000, 75000, 100000, Inf),
                              labels = c("Under $25k", "$25k-$50k", "$50k-$75k", "$75k-$100k", "Over $100k"))) |>
    filter(!is.na(income_group)) |>
    group_by(income_group, task_level3) |>
    summarise(n_occurrences = n(), .groups = "drop") |>
    group_by(income_group) |>
    #create a rank of each sport within each income group based on # of participation activity occurrence among respondent
    mutate(rank = rank(-n_occurrences)) |>
    #show sports with rank 10 or below (top 10 sports in each income group)
    filter(rank <= 10)

sports_by_income_table <- sports_by_income |>
    ggplot(aes(x = income_group, y = rank, color = task_level3, group = task_level3)) +
    geom_line() +
    geom_point(size = 3) +
    scale_y_reverse(breaks = seq(1, 10, by = 1)) +
    labs(
        title = "Relative Interest in Sports by Income Group",
        x = "Income Group",
        y = "Sport Rank",
        color = "Sport",
        caption = "Source: American Time Use Survey (ATUS), 2003-2024, U.S. Bureau of Labor Statistics"
    ) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5))

sports_by_income_table

```

I defined relative interest in the sports based on reported *participation* in sports across income groups. Rank is determined by the number of recorded activities respondents of each income group participated in and the top 10 sports are displayed. 

As income rises, interest in water sports and other playing sports steadily increases. With respondents earning under \$25k, watersports' rank starts out at 3 and doubles to 6 for those earning \$75k or more. Interest in other playing sports starts relatively high at 7 for those earning under \$25k and moderately rises to 9 at the highest income groups.

The sharpest declines as income grows are in running and biking, with both sports' ranks declining by 2 for respondents earning 50k and more.

Other trends of note: 

* Golfing is of no interest below \$50k and only becomes a sport of interest for higher earners at \$50-75k and higher. 

* Basketball has a relatively high ranking at 8 but only appears in the Under 25k group.

**5. How does the amount of time parents spend per child change as family size increases?**

```{r time-per-child}

childcare_by_family_size <- activities |>
    filter(level1_code == "030000") |>
    right_join(
        respondents |>
            filter(n_children_under18 > 0) |>
            select(participant_id, survey_weight, n_children_under18),
        join_by(participant_id)
    ) |>
    mutate(time_spent_min = coalesce(time_spent_min, 0L)) |>
    group_by(participant_id, survey_weight, n_children_under18) |>
    summarise(total_min = sum(time_spent_min), .groups = "drop") |>
    mutate(min_per_child = total_min / n_children_under18) |>
    group_by(n_children_under18) |>
    summarise(avg_min_per_child = weighted.mean(min_per_child, survey_weight))

childcare_by_family_size_table <- childcare_by_family_size |>
    ggplot(aes(x = n_children_under18, y = avg_min_per_child)) +
    geom_line() +
    geom_point(size = 3) +
    scale_x_continuous(breaks = seq(1, max(childcare_by_family_size$n_children_under18), by = 1)) +
    labs(
        title = "Time Spent per Child on Childcare by Family Size",
        x = "Number of Children Under 18",
        y = "Average Minutes per Child per Day",
        caption = "Source: American Time Use Survey (ATUS), 2003-2024, U.S. Bureau of Labor Statistics"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))

childcare_by_family_size_table

```



## Market Research on Time Use Patterns in Target Demographics

Code related to the final deliverable of the assignment goes here. 

``` {r market-research-on-demos}

#add activity labels to activities data frame
activities_labeled <- activities |>
    left_join(
        activity_codes |> select(code_level3, task_level1, task_level2, task_level3) |> distinct(),
        join_by(level3_code == code_level3)
    )

# Identify demographic groups

stay_at_home_dads <- respondents |>
    filter(sex == "M",
           labor_force_status == "Not in labor force",
           !is_retired,
           n_children_under18 > 0,
           age < 65)

college_undergraduates <- respondents |>
    filter(college_enrolled,
           age >= 18,
           age <= 22)

working_mothers <- respondents |>
    filter(sex == "F",
           labor_force_status %in% c("Employed - at work", "Employed - absent"),
           n_children_under18 > 0)



# Define target levels for each demographic group
sahd_targets <- tribble(
    ~task_level2,                                       ~target,
    "Sleeping",                                         480,
    "Caring for and Helping Household Children",        300,
    "Activities Related to Household Children's Education", 60,
    "Household Activities",                             120,
    "Participating in Sports, Exercise, or Recreation", 60,
    "Socializing and Communicating",                    60,
    "Relaxing and Leisure",                             60
)

undergrad_targets <- tribble(
    ~task_level2,                                       ~target,
    "Sleeping",                                         360,
    "Taking Class",                                     300,
    "Research or Homework",                             300,
    "Participating in Sports, Exercise, or Recreation", 60,
    "Socializing and Communicating",                    120,
    "Relaxing and Leisure",                             120
)

working_mothers_targets <- tribble(
    ~task_level2,                                                ~target,
    "Sleeping",                                                  480,
    "Working",                                                   480,
    "Caring for and Helping Household Children",                 120,
    "Household Activities",                                      60,
    "Participating in Sports, Exercise, or Recreation",          45,
    "Socializing and Communicating",                             30,
    "Relaxing and Leisure",                                      90,
    "Health-Related Self Care",                                  30
)

# Function to calculate typical levels and fraction hitting target
calc_typical <- function(demographic, activity_filter, targets) {
    activities_filtered <- demographic |>
        select(participant_id, survey_weight) |>
        left_join(activities_labeled, join_by(participant_id)) |>
        filter({{ activity_filter }}) |>
        mutate(time_spent_min = coalesce(time_spent_min, 0L),
               task_level2 = if_else(level1_code == "020000", task_level1, task_level2)) |>
        filter(!is.na(task_level2)) |>
        group_by(participant_id, survey_weight, task_level1, task_level2) |>
        summarise(total_min = sum(time_spent_min), .groups = "drop")
    
    typical <- activities_filtered |>
        group_by(task_level1, task_level2) |>
        summarise(
            low    = wtd.quantile(total_min, weights = survey_weight, probs = 0.25),
            medium = wtd.quantile(total_min, weights = survey_weight, probs = 0.50),
            high   = wtd.quantile(total_min, weights = survey_weight, probs = 0.75),
            avg    = weighted.mean(total_min, survey_weight),
            .groups = "drop"
        ) |>
        left_join(targets, join_by(task_level2))
    
    target_hit <- activities_filtered |>
        left_join(targets, join_by(task_level2)) |>
        filter(!is.na(target)) |>
        mutate(hits_target = total_min >= target) |>
        group_by(task_level1, task_level2, target) |>
        summarise(pct_hitting_target = weighted.mean(hits_target, survey_weight),
                  .groups = "drop")
    
    list(typical = typical, target_hit = target_hit)
}

# Calculate typical levels and fraction hitting target for each demographic group
sahd_results <- calc_typical(
    stay_at_home_dads,
    level1_code == "020000" |
        level2_code %in% c("030100", "030200", "120100", "120300", "130100") |
        level3_code == "010101",
    sahd_targets
)

undergrad_results <- calc_typical(
    college_undergraduates,
    level1_code == "060000" |
        level2_code %in% c("120100", "120300", "130100") |
        level3_code == "010101",
    undergrad_targets
)

working_mothers_results <- calc_typical(
    working_mothers,
    level1_code == "020000" |
        level2_code %in% c("030100", "050100", "120100", "120300", "130100", "010300") |
        level3_code == "010101",
    working_mothers_targets
)

##CHART - fraction of resp hitting target levels
#Stay at home dads hitting target levels

sahd_plot <- sahd_results$target_hit |>
    mutate(pct_hitting_target = pct_hitting_target * 100) |>
    ggplot(aes(x = reorder(task_level2, pct_hitting_target), y = pct_hitting_target)) +
    geom_col(fill = "steelblue") +
    geom_hline(yintercept = 50, linetype = "dashed", color = "red") +
    coord_flip() +
    labs(
        title = "Stay-at-Home Dads",
        x = "Activity",
        y = "% Hitting Target",
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))

undergrad_plot <- undergrad_results$target_hit |>
    mutate(pct_hitting_target = pct_hitting_target * 100) |>
    ggplot(aes(x = reorder(task_level2, pct_hitting_target), y = pct_hitting_target)) +
    geom_col(fill = "darkgreen") +
    geom_hline(yintercept = 50, linetype = "dashed", color = "red") +
    coord_flip() +
    labs(
        title = "College Undergraduates",
        x = "Activity",
        y = "% Hitting Target",
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))

working_mothers_plot <- working_mothers_results$target_hit |>
    mutate(pct_hitting_target = pct_hitting_target * 100) |>
    ggplot(aes(x = reorder(task_level2, pct_hitting_target), y = pct_hitting_target)) +
    geom_col(fill = "purple") +
    geom_hline(yintercept = 50, linetype = "dashed", color = "red") +
    coord_flip() +
    labs(
        title = "Working Mothers",
        x = "Activity",
        y = "% Hitting Target",
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))

sahd_plot

undergrad_plot

working_mothers_plot




```




College undergraduates are defined as respondents that are enrolled in college and between the ages 18-22 so your traditional "typical" undergraduate student.

Key Activities: 
Stay At Home Dads

* Sleep (Target level: 480 mins - 8 hours)
* Caring for and Helping Household Children (Target: 360 min - 6 hours)
* Household activities (Target: 2 hours)
* Exercise/Sports (participating) (Target - 1 hour)
* Socializing (Target: 1 hour)
* Relaxing and Leisure

College Undergraduates

* Educational activities (480 mins - 5 hours)
* Sleep (not really a priority but 8)
* Exercise/Sports (Target: 1 hour)
* Socializing (Target: 120 mins - 2 hours)
* Leisure/relaxing (Target: 60 mins - 1 hour)

Working Mothers

* Childcare (Target: 2 hours)
* Work 
* Exercise/sports (Target 45 mins)
* Sleep (Target: 480 min - 8 hours)
* Self-care (health related self care 10301 and other self care 10399)
* Household activities (Target: 1 hour)
* Socializing (Target: 30 mins)


::: {.callout-note title="AI Usage Statement" collapse="true"}

Claude was used to help write the code in this project and brainstorm key demographic activities, but all non-code text was generated without the use of any Generative AI tools.  

:::



------------------------------------------------------------------------

This work ©2025 by er0driguez was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #02](https://michael-weylandt.com/STA9750/mini/mini02.html)
