---
title: "Mini-Project 02: How Do You Do 'You Do You'?"
author: "Evelyn Rodriguez"
editor:
    mode: source
format:
    html:
        toc: true
        code-fold: true
        code-summary: "Display Code"
        code-overflow: wrap
        code-block-bg: true
        code-block-border-left: "#6610f2"
        highlight-style: "a11y"
---

```{r setup, include=FALSE}

library(tidyverse)
library(glue)
library(readxl)
library(httr2)
library(skimr)
library(dplyr)
library(scales)
library(gt)
library(Hmisc)
library(gridExtra)

```


## Introduction

The *[American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm)* is a joint effort between the US Bureau of Labor Statistics and the US Census Bureau to capture how Americans spend their time everyday. Over 20 years, Americans recorded what they did and in this mini-project I will analyze how that time was spent in order to find and compare time usage patterns across key demographic groups. I will explore the ATUS data using dplyr tools to combine and manipulate the data to extract important demographic data. I will analyze the data with complex analytic pipelines, which I will then summarize for a non-technical audience with data visualization. 

In my final analysis, I will pose as a product manager for a time management app and create a market research article that applies this data and focuses on the time use patterns in three target demographics.


## Data Acquisition and Preparation

### Data Acquisition from ATUS

Data is acquired from ATUS resources websites, cleaned and combined into data frames that identify survey respondents, survey activities, their corresponding activity codes and important demographic information. 

```{r data-acquisitions}
#| warning: false

load_atus_data <- function(file_base = c("resp", "rost", "act", "cps")){
    if(!dir.exists(file.path("data", "mp02"))){
        dir.create(file.path("data", "mp02"), showWarnings=FALSE, recursive=TRUE)
    }
    
    file_base <- match.arg(file_base)
    
    file_name_out <- file.path("data", 
                               "mp02", 
                               glue("atus{file_base}_0324.dat"))
        
    if(!file.exists(file_name_out)){
        
        url_end <- glue("atus{file_base}-0324.zip")
        
        temp_zip <- tempfile(fileext=".zip")
        temp_dir <- tools::file_path_sans_ext(temp_zip)
        
        request("https://www.bls.gov") |>
            req_url_path("tus", "datafiles", url_end) |>
            req_headers(`User-Agent` = "Mozilla/5.0 (Macintosh; Intel Mac OS X 10.15; rv:149.0) Gecko/20100101 Firefox/149.0") |>
            req_perform(temp_zip)
    
        unzip(temp_zip, exdir = temp_dir)
        
        temp_file <- file.path(temp_dir, basename(file_name_out))
        
        file.copy(temp_file, file_name_out)
    }
    
    file <- read_csv(file_name_out, 
                     show_col_types = FALSE)
    
    switch(file_base, 
           #add cps data to determine if respondent is married
           cps=file |>
             filter(TULINENO == 1) |>
             mutate(marital_status = case_match(PEMARITL,
               1 ~ "Married - spouse present",
               2 ~ "Married - spouse absent",
               3 ~ "Widowed",
               4 ~ "Divorced",
               5 ~ "Separated",
               6 ~ "Never married"),
             #check if respondents are married
               has_spouse = marital_status %in% c("Married - spouse present", "Married - spouse absent"),
               is_retired = PEMLR == 5,
             #check if respondent works in education by main job occupation code
               in_education = PRDTOCC1 == 8)
             ,
           resp=file |> 
                  rename(survey_weight = TUFNWGTP, 
                         time_alone = TRTALONE, 
                         survey_year = TUYEAR,
                         survey_month = TUMONTH,
                         weekly_earnings = TRERNWA) |>
                  #add a column that counts number of children from roster file
                  left_join(
                    load_atus_data("rost") |>
                      filter(relation == "Self") |>
                      select(participant_id, age, sex), 
                    join_by(TUCASEID == participant_id)
                  ) |>
                  left_join(
                    load_atus_data("rost") |>
                      filter(age < 18) |>
                      count(participant_id, name = "n_children_under18"),
                    join_by(TUCASEID == participant_id)
                  ) |>
                  # pull in CPS data to add marital_status and has_spouse columns
                  left_join(
                    load_atus_data("cps"),
                    join_by(TUCASEID == participant_id)
                  ) |>
                  #coalesce will turn NA values into 0
                  mutate(n_children_under18 = coalesce(n_children_under18, 0L),
                         labor_force_status = case_match(TELFS,
                            1 ~ "Employed - at work",
                            2 ~ "Employed - absent",
                            3 ~ "Unemployed - on layoff",
                            4 ~ "Unemployed - looking",
                            5 ~ "Not in labor force"),
                         #creates column indicating enrollment
                         school_enrollment = case_match(TESCHENR,
                           1 ~ "Enrolled",
                           2 ~ "Not Enrolled",
                           -1 ~ "Blank"),
                         #create demographic variable that checks if enrolled and teschlvl matches college/university
                         college_enrolled = TESCHENR == 1 & TESCHLVL == 2)
           ,
           rost=file |>
                  mutate(sex = case_match(TESEX, 1 ~ "M", 2 ~ "F"),
                         #adding relation to respondent
                         relation = case_match(TERRP,
                           c(18,19) ~ "Self",
                           20 ~ "Spouse",
                           21 ~ "Unmarried partner",
                           22 ~ "Own household child",
                           23 ~ "Grandchild",
                           24 ~ "Parent",
                           25 ~ "Brother/sister",
                           26 ~ "Other related person",
                           27 ~ "Foster child",
                           28 ~ "Housemate/roommate",
                           29 ~ "Roomer/boarder",
                           30 ~ "Other nonrelative",
                           40 ~ "Own nonhousehold child < 18")) |>
                  rename(age = TEAGE), 
           act =file |>
                mutate(activity_n_of_day = TUACTIVITY_N, 
                       time_spent_min = TUACTDUR24, 
                       start_time = TUSTARTTIM,
                       stop_time = TUSTOPTIME, 
                       level1_code = paste0(TRTIER1P, "0000"), 
                       level2_code = paste0(TRTIER2P, "00"),
                       level3_code = TRCODEP,
                       #location of activity
                       location = case_match(TEWHERE,
                         1 ~ "Home",
                         2 ~ "Workplace",
                         3 ~ "Someone else's home",
                         4 ~ "Restaurant bar",
                         5 ~ "Place of worship",
                         6 ~ "Grocery store",
                         7 ~ "Other store/mall",
                         8 ~ "School",
                         9 ~ "Outdoors",
                         10 ~ "Library",
                         11 ~ "Other place",
                         12 ~ "Car, truck, or motorcycle (driver)",
                         13 ~ "Car, truck, or motorcycle (passenger)",
                         14 ~ "Walking",
                         15 ~ "Bus",
                         16 ~ "Subway/train",
                         17 ~ "Bicycle",
                         18 ~ "Boat/ferry",
                         19 ~ "Taxi/limousine service",
                         20 ~ "Airplane",
                         21 ~ "Other transportation",
                         30 ~ "Bank",
                         31 ~ "Gym/health club",
                         32 ~ "Post office",
                         89 ~ "Unspecificed place",
                         99 ~ "Unspecified transportation",
                         (-1) ~ "Blank"
                       )) 
           ) |> 
                rename(participant_id = TUCASEID) |>
                select(matches("[:lower:]", 
                               ignore.case=FALSE))
}

load_atus_activities <- function(){
    if(!dir.exists(file.path("data", "mp02"))){
        dir.create(file.path("data", "mp02"), showWarnings=FALSE, recursive=TRUE)
    }
    
    dest_file <- file.path("data",
                           "mp02",
                           "atus_activity_codes.csv")
    
    if(!file.exists(dest_file)){
        download.file("https://michael-weylandt.com/STA9750/mini/atus_activity_codes.csv", 
                      quiet=TRUE, 
                      destfile=dest_file, 
                      mode="wb")
    }
    
    read_csv(dest_file, show_col_types = FALSE) 
}

respondents  <- load_atus_data("resp")
cps <- load_atus_data("cps") 
roster       <- load_atus_data("rost")
activities   <- load_atus_data("act")
activity_codes <- load_atus_activities()

```

## Exploratory Analysis

### Inline Values

**1. How many different respondents have answered an ATUS survey since 2003?**

```{r unique-respondents}

unique_respondents <- respondents |> 
  distinct(participant_id) |>
  nrow()

```

There are `r comma(unique_respondents)` respondents that answered the ATUS survey since 2003.


**2. Americans enjoy watching a variety of sports. How many different sports does ATUS ask about watching as a “Level 3” task?**

```{r number-of-sports-watched}

n_sports_watched <- activity_codes |>
  filter(code_level2 == 130200) |>
  summarise(n = n()) |>
  pull(n)

```

ATUS asks about watching `r n_sports_watched` different sports as a Level 3 task. 

**3. Approximately what percent of Americans are retired?**

```{r percent-retired}

percent_retired <- respondents |>
  summarise(pct_retired = weighted.mean(is_retired, survey_weight)) |>
  pull(pct_retired)

```

Approximately `r percent(percent_retired)` of Americans are retired.


**4. On average, how many hours do Americans sleep per night?**

```{r average-sleep-hours}

avg_sleep_hours <- activities |>
  #filtering only on level 3 code 10101 since we don't know if Other Sleep includes actually sleeping
  filter(level3_code == "010101") |>
  group_by(participant_id) |>
  summarise(total_sleep_min = sum(time_spent_min)) |>
  left_join(
    respondents |> select(participant_id, survey_weight),
    join_by(participant_id)
  ) |>
  summarise(avg_sleep_hours = weighted.mean(total_sleep_min, survey_weight) / 60) |>
  pull(avg_sleep_hours)

```

On average, Americans sleep `r round(avg_sleep_hours, 1)` hours per night.

**5. On average, how many hours do Americans with one or more children in the home spend taking care of their children (all activities including healthcare and involvement in their children’s education)?**

```{r average-childcare-hours}

avg_childcare_hours <- activities |>
  filter(
    level1_code == "030000",
    level2_code %in% c("030100", "030200", "030300")) |>
  group_by(participant_id) |>
  summarise(total_childcare_min = sum(time_spent_min)) |>
  #right join ensures we keep all respondents with children even if they recorded zero childcare activities
  #this will prevent an upward bias in the average
  right_join(
    respondents |>
      filter(n_children_under18 >= 1) |>
      select(participant_id, survey_weight),
    join_by(participant_id)
  ) |>
  mutate(total_childcare_min = coalesce(total_childcare_min, 0L)) |>
  summarise(avg_childcare_hours = weighted.mean(total_childcare_min, survey_weight) / 60) |>
  pull(avg_childcare_hours)

```

On average, Americans with one or more children spend `r round(avg_childcare_hours, 2)` hours taking care of their children. 


### Tables

**1. Do retired persons spend more time on lawn-care than people who have full-time employment?**

```{r lawn-care-time}

lawn_care <- activities |>
  filter(
    level2_code == "020500",
    #filter only one level 3 activity code that specifically talks about lawns since we don't know what is included in Other category
    level3_code == "020501"
  ) |>
  group_by(participant_id) |>
  summarise(total_lawn_min = sum(time_spent_min)) |>
  right_join(
    respondents |>
      filter(is_retired | labor_force_status == "Employed - at work") |>
      select(participant_id, survey_weight, is_retired, labor_force_status), 
    join_by(participant_id)
  ) |>
  mutate(total_lawn_min = coalesce(total_lawn_min, 0L)) |>
  group_by(is_retired) |>
  summarise(avg_lawn_hours = weighted.mean(total_lawn_min, survey_weight) / 60)

#table
lawn_care_table <- lawn_care |>
  mutate(is_retired = if_else(is_retired, "Retired", "Employed Full-Time")) |>
  gt() |>
  cols_label(
    is_retired = md("**Employment Status**"),
    avg_lawn_hours = md("**Average Hours Spent<br>on Lawn Care**")
  ) |>
  cols_align(align = "center", columns = avg_lawn_hours) |>
  fmt_number(avg_lawn_hours, decimals = 2) |>
  tab_header(
    title = md("**Lawn Care Hours by Employment Status**"), 
    subtitle = "Average Time Spent on Lawn-Care for Retired vs. Full-Time Employed Persons"
  ) |>
  tab_source_note(
    source_note = md("*Source: [American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm), 2003-2024, U.S. Bureau of Labor Statistics*")
  )

lawn_care_table

```

On average, retired people spend more time on lawn-care than full-time employed people.


**2. How does the amount of time spent on social and recreational activities typically compare for married and unmarried people?**

```{r social-and-recreational-activities-married-vs-unmarried}

social_rec <- activities |>
  filter(level1_code %in% c("120000", "130000")) |>
  group_by(participant_id) |>
  summarise(total_social_rec_min = sum(time_spent_min)) |>
  right_join(
    respondents |>
      select(participant_id, survey_weight, has_spouse),
    join_by(participant_id)
  ) |>
  mutate(total_social_rec_min = coalesce(total_social_rec_min, 0L)) |>
  group_by(has_spouse) |>
  summarise(avg_social_rec_hours = weighted.mean(total_social_rec_min, survey_weight) / 60)

social_rec_table <- social_rec |>
  mutate(has_spouse = if_else(has_spouse, "Married", "Unmarried")) |>
  gt() |>
  cols_label(
    has_spouse = md("**Marital Status**"),
    avg_social_rec_hours = md("**Avg. Hours Spent on<br>Social/Recreational Activities**")
  ) |>
  cols_align(align = "center", columns = avg_social_rec_hours) |>
  fmt_number(avg_social_rec_hours, decimals = 2) |>
  tab_header(
    title = md("**Social and Recreational Activity Time by Marital Status**"), 
    subtitle = md("Average Time Spent on Social and Recreational Activities for Married vs. Non-Married People")
  ) |>
  tab_source_note(
    source_note = md("*Source: [American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm), 2003-2024, U.S. Bureau of Labor Statistics*")
  )

social_rec_table
  

```

Married people spend  less time on social and recreational activities as compared to unmarried people. 

**3. Where do high-earning Americans spend the majority of their time?**

***Note:** High earners are defined as those earning above the 75th percentile for weekly earnings*


```{r high-earner-time}

high_earner_threshold <- respondents |>
  filter(weekly_earnings > 0) |>
  #define high earner threshold as someone earning above the 75th percentile
  summarise(threshold = quantile(weekly_earnings, 0.75)) |>
  pull(threshold)

high_earner_location <- activities |>
  right_join(
    respondents |>
      filter(weekly_earnings >= high_earner_threshold) |>
      select(participant_id, survey_weight),
    join_by(participant_id)
  ) |>
  #remove rows where location is missing or blank
  filter(!is.na(location), location != "Blank") |>
  group_by(location) |>
  summarise(avg_hours = weighted.mean(time_spent_min, survey_weight) / 60) |>
  arrange(desc(avg_hours))

high_earner_location_table <- high_earner_location |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(
    location = md("**Location**"),
    avg_hours = md("**Average Hours per Day**")
  ) |>
  cols_align(align = "center", columns = avg_hours) |>
  fmt_number(avg_hours, decimals = 2) |>
  tab_header(
    title = md("**Where Do High-Earning Americans Spend Their Time**"), 
    subtitle = "Locations where respondents above the 75th percentile of weekly earnings spend the majority of their time"
  ) |>
  tab_source_note(
    source_note = md("*Source: [American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm), 2003-2024, U.S. Bureau of Labor Statistics*")
  )

high_earner_location_table  
  

```

High-earning Americans spend the majority of their time in the workplace, on an airplane or in a place of worship. 



**4. What are the most common activities reported by Americans when on a bus, train, subway, boat, or ferry?**

***Note:** Most common, in this case, means most frequently reported activity as opposed to activity with most reported time spent on activity*

```{r common-transit-activities}

transit_activities <- activities |>
  filter(location %in% c("Bus", "Subway/train", "Boat/ferry")) |>
  left_join(
    activity_codes |> select(code_level3, task_level3) |> distinct(), 
    join_by(level3_code == code_level3)
  ) |>
  filter(!is.na(task_level3)) |>
  group_by(task_level3) |>
  summarise(n_occurrences = n()) |>
  arrange(desc(n_occurrences)) |>
  slice_head(n = 10)

transit_activities_table <- transit_activities |>
  gt() |>
  cols_label(
    task_level3 = md("**Activity**"),
    n_occurrences = md("**Number of Occurrences**")
  ) |>
  cols_align(align = "center", columns = n_occurrences) |>
  tab_header(
    title = md("**Most Common Activities on Public Transportation**"),
    subtitle = "Activities on Bus, Subway, Train, Boat or Ferry"
  ) |>
  tab_source_note(
    source_note = md("*Source: [American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm), 2003-2024, U.S. Bureau of Labor Statistics*")
  )

transit_activities_table

```

The most common activities reported when on a bus, train, subway, boat or ferry are traveling to work, traveling to class, or traveling to eat or drink. 

**5. How much time per day do Americans spend on your four favorite activities and how does this vary by age group?**

**Evelyn's Favorite Activities:**

* Reading *(ATUS Level 3 Task Name: Reading for personal interest)*

* Playing Video and Board Games *(ATUS Level 3 Task Name: Playing games)*

* Lifting at the Gym *(ATUS Level 3 Task Name: Weightlifting or strength training)*

* Scuba Diving *(ATUS Level 3 Task Name: Participating in water sports)*

***Note:** Analysis in first table includes those who reported 0 time spent on these activities so we can see prevalence in the general population*

```{r favorite-activities-by-age-group-one-table}

c("Reading for personal interest", "Playing games", "Weightlifting or strength training", "Participating in water sports") |>
    map(\(activity_name) {
        code <- activity_codes |>
            filter(task_level3 == activity_name) |>
            pull(code_level3)
        
        activities |>
            filter(level3_code == code) |>
            right_join(
                respondents |> select(participant_id, survey_weight),
                join_by(participant_id)
            ) |>
            mutate(time_spent_min = coalesce(time_spent_min, 0L)) |>
            left_join(
                roster |>
                    filter(relation == "Self") |>
                    select(participant_id, age),
                join_by(participant_id)
            ) |>
            mutate(age_group = cut(age, breaks = c(0, 24, 34, 44, 54, 64, Inf),
                                   labels = c("Under 25", "25-34", "35-44",
                                              "45-54", "55-64", "65 and over")),
                   activity = activity_name) |>
            group_by(activity, age_group) |>
            summarise(avg_min = weighted.mean(time_spent_min, survey_weight),
                      .groups = "drop")
    }) |>
    list_rbind() |>
    pivot_wider(names_from = activity, values_from = avg_min) |>
    gt() |>
    cols_label(
        age_group = md("**Age Group**"), 
        `Reading for personal interest` = md("**Reading**"),
        `Playing games` = md("**Playing Games**"),
        `Weightlifting or strength training` = md("**Weightlifting or<br>Strength Training**"), 
        `Participating in water sports` = md("**Participating in<br>Water Sports**")
    ) |>
    fmt_number(c("Reading for personal interest", "Playing games", "Weightlifting or strength training", "Participating in water sports"), 
               decimals = 0) |>
    cols_align(align = "center", 
               columns = c("Reading for personal interest", "Playing games", "Weightlifting or strength training", "Participating in water sports")) |>
    tab_header(
        title = md("**Time Spent on Evelyn's Favorite Activities by Age Group**"), 
        subtitle = "Average Minutes per Day for Each Activity"
    ) |>
    tab_source_note(
        source_note = md("*Source: [American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm), 2003-2024, U.S. Bureau of Labor Statistics*")
    )


c("Reading for personal interest", "Playing games", "Weightlifting or strength training", "Participating in water sports") |>
    map(\(activity_name) {
        code <- activity_codes |>
            filter(task_level3 == activity_name) |>
            pull(code_level3)
        
        activities |>
            filter(level3_code == code) |>
            inner_join(
                respondents |> select(participant_id, survey_weight),
                join_by(participant_id)
            ) |>
            left_join(
                roster |>
                    filter(relation == "Self") |>
                    select(participant_id, age),
                join_by(participant_id)
            ) |>
            mutate(age_group = cut(age, breaks = c(0, 24, 34, 44, 54, 64, Inf),
                                   labels = c("Under 25", "25-34", "35-44",
                                              "45-54", "55-64", "65 and over")),
                   activity = activity_name) |>
            group_by(activity, age_group) |>
            summarise(avg_min = weighted.mean(time_spent_min, survey_weight),
                      .groups = "drop")
    }) |>
    list_rbind() |>
    pivot_wider(names_from = activity, values_from = avg_min) |>
    gt() |>
    cols_label(
        age_group = md("**Age Group**"),
        `Reading for personal interest` = md("**Reading for Personal Interest**"),
        `Playing games` = md("**Playing Games**"),
        `Weightlifting or strength training` = md("**Weightlifting or Strength Training**"),
        `Participating in water sports` = md("**Participating in Water Sports**")
    ) |>
    fmt_number(c("Reading for personal interest", "Playing games", 
                 "Weightlifting or strength training", "Participating in water sports"),
               decimals = 0) |>
    cols_align(align = "center",
               columns = c("Reading for personal interest", "Playing games",
                           "Weightlifting or strength training", "Participating in water sports")) |>
    tab_header(
        title = md("**Time Spent on Selected Activities by Age Group (Among Participants Only)**"), 
        subtitle = "Of those who participated in activity"
    ) |>
    tab_source_note(
        source_note = md("*Source: [American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm), 2003-2024, U.S. Bureau of Labor Statistics*")
    )

```


```{r favorite-activities-by-age-group}

# time_by_age_group_table <- function(activity_name){
#   act_code <- activity_codes |>
#     filter(task_level3 == activity_name) |>
#     pull(code_level3)
#   
#   activities |> 
#     filter(level3_code == act_code) |>
#     right_join(
#       respondents |> select(participant_id, survey_weight),
#       join_by(participant_id)
#     ) |>
#     mutate(time_spent_min = coalesce(time_spent_min, 0L)) |>
#     left_join(
#       roster |> 
#         filter(relation == "Self") |>
#         select(participant_id, age),
#       join_by(participant_id)
#     ) |>
#     mutate(age_group = cut(age, breaks = c(0, 24, 34, 44, 54, 64, Inf),
#                             labels = c("Under 25", "25-34", "35-44", "45-54", "55-64", "65 and Over"))) |>
#     group_by(age_group) |>
#     summarise(avg_min = weighted.mean(time_spent_min, survey_weight)) |>
#     arrange(age_group) |>
#     gt() |>
#     cols_label(
#       age_group = md("**Age Group**"), 
#       avg_min = md("**Average Minutes per Day**")
#     ) |>
#     cols_align(align = "center", columns = c(age_group, avg_min)) |>
#     fmt_number(avg_min, decimals = 0) |>
#     tab_header(
#       title = md(glue("**Time Spent on {activity_name} by Age Group**")),
#       subtitle = glue("Average Minutes per Day {activity_name}")
#     ) |>
#     tab_source_note(
#       source_note = md("*Source: [American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm), 2003-2024, U.S. Bureau of Labor Statistics*")
#     )
# }
# 
# time_by_age_group_table("Reading for personal interest")
# time_by_age_group_table("Playing games")
# time_by_age_group_table("Weightlifting or strength training")
# time_by_age_group_table("Participating in water sports")

```


### Plots

**1. How does the amount of time spent on household activities change as Americans age?**

```{r household-activities-as-Americans-age}

household_by_age <- activities |>
  filter(level1_code == "020000") |>
  right_join(
    respondents |> select(participant_id, survey_weight),
    join_by(participant_id)
  ) |>
  mutate(time_spent_min = coalesce(time_spent_min, 0L)) |>
  left_join(
    roster |>
      filter(relation == "Self") |>
      select(participant_id, age),
    join_by(participant_id)
  ) |>
  group_by(age) |>
  summarise(avg_min = weighted.mean(time_spent_min, survey_weight, na.rm = TRUE))

household_act_by_age_table <- household_by_age |>
  ggplot(aes(x = age, y = avg_min)) +
  geom_line() +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
  scale_x_continuous(breaks = seq(0, 90, by = 10)) +
  labs(
    title = "Time Spent on Household Activities by Age",
    x = "Age",
    y = "Average Minutes per Day",
    caption = "Source: American Time Use Survey (ATUS), 2003-2024, U.S. Bureau of Labor Statistics"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

household_act_by_age_table

```

Generally, the amount of time spent on household activities increases as Americans age starting with ~33 minutes per day on average at 20 years old and peaking at 65-70 years old with ~44 minutes per day. After this point, time spent on household activities seems to decline again. 

**2. A major perk of working in education is the freedom to spend summers how you wish. What are the most popular activities for Americans employed in education during the summer months (June, July, and August)?**

```{r education-employees-summer-activities}

education_summer_activities <- activities |>
    inner_join(
        respondents |>
            filter(in_education,
                   labor_force_status %in% c("Employed - at work", "Employed - absent"),
                   survey_month %in% c(6, 7, 8)) |>
            select(participant_id, survey_weight),
        join_by(participant_id)
    ) |>
    inner_join(
        activity_codes,
        join_by(level3_code == code_level3)
    ) |>
    group_by(task_level3) |>
    summarise(n_occurrences = n()) |>
    arrange(desc(n_occurrences)) |>
    slice_head(n = 10)

education_summer_activities_table <- education_summer_activities |>
    ggplot(aes(x = reorder(task_level3, n_occurrences), y = n_occurrences)) +
    geom_col() +
    coord_flip() +
    labs(
        title = "Most Popular Summer Activities for Education Workers",
        x = "Activity",
        y = "Number of Occurrences",
        caption = "Source: American Time Use Survey (ATUS), 2003-2024, U.S. Bureau of Labor Statistics"
    ) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5))

education_summer_activities_table

```

The top three most popular activities during the summer months for American employed in education are sleeping, eating/drinking, and washing/dressing/grooming oneself. Most  popular is measured by most frequently reported activity.

**3. It is often remarked that “education is a privilege wasted on the young”. Among respondents who report being enrolled in a higher-education program, how does the amount of time allocated to various educational activities (attending class, doing homework, studying, etc.) vary by age? Does it seem to be true that older students dedicate more time to their education?**

```{r educational-activity-time-by-age}

education_time_by_age <- activities |>
    filter(level1_code == "060000") |>
    right_join(
        respondents |>
            filter(college_enrolled) |>
            select(participant_id, survey_weight, age),
        join_by(participant_id)
    ) |>
    mutate(time_spent_min = coalesce(time_spent_min, 0L)) |>
    mutate(age_group = cut(age, breaks = c(0, 24, 34, 44, 54, 64, Inf),
                           labels = c("Under 25", "25-34", "35-44",
                                      "45-54", "55-64", "65 and over"))) |>
    filter(!is.na(age_group)) |>
    group_by(age_group) |>
    summarise(avg_min = weighted.mean(time_spent_min, survey_weight)) |>
    arrange(age_group)

education_time_by_age_table <- education_time_by_age |>
    ggplot(aes(x = age_group, y = avg_min, fill = age_group)) +
    geom_col() +
    scale_fill_brewer(palette = "Set2") +
    scale_y_continuous(breaks = seq(0, max(education_time_by_age$avg_min), by = 10)) +
    labs(
        title = "Time Spent on Educational Activities by Age Group",
        subtitle = "Among respondents enrolled in higher education",
        x = "Age Group",
        y = "Average Minutes per Day",
        caption = "Source: American Time Use Survey (ATUS), 2003-2024, U.S. Bureau of Labor Statistics"
    ) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5))

education_time_by_age_table

```

Among respondents enrolled in a higher education program, students who are younger on average spend more time on educational activities than older students. Students under 25 report the highest average time spent on educational activities with ~87 minutes per day on average, followed by students 25-34. Students between 35-44 report the lowest average time spent at ~78 minutes. 

**4. A “bump plot” is an effective way to display ranking (ordinal) data and to show how it changes over time. Use a bump plot to show how the relative interest in different sports changes as a function of either age or income (your choice).**

```{r sports-by-income}

sports_by_income <- activities |>
    #filter by participation in sports
    filter(level2_code == "130100") |>
    inner_join(
        respondents |>
            filter(weekly_earnings > 0) |>
            select(participant_id, survey_weight, weekly_earnings),
        join_by(participant_id)
    ) |>
    inner_join(
        activity_codes |> select(code_level3, task_level3) |> distinct(),
        join_by(level3_code == code_level3)
    ) |>
    mutate(income_group = cut((weekly_earnings / 100) * 52,
                              breaks = c(0, 25000, 50000, 75000, 100000, Inf),
                              labels = c("Under $25k", "$25k-$50k", "$50k-$75k", "$75k-$100k", "Over $100k"))) |>
    filter(!is.na(income_group)) |>
    group_by(income_group, task_level3) |>
    summarise(n_occurrences = n(), .groups = "drop") |>
    group_by(income_group) |>
    #create a rank of each sport within each income group based on # of participation activity occurrence among respondent
    mutate(rank = rank(-n_occurrences)) |>
    #show sports with rank 10 or below (top 10 sports in each income group)
    filter(rank <= 10)

sports_by_income_table <- sports_by_income |>
    ggplot(aes(x = income_group, y = rank, color = task_level3, group = task_level3)) +
    geom_line() +
    geom_point(size = 3) +
    scale_y_reverse(breaks = seq(1, 10, by = 1)) +
    labs(
        title = "Relative Interest in Sports by Income Group",
        x = "Income Group",
        y = "Sport Rank",
        color = "Sport",
        caption = "Source: American Time Use Survey (ATUS), 2003-2024, U.S. Bureau of Labor Statistics"
    ) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5))

sports_by_income_table

```

**Note:** I defined relative interest in the sports based on reported *participation* in sports across income groups. Rank is determined by the number of recorded activities respondents participated in and the top 10 sports are displayed.

As income rises, interest in water sports and other playing sports steadily increases. With respondents earning under \$25k, watersports' rank starts out at 3 and doubles to 6 for those earning \$75k or more. Interest in other playing sports starts relatively high at 7 for those earning under \$25k and moderately rises to 9 at the highest income groups.

The sharpest declines as income grows are in running and biking, with both sports' ranks declining by 2 for respondents earning 50k and more.

Other trends of note: 

* Golfing is of no interest below \$50k and only becomes a sport of interest for higher earners at \$50-75k and higher. 

* Basketball has a relatively high ranking at 8 but only appears in the Under 25k group.

**5. How does the amount of time parents spend per child change as family size increases?**

```{r time-per-child}

childcare_by_family_size <- activities |>
    filter(level1_code == "030000") |>
    right_join(
        respondents |>
            filter(n_children_under18 > 0) |>
            select(participant_id, survey_weight, n_children_under18),
        join_by(participant_id)
    ) |>
    mutate(time_spent_min = coalesce(time_spent_min, 0L)) |>
    group_by(participant_id, survey_weight, n_children_under18) |>
    summarise(total_min = sum(time_spent_min), .groups = "drop") |>
    mutate(min_per_child = total_min / n_children_under18) |>
    group_by(n_children_under18) |>
    summarise(avg_min_per_child = weighted.mean(min_per_child, survey_weight))

childcare_by_family_size_table <- childcare_by_family_size |>
    ggplot(aes(x = n_children_under18, y = avg_min_per_child)) +
    geom_line() +
    geom_point(size = 3) +
    scale_x_continuous(breaks = seq(1, max(childcare_by_family_size$n_children_under18), by = 1)) +
    labs(
        title = "Time Spent per Child on Childcare by Family Size",
        x = "Number of Children Under 18",
        y = "Average Minutes per Child per Day",
        caption = "Source: American Time Use Survey (ATUS), 2003-2024, U.S. Bureau of Labor Statistics"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))

childcare_by_family_size_table

```

The amount of time parents spend per child steadily decreases from 45 minutes when parents have one child to 6 minutes when there are 9 children. After 9 children, time spent per child increases slightly to 15 minutes until they reach child 12, where there is a sharp decline down to almost 0 minutes. 


## Market Research on Time Use Patterns in Target Demographics

``` {r market-research-on-demos}

#add activity labels to activities data frame
activities_labeled <- activities |>
    left_join(
        activity_codes |> select(code_level3, task_level1, task_level2, task_level3) |> distinct(),
        join_by(level3_code == code_level3)
    )

# Identify demographic groups
stay_at_home_dads <- respondents |>
    filter(sex == "M",
           labor_force_status == "Not in labor force",
           !is_retired,
           n_children_under18 > 0,
           age < 65)

college_undergraduates <- respondents |>
    filter(college_enrolled,
           age >= 18,
           age <= 22)

working_mothers <- respondents |>
    filter(sex == "F",
           labor_force_status %in% c("Employed - at work", "Employed - absent"),
           n_children_under18 > 0,
           age < 65)


# Define target levels for time spent of key activities of each demographic group
sahd_targets <- tribble(
    ~task_level2,                                       ~target,
    "Sleeping",                                         480,
    "Caring for and Helping Household Children",        300,
    "Activities Related to Household Children's Education", 60,
    "Household Activities",                             120,
    "Participating in Sports, Exercise, or Recreation", 60,
    "Socializing and Communicating",                    60,
    "Relaxing and Leisure",                             60
)

undergrad_targets <- tribble(
    ~task_level2,                                       ~target,
    "Sleeping",                                         360,
    "Taking Class",                                     240,
    "Research or Homework",                             240,
    "Extracurricular School Activities (except sports)", 90,
    "Participating in Sports, Exercise, or Recreation", 90,
    "Socializing and Communicating",                    120,
    "Relaxing and Leisure",                             120
)

working_mothers_targets <- tribble(
    ~task_level2,                                                ~target,
    "Sleeping",                                                  480,
    "Working",                                                   420,
    "Caring for and Helping Household Children",                 120,
    "Household Activities",                                      60,
    "Participating in Sports, Exercise, or Recreation",          90,
    "Socializing and Communicating",                             30,
    "Relaxing and Leisure",                                      90,
    "Health-Related Self Care",                                  90
)

# Function to calculate typical levels and fraction hitting target
calc_typical <- function(demographic, activity_filter, targets) {
    activities_filtered <- demographic |>
        select(participant_id, survey_weight) |>
        left_join(activities_labeled, join_by(participant_id)) |>
        filter({{ activity_filter }}) |>
        mutate(time_spent_min = coalesce(time_spent_min, 0L),
          #if come across level1_code, will collapse all corresponding level2 tasks into one row
               task_level2 = if_else(level1_code == "020000", task_level1, task_level2)) |>
        filter(!is.na(task_level2)) |>
        group_by(participant_id, survey_weight, task_level1, task_level2) |>
        summarise(total_min = sum(time_spent_min), .groups = "drop")
    
    typical <- activities_filtered |>
        group_by(task_level1, task_level2) |>
        summarise(
            low    = wtd.quantile(total_min, weights = survey_weight, probs = 0.25),
            medium = wtd.quantile(total_min, weights = survey_weight, probs = 0.50),
            high   = wtd.quantile(total_min, weights = survey_weight, probs = 0.75),
            avg    = weighted.mean(total_min, survey_weight),
            .groups = "drop"
        ) |>
        left_join(targets, join_by(task_level2))
    
    target_hit <- activities_filtered |>
        left_join(targets, join_by(task_level2)) |>
        filter(!is.na(target)) |>
        mutate(hits_target = total_min >= target) |>
        group_by(task_level1, task_level2, target) |>
        summarise(pct_hitting_target = weighted.mean(hits_target, survey_weight),
                  .groups = "drop")
    
    list(typical = typical, target_hit = target_hit)
}

# Calculate typical levels and fraction hitting target for each demographic group
sahd_results <- calc_typical(
    stay_at_home_dads,
    level1_code == "020000" |
        level2_code %in% c("030100", "030200", "120100", "120300", "130100") |
        level3_code == "010101",
    sahd_targets
)

undergrad_results <- calc_typical(
    college_undergraduates,
        level2_code %in% c("060100", "060200", "060300", "120100", "120300", "130100") |
        level3_code == "010101",
    undergrad_targets
)

working_mothers_results <- calc_typical(
    working_mothers,
    level1_code == "020000" |
        level2_code %in% c("030100", "050100", "120100", "120300", "130100", "010300") |
        level3_code == "010101",
    working_mothers_targets
)


###CHARTS FOR FINAL DELIVERABLE 
###
##CHART - fraction of respondents hitting target levels for each key activity
#Stay at home dads hitting target levels

sahd_hitting_target_plot <- sahd_results$target_hit |>
    mutate(pct_hitting_target = pct_hitting_target * 100) |>
    ggplot(aes(x = reorder(task_level2, pct_hitting_target), y = pct_hitting_target)) +
    geom_col(fill = "steelblue") +
    geom_hline(yintercept = 50, linetype = "dashed", color = "red") +
    coord_flip() +
    labs(
        title = "Stay-at-Home Dads",
        x = "Activity",
        y = "% Hitting Target",
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))

#Undergrads hitting target levels
undergrad_hitting_target_plot <- undergrad_results$target_hit |>
    mutate(pct_hitting_target = pct_hitting_target * 100) |>
    ggplot(aes(x = reorder(task_level2, pct_hitting_target), y = pct_hitting_target)) +
    geom_col(fill = "darkgreen") +
    geom_hline(yintercept = 50, linetype = "dashed", color = "red") +
    coord_flip() +
    labs(
        title = "College Undergraduates",
        x = "Activity",
        y = "% Hitting Target",
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))

#working mothers hitting target demos
working_mothers_hitting_target_plot <- working_mothers_results$target_hit |>
    mutate(pct_hitting_target = pct_hitting_target * 100) |>
    ggplot(aes(x = reorder(task_level2, pct_hitting_target), y = pct_hitting_target)) +
    geom_col(fill = "purple") +
    geom_hline(yintercept = 50, linetype = "dashed", color = "red") +
    coord_flip() +
    labs(
        title = "Working Mothers",
        x = "Activity",
        y = "% Hitting Target",
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))


##CHART - Average time spent on common activities across demographics 
##
#Find common activities across the three demographics
common_activities_across_demos <- bind_rows(
    sahd_results$typical |> mutate(demographic = "Stay-at-Home Dads"),
    undergrad_results$typical |> mutate(demographic = "College Undergraduates"),
    working_mothers_results$typical |> mutate(demographic = "Working Mothers")
) |>
    filter(task_level2 %in% c("Sleeping", "Socializing and Communicating", 
                               "Relaxing and Leisure",
                               "Participating in Sports, Exercise, or Recreation")) |>
    ggplot(aes(x = task_level2, y = avg, fill = demographic)) +
    geom_col(position = "dodge") +
    coord_flip() +
    scale_fill_brewer(palette = "Set2") +
    labs(
        title = "Average Time Spent on Common Activities Between Demographic Groups",
        x = "Activity",
        y = "Average Minutes per Day",
        fill = "Demographic",
        caption = "Source: American Time Use Survey (ATUS), 2003-2024, U.S. Bureau of Labor Statistics"
    ) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5))


##CHART - Typical Time Use across demos
##
make_typical_table <- function(results, title) {
    results$typical |>
        select(-task_level1) |>
        gt() |>
        cols_label(
          task_level2 = md("**Activity**"),
          low = md("**Low**<br>(25th %ile)"),
          medium = md("**Medium**<br>(50th %ile)"),
          high = md("**High**<br>(75th %ile)"),
          avg = md("**Average**"),
          target = md("**Target**")
        ) |>
        cols_width(task_level2 ~ px(150)) |>
        fmt_number(columns = c(low, medium, high, avg, target), decimals = 0) |>
        cols_align(align = "center", columns = c(low, medium, high, avg, target)) |>
        #if target is greater than average it indicates an opportunity for users to push themselves
        tab_style(
          style = cell_text(color = "darkgreen"),
          locations = cells_body(
            columns = target,
            rows = target >= avg
          )
        ) |>
      #if target is less than average, it indicates an easy "win" for users
        tab_style(
          style = cell_text(color = "darkred"),
          locations = cells_body(
            columns = target, 
            rows = target <= avg
          )
        ) |>
        tab_header(
            title = md(glue("**{title}**")),
            subtitle = "Minutes per day"
        ) |>
        tab_source_note(
            source_note = "Source: American Time Use Survey (ATUS), 2003-2024, U.S. Bureau of Labor Statistics"
        )
}


```


What have you been up to? A common question among friends, coworkers, and acquaintances that give us just a little bit of insight into who we are. Luckily, the Bureau of Labor Statistics and US Census Bureau has taken some of the mystery out of this guessing game with its *[American Time Use Survey (ATUS)](https://www.bls.gov/tus/data/datafiles-0324.htm)*. With this treasure trove of time usage data from the past 20 years, we're able to dig into our key demographic of users and their daily habits so that we can answer the core questions of our user research: *Who are they?*, *What do they do?*, and *Are we engaging and realistic in our time goals for users?*.

### Who Are They?
Our core demographic thus far is made up of Stay at Home Dads, College Undergraduates and Working Mothers all with good reason to need a helping hand when it comes to making their time allocation work for their unique needs. 

We broke down and defined these three groups as follows: 

* **Stay at Home Dads** are males with one or more children who reported their labor force status as 'Not in the Labor Force' and who are not of retirement age (below 65) or retirement status. 

* **College undergraduates** are individuals of all genders enrolled in college and between the ages 18-22. This captures the traditional "typical" first time undergraduate student not far removed from high school and still figuring out how to manage their time wisely.

* **Working Mothers** are women with one or more children who reported a labor force status of 'Employed' and were not retirement age (below 65).

### What Do They Do?

For each of these groups, we've identified key activities that take up a significant amount of time in these users' days and that users would like to improve upon like hitting the gym for longer. To gauge how useful our application may be to these individuals, we analyzed how each of these groups have historically spent time on these activities on average and how they measure up to what we believe would be desired target times for these groups. 

Among these groups, we see that our app may be most useful for working mothers since we've identified more key activities and responsibilities in general for this group. When comparing working mothers' typical time usage against our key activity targets, we also identified the most opportunities for time improvement in activities such as caring for household children, health-related self care, and participating in sports/exercise. 

```{r typical-time-use-tables, echo = FALSE}

make_typical_table(sahd_results, "Typical Time Use for Stay-at-Home Dads")

make_typical_table(undergrad_results, "Typical Time Use for College Undergraduates")

make_typical_table(working_mothers_results, "Typical Time Use for Working Mothers")

```

### Are We Engaging and Realistic in Our Time Goals?

Our target time levels are designed to be close enough that users won't get discouraged by difficulty but far enough to ignite the fire of competition among their peers. When measuring up the fraction of users in each group who can hit their respective target time goals, we see that users in each group can easily achieve half or more of the intended time goals with two or three outstanding goals in each group. It's our hope that this balance will keep the completionists among us engaged and the more casual users more than happy with their outcomes. 

Undergraduates are the group with the least amount of activities where more than 50% of users can hit target levels, which may indicate difficulty keeping up with taking classes, doing homework and socializing. 

```{r hitting-target-levels-charts, echo=FALSE}

sahd_hitting_target_plot
undergrad_hitting_target_plot
working_mothers_hitting_target_plot

```

Among activities that are shared across all demographic groups, working mothers spend on average less time on all activities while stay at home dads seem to have the luxury of time at hand. These two groups may seem somewhat similar at the outset with many overlapping activities but we see that working moms aren't able to hit targets in `participate in sports/exercise` and `care for household children` even though their target thresholds are significantly lower than stay at home dads. Moreover, undergraduate students only slightly led in time spent socializing category and for the most part played second fiddle to stay at home dads, which makes sense since none of a student's main priorities (going to class or studying) are present here.  

```{r common-activities-across-demos, echo=FALSE}

common_activities_across_demos

```

These disparities highlight and suggest segmentation in prioritization of activities among these groups. Our approach is correct in using dynamic thresholds for target levels since each major group's needs are clearly different. Working moms may need to do more with less time so they need lower targets in certain activities to hit target levels and get a sense of accomplishment out of our application. Stay at home dads are able to handle higher thresholds due to them having more time in the day not being taken up by work. Undergraduate students want to work hard in school and are not very concerned with housework.

Overall, our target demographics are well defined as a great starting point for application launch. Time usage analysis across these groups indicates we will have a varied audience of users with differing priorities that will require dynamic tiers of target thresholds in order to keep these groups engaged, happy and challenged. 



::: {.callout-note title="AI Usage Statement" collapse="true"}

Claude was used to help write the code in this project and brainstorm key demographic activities, but all non-code text was generated without the use of any Generative AI tools.  

:::



------------------------------------------------------------------------

This work ©2025 by er0driguez was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #02](https://michael-weylandt.com/STA9750/mini/mini02.html)
