---
title: "Mini-Project 03 - Who Goes There? US Internal Migration and Implications for Congressional Reapportionment"
author: "Evelyn Rodriguez"
editor:
    mode: source
format:
    html:
        toc: true
        code-fold: true
        code-summary: "Display Code"
        code-overflow: wrap
        code-block-bg: true
        code-block-border-left: "#6610f2"
        highlight-style: "a11y"
---

```{r setup, include=FALSE}

library(tidyverse)
library(tidycensus)
library(glue)
library(readxl)
library(httr2)
library(skimr)
library(dplyr)
library(scales)
library(gt)
library(jsonlite)
library(patchwork)
library(shadowtext)

```

## Introduction

What will your state look like in 2030? Many inquiring politicos are hard at work figuring this out and in this mini-project, we will examine [migration flows](https://www.census.gov/topics/population/migration/guidance/migration-flows.html) within the US to forecast the size of state congressional delegations for the 2032 midterm elections. Using state-level and metro-level data from the US Census Bureau's [American Community Survey (ACS)](https://en.wikipedia.org/wiki/American_Community_Survey), I intend to paint a picture of what 2030 will hold for the United States and, in particular, New York. 

I'll acquire and prepare U.S. Census data from several sources (tidycensus, excel sheets, and census API) in order to model projections that predict state populations through 2030 and calculate what congressional reapportionment may be for New York. To wrap up my findings, I'll write an Advertising Strategy report that lays outs the key insights I've found in state and metro migration flows and how we can use those to build an advertising campaign that brings in enough New Yorkers for an additional congressional seat. 

## Data Acquisition and Preparation

### Tasks 1-5

* Acquire data from the `tidycensus` package in R to find total population by state from 2015 - 2024.
* Acquire data from Census Bureau Excel files for state-to-state migration flows 2023 and 2024.
* Acquire data via API request for metro-to-metro migration flows.

***Note:** Data for 2020 is skipped since regular 1-year ACS for 2020 was not released and is not available in `tidycensus`.*

```{r acs-data-acquisition}
#| warning: false 


acquire_acs_data <- function(dataset = c("population", "state_to_state_migration", "metro_to_metro_migration", "international_migration")) {
  
  # Validate dataset argument
  dataset <- match.arg(dataset)
  
  # Define and create the project cache directory if it doesn't already exist
  data_dir <- file.path("data", "mp03")
  if (!dir.exists(file.path("data", "mp03"))) {
    dir.create(file.path("data", "mp03"), showWarnings = FALSE, recursive = TRUE)
  }
  
  # Helper to convert state abbreviations to full names
  # Extends R's built-in state.abb/state.name with DC and Puerto Rico
  abb_to_name <- c(setNames(state.name, state.abb),
                   "DC" = "District of Columbia",
                   "PR" = "Puerto Rico")
  
  expand_state_name <- function(x) {
    ifelse(x %in% names(abb_to_name), abb_to_name[x], x)
  }
  
  switch(dataset,
    
    ## TASK 1 - STATE POPULATION AND BIRTH INFORMATION     
    # --- Population: ACS-1 state-level total population, 2015-2024 (no 2020) ---
    population = {
      
      acs_years <- c(2015:2019, 2021:2024)
      
      POP_ALL <- map(acs_years, function(yy) {
        year_cache_file <- file.path(data_dir, glue("state_population_{yy}.csv"))
        
        if (file.exists(year_cache_file)) {
          message(glue("Loading cached population data for {yy}"))
          return(read_csv(year_cache_file, show_col_types = FALSE))
        }
        
        message(glue("Downloading ACS-1 state population data for {yy}"))
        
        # B01003_001 is the ACS variable for total population
        state_pop_raw <- get_acs(
          variables   = "B01003_001",
          geography   = "state",
          year        = yy,
          survey      = "acs1",
          geometry    = FALSE,
          cache_table = TRUE
        ) |>
          mutate(year = yy) |>
          select(
            state      = NAME,
            population = estimate,
            year
          )
        
        # Cache the cleaned result to disk before returning
        write_csv(state_pop_raw, year_cache_file)
        state_pop_raw
        
      }) |> bind_rows()
      
      POP_ALL
    },
    
    ##TASK 4: COMBINED STATE TO STATE FLOWS
    # --- State-to-State Migration: combines 2024 and 2023 ACS-1 flows ---
    state_to_state_migration = {
      
      ##TASK 2: 2024 STATE TO STATE FLOWS
      # -- Internal function: 2024 flows and stationary population --
      state_to_state_migration_2024 <- function() {
        
        migration_url_2024  <- "https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2024/state-to-state-migration/State_to_State_Migration_Table_2024_T13.xlsx"
        migration_file_2024 <- file.path(data_dir, "State_to_State_Migration_Table_2024_T13.xlsx")
        
        if (!file.exists(migration_file_2024)) {
          message("Downloading 2024 state-to-state migration file...")
          download.file(migration_url_2024, destfile = migration_file_2024, quiet = TRUE, mode = "wb")
        } else {
          message("Loading cached 2024 state-to-state migration file...")
        }
        
        # Table sheet: long format, col A = current state, col B = state 1yr ago, col C = estimate
        # "X" = same-state non-movers (filtered out), "N" = suppressed values (replaced with 0)
        flows_2024 <- read_excel(
          migration_file_2024,
          sheet     = "Table",
          skip      = 8,
          col_names = c("state_current", "state_1y", "population", "moe")
        ) |>
          select(state_current, state_1y, population) |>
          filter(!is.na(state_current), population != "X") |>
          mutate(
            population    = replace(population, population == "N", "0"),
            population    = as.integer(population),
            state_current = expand_state_name(state_current),
            state_1y      = expand_state_name(state_1y),
            year_current  = 2024L,
            year_1y       = 2023L
          ) |>
          filter(state_1y != "Foreign country") |>
          select(state_current, state_1y, population, year_current, year_1y)
        
        # Supplemental sheet: stationary population (did not move)
        # skip = 10 to bypass headers, empty row, and US aggregate row
        # col A = state name, col B = total population; sheet has 11 columns total
        stationary_2024 <- read_excel(
          migration_file_2024,
          sheet     = "Supplemental - Current Res",
          skip      = 10,
          col_names = c("state_current", "population", "moe_1", "est_2", "moe_2",
                        "est_3", "moe_3", "est_4", "moe_4", "est_5", "moe_5")
        ) |>
          select(state_current, population) |>
          filter(!is.na(state_current), !is.na(population)) |>
          mutate(
            population   = as.integer(population),
            state_1y     = state_current,
            year_current = 2024L,
            year_1y      = 2023L
          ) |>
          select(state_current, state_1y, population, year_current, year_1y)
        
        bind_rows(flows_2024, stationary_2024)
      }
      
      ##TASK 3: 2023 STATE TO STATE MIGRATION
      # -- Internal function: 2023 flows and stationary population --
      state_to_state_migration_2023 <- function() {
        
        migration_url_2023  <- "https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2023/state-to-state-migration/State_to_State_Migration_Table_2023_T13.xlsx"
        migration_file_2023 <- file.path(data_dir, "State_to_State_Migration_Table_2023_T13.xlsx")
        
        if (!file.exists(migration_file_2023)) {
          message("Downloading 2023 state-to-state migration file...")
          download.file(migration_url_2023, destfile = migration_file_2023, quiet = TRUE, mode = "wb")
        } else {
          message("Loading cached 2023 state-to-state migration file...")
        }
        
        # tool_input sheet: hidden sheet, long format, col A = state_1y, col B = state_current
        # skip = 1 to skip header row; "N/A" = same-state non-movers (filtered out)
        # "N" = suppressed values (replaced with 0)
        flows_2023 <- read_excel(
          migration_file_2023,
          sheet     = "tool_input",
          skip      = 1,
          col_names = c("state_1y", "state_current", "population", "moe",
                        "blank", "state", "in_est", "in_moe", "out_est", "out_moe",
                        "blank2", "state_names")
        ) |>
          select(state_1y, state_current, population) |>
          filter(!is.na(state_1y), population != "N/A") |>
          mutate(
            population    = replace(population, population == "N", "0"),
            population    = as.integer(population),
            # trim whitespace - DC has trailing space in source data
            state_current = str_trim(state_current),
            state_1y      = str_trim(state_1y),
            year_current  = 2023L,
            year_1y       = 2022L
          ) |>
          filter(state_1y != "Foreign country") |>
          select(state_current, state_1y, population, year_current, year_1y)
        
        # Stationary population from Table sheet
        # range A9:D74 reads only cols A-D starting after 8 header rows
        # col A = state name, col D = same residence estimate
        stationary_2023 <- read_excel(
          migration_file_2023,
          sheet     = "Table",
          range     = "A9:D74",
          col_names = c("state_current", "pop_est", "pop_moe", "population")
        ) |>
          select(state_current, population) |>
          filter(!is.na(state_current), !is.na(population),
                 !state_current %in% c("United States2", "Current residence in")) |>
          mutate(
            population   = as.integer(population),
            state_1y     = state_current,
            year_current = 2023L,
            year_1y      = 2022L
          ) |>
          select(state_current, state_1y, population, year_current, year_1y)
        
        bind_rows(flows_2023, stationary_2023)
      }
      
      # Combine both years into one tidy data frame
      bind_rows(state_to_state_migration_2024(), state_to_state_migration_2023())
    },
    # --- Metro-to-Metro Migration: ACS-5 2016-2020 metro-level flows ---
    metro_to_metro_migration = {
      
      # API URL for ACS-5 2016-2020 metro-to-metro migration flows
      # Requests MOVEDIN, MOVEDOUT, FULL1_NAME, FULL2_NAME, GEOID1, GEOID2
      # for all metropolitan/micropolitan statistical areas
      metro_migration_url  <- "https://api.census.gov/data/2020/acs/flows?get=MOVEDIN,MOVEDOUT,FULL1_NAME,FULL2_NAME,GEOID1,GEOID2&for=metropolitan%20statistical%20area/micropolitan%20statistical%20area:*"
      metro_migration_file <- file.path(data_dir, "metro_to_metro_migration.json")
      
      if (!file.exists(metro_migration_file)) {
        message("Downloading metro-to-metro migration data from Census API...")
        download.file(metro_migration_url, destfile = metro_migration_file, quiet = TRUE, mode = "wb")
      } else {
        message("Loading cached metro-to-metro migration data...")
      }
      
      # read_json() interprets first row as data instead of column names
      # move_row_to_colnames() fixes this by promoting first row to column headers
      move_row_to_colnames <- function(X, row = 1) {
        X <- as_tibble(X)
        Xrow <- X[row, ]
        X <- X[-row, ]
        colnames(X) <- Xrow
        readr::type_convert(X)
      }
      
      # Extracts the primary state abbreviation from a metro area name
      # For single-state metros (e.g. "Los Angeles-Long Beach, CA Metro Area") -> "CA"
      # For multi-state metros (e.g. "New York-Newark-Jersey City, NY-NJ-PA Metro Area") -> "NY"
      pull_state_from_metro <- function(metro_name) {
        str_extract(metro_name, ".*, (\\S{2})[-[:alpha:]]* Metro Area", group = 1)
      }
      
      read_json(metro_migration_file, simplifyVector = TRUE) |>
        as.data.frame() |>
        move_row_to_colnames() |>
        select(
          geoid1     = GEOID1,
          geoid2     = GEOID2,
          full1_name = FULL1_NAME,
          full2_name = FULL2_NAME,
          moved_in   = MOVEDIN,
          moved_out  = MOVEDOUT
        ) |>
        filter(!full2_name %in% c("Africa", "Asia", "Caribbean", "Central America", 
                           "Europe", "Northern America", "Oceania and At Sea", 
                           "South America", "U.S. Island Areas")) |>
        mutate(
          state1 = pull_state_from_metro(full1_name),
          state2 = pull_state_from_metro(full2_name)
        )
    },
    
    # --- International Migration: people moving into each state from abroad ---
    # Extracted from "Foreign country" rows in 2024 and 2023 migration files
    international_migration = {
      
      # -- 2024 international migration --
      migration_url_2024  <- "https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2024/state-to-state-migration/State_to_State_Migration_Table_2024_T13.xlsx"
      migration_file_2024 <- file.path(data_dir, "State_to_State_Migration_Table_2024_T13.xlsx")
      
      if (!file.exists(migration_file_2024)) {
        message("Downloading 2024 state-to-state migration file...")
        download.file(migration_url_2024, destfile = migration_file_2024, quiet = TRUE, mode = "wb")
      } else {
        message("Loading cached 2024 state-to-state migration file...")
      }
      
      # Extract only rows where state_1y == "Foreign country"
      intl_2024 <- read_excel(
        migration_file_2024,
        sheet     = "Table",
        skip      = 8,
        col_names = c("state_current", "state_1y", "population", "moe")
      ) |>
        select(state_current, state_1y, population) |>
        filter(!is.na(state_current), population != "X") |>
        mutate(
          population    = replace(population, population == "N", "0"),
          population    = as.integer(population),
          state_current = expand_state_name(state_current),
          year          = 2024L
        ) |>
        filter(state_1y == "Foreign country") |>
        select(state = state_current, international_migration = population, year) |>
        filter(!state %in% c("District of Columbia", "Puerto Rico"))
      
      # -- 2023 international migration --
      migration_url_2023  <- "https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2023/state-to-state-migration/State_to_State_Migration_Table_2023_T13.xlsx"
      migration_file_2023 <- file.path(data_dir, "State_to_State_Migration_Table_2023_T13.xlsx")
      
      if (!file.exists(migration_file_2023)) {
        message("Downloading 2023 state-to-state migration file...")
        download.file(migration_url_2023, destfile = migration_file_2023, quiet = TRUE, mode = "wb")
      } else {
        message("Loading cached 2023 state-to-state migration file...")
      }
      
      # Extract only rows where state_1y == "Foreign country" from tool_input sheet
      intl_2023 <- read_excel(
        migration_file_2023,
        sheet     = "tool_input",
        skip      = 1,
        col_names = c("state_1y", "state_current", "population", "moe",
                      "blank", "state", "in_est", "in_moe", "out_est", "out_moe",
                      "blank2", "state_names")
      ) |>
        select(state_1y, state_current, population) |>
        filter(!is.na(state_1y), population != "N/A") |>
        mutate(
          population    = replace(population, population == "N", "0"),
          population    = as.integer(population),
          state_current = str_trim(state_current),
          state_1y      = str_trim(state_1y),
          year          = 2023L
        ) |>
        filter(state_1y == "Foreign country") |>
        select(state = state_current, international_migration = population, year) |>
        filter(!state %in% c("District of Columbia", "Puerto Rico"))
      
      # Combine both years
      bind_rows(intl_2024, intl_2023)
    }
  )
}

STATE_POP                <- acquire_acs_data("population")
STATE_TO_STATE_MIGRATION <- acquire_acs_data("state_to_state_migration")
METRO_TO_METRO_MIGRATION <- acquire_acs_data("metro_to_metro_migration")
INTERNATIONAL_MIGRATION  <- acquire_acs_data("international_migration")

```

***Note:** My data frame will have more than 5000 rows because I've taken Puerto Rico and U.S. Island Areas into consideration and not just the 50 mainland(-ish) states. In-migration from Puerto Rico and islands may represent significant voting blocks that can be useful for analysis later on.*


```{r initial-data-check, include=FALSE}
#Checking if Data is correct for Task #1
dim(STATE_POP)
glimpse(STATE_POP)
head(STATE_POP)

#should be 9 years
distinct(STATE_POP, year)

#should be 52 states
STATE_POP |> count(year)

#should be california
STATE_POP |> group_by(year) |> slice_max(population)

#should be wyoming
STATE_POP |> group_by(year) |> slice_min(population)

#should increase
STATE_POP |> group_by(year) |> summarize(total_population = sum(population))
STATE_POP |> filter(state == "California", year == 2023)
STATE_POP |> summarize(across(everything(), ~sum(is.na(.))))
STATE_POP |> filter(population < 100000 | population > 50000000)
```

## Exploratory Analysis

### EDA: General

**1. Which states have had the highest net population growth rates over the past decade?**

```{r highest-net-pop-growth}

top_net_pop_growth_state <- STATE_POP |>
  filter(year %in% c(2015, 2024)) |>
  pivot_wider(
    names_from   = year,
    values_from  = population,
    names_prefix = "pop_"
  ) |>
  mutate(
    net_growth      = pop_2024 - pop_2015,
    net_growth_rate = (pop_2024 - pop_2015) / pop_2015 * 100
  ) |>
  arrange(desc(net_growth_rate)) |>
  select(state, pop_2015, pop_2024, net_growth, net_growth_rate) |>
  slice_max(net_growth_rate, n = 4)

growth_labels <- top_net_pop_growth_state |>
  select(state, net_growth_rate)

top_net_growth_facet <- STATE_POP |>
  filter(state %in% top_net_pop_growth_state$state) |>
  ggplot(aes(x = year, y = population, color = state)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  geom_text(
    data        = growth_labels,
    aes(
      x     = 2015,
      y     = Inf,
      label = paste0(round(net_growth_rate, 1), "%")
    ),
    hjust       = -0.1,
    vjust       = 1.5,
    size        = 6,
    color       = "darkgreen",
    inherit.aes = FALSE
  ) +  
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = c(2015, 2018, 2021, 2024)) +
  facet_wrap(~state, scales = "free_y") +
  labs(
    title    = "Population Change Over Time for Fastest Growing States",
    subtitle = "ACS-1 annual estimates, 2015-2024 (excluding 2020)",
    x        = NULL,
    y        = "Total Population",
    color    = NULL,
    caption  = "Source: ACS-1, US Census Bureau"
  ) +
  theme_minimal() +
  theme(
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 6),
    panel.spacing    = unit(1.5, "lines")
  )

top_net_growth_facet


#table
# top_net_growth_table <- top_net_pop_growth_state |>
#   gt() |>
#   tab_header(
#     title    = md("**Top 10 States by Net Population Growth Rate (2015-2024)**"),
#     subtitle = "Based on ACS-1 annual estimates"
#   ) |>
#   cols_label(
#     state           = md("**State**"),
#     pop_2015        = md("**Population<br>(2015)**"),
#     pop_2024        = md("**Population<br>(2024)**"),
#     net_growth      = md("**Net Growth**"),
#     net_growth_rate = md("**Growth Rate (%)**")
#   ) |>
#   cols_align(
#     align   = "center",
#     columns = c(pop_2015, pop_2024, net_growth, net_growth_rate)
#   ) |>
#   fmt_number(
#     columns  = c(pop_2015, pop_2024, net_growth),
#     decimals = 0
#   ) |>
#   fmt_number(
#     columns  = net_growth_rate,
#     decimals = 2
#   ) |>
#   data_color(
#     columns = net_growth_rate,
#     palette = c("#90CAF9", "#1565C0")
#   ) |>
#   tab_source_note(
#     source_note = "Source: ACS-1 Baseline State Population Data, US Census Bureau"
#   )

# # bar chart
# top_net_growth_bar <- top_net_pop_growth_state |>
#   mutate(state = fct_reorder(state, net_growth_rate)) |>
#   ggplot(aes(x = net_growth_rate, y = state, fill = net_growth_rate)) +
#   geom_col() +
#   geom_vline(xintercept = 0, linewidth = 0.8, color = "black") +
#   geom_text(
#     aes(label = paste0(round(net_growth_rate, 1), "%")),
#     hjust    = -0.2,
#     size     = 3
#   ) +
#   scale_fill_gradient(low = "#90CAF9", high = "#1565C0") +
#   labs(
#     title    = "Top 15 States by Net Population Growth Rate (2015-2024)",
#     subtitle = "Percentage change in total population",
#     x        = "Net Growth Rate (%)",
#     y        = NULL,
#     fill     = "Growth Rate (%)",
#     caption  = "Source: ACS-1, US Census Bureau"
#   ) +
#   theme_minimal() +
#   theme(
#     legend.position    = "bottom",
#     panel.grid.major.y = element_blank()
#   )

```

Idaho (20.9%), Utah (16.9%), Florida (15.3%) and Texas (13.9%) have the highest net population growth rates over the past decade.

**2. If you meet someone who moved to New York State in the last year, what are the most likely states they moved from? If one of your friends announces they are moving out of the state, where are they most likely to move?**

```{r most-likely-states-moved-to-from-NY}

# most likely states people moved FROM into New York
moved_to_ny <- STATE_TO_STATE_MIGRATION |>
  filter(
    state_current == "New York",
    state_1y != "New York",
    year_current == 2024
  ) |>
  arrange(desc(population)) |>
  select(state_1y, population) |>
# %age out of total migration to NY
  mutate(pct = population / sum(population) * 100) |>
  slice_max(population, n = 5)

# most likely states people moved TO from New York
moved_from_ny <- STATE_TO_STATE_MIGRATION |>
  filter(
    state_1y == "New York",
    state_current != "New York",
    year_current == 2024
  ) |>
  arrange(desc(population)) |>
  select(state_current, population) |>
  mutate(pct = population / sum(population) * 100) |>
  slice_max(population, n = 5)

# top 5 states people moved from into New York
plot_to_ny <- moved_to_ny |>
  ggplot(aes(x = population, y = fct_reorder(state_1y, population), fill = population)) +
  geom_col() +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    hjust = -0.2,
    size  = 3
  ) +
  scale_fill_gradient(low = "#90CAF9", high = "#1565C0") +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title   = "Top 5 States People Moved From Into New York (2024)",
    x       = "Number of People",
    y       = NULL,
    caption = "Source: ACS-1, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position    = "none",
    panel.grid.major.y = element_blank()
  )

# top 5 states people moved to from New York
plot_from_ny <- moved_from_ny |>
  ggplot(aes(x = population, y = fct_reorder(state_current, population), fill = population)) +
  geom_col() +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    hjust = -0.2,
    size  = 3
  ) +
  scale_fill_gradient(low = "#FFCDD2", high = "#C62828") +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title   = "Top 5 States People Moved To From New York (2024)",
    x       = "Number of People",
    y       = NULL,
    caption = "Source: ACS-1, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position    = "none",
    panel.grid.major.y = element_blank()
  )

# stack vertically using patchwork
plot_to_ny / plot_from_ny

```

The top 5 states people moved from into New York in 2024 were New Jersey, California, Florida, Pennsylvania and Massachusetts.
The top 5 states New Yorkers moved to in 2024 were New Jersey, Florida, Pennsylvania, Texas and Connecticut.

**3. If you meet someone who moved to New York City in the last year, what are the most likely metro areas they moved from? If one of your friends announces they are moving out of the city, where are they most likely to move?**

```{r most-likely-states-moved-to-from-nyc}

nyc_metro <- "New York-Newark-Jersey City, NY-NJ-PA Metro Area"

# top 5 metro areas people moved FROM into NYC
moved_to_nyc <- METRO_TO_METRO_MIGRATION |>
  filter(
    full1_name == nyc_metro,
    full2_name != nyc_metro,
# excluded outside metro area since these are places not considered Metro Areas    
    full2_name != "Outside Metro Area within U.S. or Puerto Rico"
  ) |>
  mutate(pct = moved_in / sum(moved_in) * 100) |>
  arrange(desc(moved_in)) |>
  slice_max(moved_in, n = 5) |>
  mutate(full2_name = str_replace(full2_name, "(.*),\\s*(\\S+)\\s*Metro Area", "\\1, \\2"))

# top 5 metro areas people moved TO from NYC
moved_from_nyc <- METRO_TO_METRO_MIGRATION |>
  filter(
    full1_name == nyc_metro,
    full2_name != nyc_metro,
# excluded outside metro area since these are places not considered Metro Areas
    full2_name != "Outside Metro Area within U.S. or Puerto Rico"
  ) |>
  mutate(pct = moved_out / sum(moved_out) * 100) |>
  arrange(desc(moved_out)) |>
  slice_max(moved_out, n = 5) |>
  mutate(full2_name = str_replace(full2_name, "(.*),\\s*(\\S+)\\s*Metro Area", "\\1, \\2"))

plot_to_nyc <- moved_to_nyc |>
  ggplot(aes(x = moved_in, y = fct_reorder(full2_name, moved_in), fill = moved_in)) +
  geom_col() +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    hjust = -0.2,
    size  = 3
  ) +
  scale_fill_gradient(low = "#90CAF9", high = "#1565C0") +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title   = "Top 5 Metro Areas People Moved From Into NYC",
    x       = "Number of People",
    y       = NULL,
    caption = "Source: ACS-5 2016-2020 Metropolitan Flows, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position    = "none",
    panel.grid.major.y = element_blank()
  )

plot_from_nyc <- moved_from_nyc |>
  ggplot(aes(x = moved_out, y = fct_reorder(full2_name, moved_out), fill = moved_out)) +
  geom_col() +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    hjust = -0.2,
    size  = 3
  ) +
  scale_fill_gradient(low = "#FFCDD2", high = "#C62828") +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title   = "Top 5 Metro Areas People Moved To From NYC",
    x       = "Number of People",
    y       = NULL,
    caption = "Source: ACS-5 2016-2020 Metropolitan Flows, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position    = "none",
    panel.grid.major.y = element_blank()
  )

plot_to_nyc / plot_from_nyc


```

The Philadelphia, PA metro area is where someone most likely would have moved to NYC from and where New Yorkers are most likely to move to. 


**4. Which states have had the highest amounts of in-migration and out-migration? Furthermore, which states have had the highest amount of net migration?**

```{r in-out-net-migration-state}

# total in-migration per state (people who moved INTO that state)
in_migration <- STATE_TO_STATE_MIGRATION |>
  filter(
    state_current != state_1y,
    year_current  == 2024
  ) |>
  group_by(state_current) |>
  summarize(total_in = sum(population))



# total out-migration per state (people who moved OUT of that state)
out_migration <- STATE_TO_STATE_MIGRATION |>
  filter(
    state_current != state_1y,
    year_current  == 2024
  ) |>
  group_by(state_1y) |>
  summarize(total_out = sum(population))

# combine and compute net migration
net_migration <- in_migration |>
  inner_join(out_migration, join_by(state_current == state_1y)) |>
  mutate(net = total_in - total_out) |>
  rename(state = state_current)

# plot 1: top 10 states by in-migration
plot_in_migration <- net_migration |>
  slice_max(total_in, n = 10) |>
  mutate(state = fct_reorder(state, total_in)) |>
  ggplot(aes(x = total_in, y = state, fill = total_in)) +
  geom_col() +
  geom_text(
    aes(label = scales::comma(total_in)),
    hjust = 1.2,
    size  = 3,
    color = "white"
  ) +
  scale_fill_gradient(low = "#90CAF9", high = "#1565C0") +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title   = "States by Total In-Migration",
    subtitle = "Top 10 (2024)",
    x       = "Number of People",
    y       = NULL,
    caption = "Source: ACS-1, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position    = "none",
    panel.grid.major.y = element_blank()
  )

# plot 2: top 10 states by out-migration
plot_out_migration <- net_migration |>
  slice_max(total_out, n = 10) |>
  mutate(state = fct_reorder(state, total_out)) |>
  ggplot(aes(x = total_out, y = state, fill = total_out)) +
  geom_col() +
  geom_text(
    aes(label = scales::comma(total_in)),
    hjust = 1.2,
    size  = 3,
    color = "white"
  ) +
  scale_fill_gradient(low = "#E1BEE7", high = "#4A148C") +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title   = "States by Total Out-Migration",
    subtitle = "Top 10 (2024)",
    x       = "Number of People",
    y       = NULL,
    caption = "Source: ACS-1, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position    = "none",
    panel.grid.major.y = element_blank()
  )

net_migration_table <- net_migration |>
  slice_max(net, n = 10) |>
  arrange(desc(net)) |>
  select(state, total_in, total_out, net) |>
  gt() |>
  tab_header(
    title = md("**Top 10 States by Net Migration (2024)**")
  ) |>
  cols_label(
    state     = md("**State**"),
    total_in  = md("**Total In-Migration**"),
    total_out = md("**Total Out-Migration**"),
    net       = md("**Net Migration**")
  ) |>
  cols_align(
    align   = "center",
    columns = c(total_in, total_out, net)
  ) |>
  fmt_number(
    columns  = c(total_in, total_out, net),
    decimals = 0
  ) |>
  data_color(
    columns = net,
    palette = c("#C8E6C9", "#1B5E20")
  ) |>
  tab_source_note(
    source_note = "Source: ACS-1, US Census Bureau"
  ) |>
  tab_options(
    heading.align = "center"
  )

plot_in_migration | plot_out_migration

net_migration_table

```

Florida, Texas, and California had the highest amounts of in-migration.

The same states had the highest amounts of out-migration except in a differing order (California, Florida, Texas).

Texas, Florida and North Carolina had the highest amount of net migration.


**5. Which metro areas have had the highest amounts of total in-migration and out-migration? Furthermore, which metro areas have had the highest amount of net migration?**

```{r in-out-net-migration-metro}

# total in-migration per metro area
metro_in_migration <- METRO_TO_METRO_MIGRATION |>
  filter(full1_name != full2_name) |>
  group_by(full1_name) |>
  summarize(total_in = sum(moved_in, na.rm = TRUE))

# total out-migration per metro area
metro_out_migration <- METRO_TO_METRO_MIGRATION |>
  filter(full1_name != full2_name) |>
  group_by(full1_name) |>
  summarize(total_out = sum(moved_out, na.rm = TRUE))

# combine and compute net migration
metro_net_migration <- metro_in_migration |>
  inner_join(metro_out_migration, join_by(full1_name)) |>
  mutate(
    net       = total_in - total_out,
    full1_name = str_replace(full1_name, "(.*),\\s*(\\S+)\\s*Metro Area", "\\1, \\2")
  ) |>
  rename(metro = full1_name)

# plot 1: top 10 metro areas by in-migration
plot_metro_in <- metro_net_migration |>
  slice_max(total_in, n = 10) |>
  mutate(
    metro = paste0(str_extract(metro, "^[^-]+"), ", ", str_extract(metro, ",\\s*(\\w{2})", group = 1)),
    metro = fct_reorder(metro, total_in)
  ) |>
  ggplot(aes(x = total_in, y = metro, fill = total_in)) +
  geom_col() +
  geom_text(
    aes(label = scales::comma(total_in)),
    hjust = 1.2,
    size = 3,
    color = "white"
  ) +
  scale_fill_gradient(low = "#90CAF9", high = "#1565C0") +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Metro Areas by In-Migration",
    subtitle = "Top 10 (2016-2020)",
    x = "Number of People",
    y = NULL,
    caption = "Source: ACS-5, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank()
  )

# plot 2: top 10 metro areas by out-migration
plot_metro_out <- metro_net_migration |>
  slice_max(total_out, n = 10) |>
  mutate(
    metro = paste0(str_extract(metro, "^[^-]+"), ", ", str_extract(metro, ",\\s*(\\w{2})", group = 1)),
    metro = fct_reorder(metro, total_out)
  ) |>
  ggplot(aes(x = total_out, y = metro, fill = total_out)) +
  geom_col() +
  geom_text(
    aes(label = scales::comma(total_out)),
    hjust = 1.2,
    size  = 3,
    color = "white"
  ) +
  scale_fill_gradient(low = "#E1BEE7", high = "#4A148C") +
  scale_x_continuous(
    labels = scales::comma,
    breaks = seq(0, max(metro_net_migration$total_out), by = 200000),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Metro Areas by Out-Migration",
    subtitle = "Top 10 (2016-2020)",
    x = "Number of People",
    y = NULL,
    caption = "Source: ACS-5, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank()
  )


# gt table for top 10 metro areas by net migration
metro_net_table <- metro_net_migration |>
  slice_max(net, n = 10) |>
  arrange(desc(net)) |>
  select(metro, total_in, total_out, net) |>
  gt() |>
  tab_header(
    title = "Top 10 Metro Areas by Net Migration (2016-2020)"
  ) |>
  cols_label(
    metro     = "Metro Area",
    total_in  = "Total In-Migration",
    total_out = "Total Out-Migration",
    net       = "Net Migration"
  ) |>
  cols_align(
    align   = "center",
    columns = c(total_in, total_out, net)
  ) |>
  fmt_number(
    columns  = c(total_in, total_out, net),
    decimals = 0
  ) |>
  data_color(
    columns = net,
    palette = c("#C8E6C9", "#1B5E20")
  ) |>
  tab_source_note(
    source_note = "Source: ACS-5, US Census Bureau"
  ) |>
  tab_options(
    heading.align = "center"
  )

# display in and out migration side by side
plot_metro_in | plot_metro_out

metro_net_table

```

Los Angeles, CA, Dallas, TX, New York, NY metro areas had the highest amounts of total in-migration.

New York, NY, Los Angeles, CA, and Chicago, IL metro areas had the highest amounts of total out-migration.

Phoenix, AZ, Riverside, CA, and Austin, TX had the highest amount of net migration.


**6. Which states have the highest fraction of residents who lived there last year? (That is, which states have the lowest proportional total in-migration?)**

```{r state-pop-stability}

# total state populations
state_pop_2024 <- STATE_POP |>
  filter(year == 2024) |>
  select(state, total_pop = population)

# calculate in-migration fraction per state
state_stability <- STATE_TO_STATE_MIGRATION |>
  filter(year_current == 2024) |>
  group_by(state_current) |>
  summarize(in_migration = sum(population[state_current != state_1y])) |>
  rename(state = state_current) |>
  left_join(state_pop_2024, join_by(state)) |>
  mutate(
    stability_rate   = 1 - (in_migration / total_pop),
    in_migration_pct = in_migration / total_pop * 100
  ) |>
  filter(state != "Puerto Rico")

state_stability_table <- state_stability |>
  slice_max(stability_rate, n = 10) |>
  arrange(desc(stability_rate)) |>
  select(state, total_pop, in_migration, stability_rate, in_migration_pct) |>
  gt() |>
  tab_header(
    title = md("**Top States by Resident Stability Rate**"),
    subtitle = md("States with Lowest Proportional Total In-Migration (2024)")
  ) |>
  cols_label(
    state = "State",
    total_pop = "Total Population",
    in_migration = "Total In-Migration",
    stability_rate = "Stability Rate",
    in_migration_pct = "In-Migration (%)"
  ) |>
  cols_align(
    align = "center",
    columns = c(total_pop, in_migration, stability_rate, in_migration_pct)
  ) |>
  fmt_number(
    columns = c(total_pop, in_migration),
    decimals = 0
  ) |>
  fmt_percent(
    columns = stability_rate,
    decimals = 2
  ) |>
  fmt_number(
    columns = in_migration_pct,
    decimals = 2
  ) |>
  data_color(
    columns = stability_rate,
    palette = c("#C8E6C9", "#1B5E20")
  ) |>
  tab_source_note(
    source_note = "Source: ACS-1, US Census Bureau"
  ) |>
  tab_options(
    heading.align = "center"
  )

state_stability_table

```

California, Michigan and New York had the lowest proportional total in-migration.


**7. Which state has the highest fraction of its population growth attributable to net internal migration? That is, determine the total population growth of each state and see what fraction of it corresponds to net migration (as opposed to deaths, births, or international migration).**

```{r pop-growth-net-migration}

# compute population growth from 2023 to 2024 using STATE_POP
pop_growth <- STATE_POP |>
  filter(year %in% c(2023, 2024)) |>
  pivot_wider(
    names_from = year,
    values_from = population,
    names_prefix = "pop_"
  ) |>
  mutate(pop_growth = pop_2024 - pop_2023)

# combine with net_migration (from question #4) and compute fraction of growth attributable to net migration
migration_growth_share <- pop_growth |>
  left_join(net_migration, join_by(state)) |>
  mutate(
    migration_share = net / pop_growth * 100
  ) |>
  filter(state != "Puerto Rico") |>
  select(state, pop_2023, pop_2024, pop_growth, net, migration_share)

top_migration_growth_state <- migration_growth_share |>
  slice_max(migration_share, n = 1)

```

The state with the highest fraction of population growth attributable to net internal migration is `r top_migration_growth_state$state`, with `r round(top_migration_growth_state$migration_share, 1)`% of its growth coming from internal migration.

***Note:** Migration Share can exceed 100% if net migration is larger than total population growth (meaning natural growth or international migration was negative), or be negative if net migration was negative while the state still grew overall. It means some states are growing purely because of internal migration despite losing population to deaths or international emigration.*

*For example: *

* *Net internal migration: +100,000 (people moving in from other states)*

* *Natural increase: -20,000 (more deaths than births, common in older states)*

* *Net international migration: -10,000*

*Then total population growth would be +100,000 - 20,000 - 10,000 = +70,000, but net migration alone was +100,000. So migration share = 100,000 / 70,000* * *100 = 142.9%, which exceeds 100% because the other components partially offset the migration gains.*


### EDA: New York

**8. Which state had the highest amount of migration into your state? Of the people who moved out of your state, what was the most popular destination state?**

```{r nys-state-migration-flow}

# state with highest amount of migration into New York
top_inflow_to_ny <- STATE_TO_STATE_MIGRATION |>
  filter(
    state_current == "New York",
    state_1y != "New York",
    year_current == 2024
  ) |>
  slice_max(population, n = 1)

# most popular destination state for people leaving New York
top_outflow_from_ny <- STATE_TO_STATE_MIGRATION |>
  filter(
    state_1y      == "New York",
    state_current != "New York",
    year_current  == 2024
  ) |>
  slice_max(population, n = 1)

```
The state with the highest amount of migration into New York was `r top_inflow_to_ny$state_1y`, with `r scales::comma(top_inflow_to_ny$population)` people moving in.

Of the people who moved out of New York, the most popular destination was `r top_outflow_from_ny$state_current`, with `r scales::comma(top_outflow_from_ny$population)` people moving there.


**9. What is the largest metropolitan area in your state? (You can answer this just using background knowledge; you do not have to use Census data here.) Which metro area, not located in your state, has the highest amount of migration into your state’s largest metro? To which metro area outside of your state do the most residents of your largest metro area move?**

```{r nys-metro-migration-flow}

nyc_metro <- "New York-Newark-Jersey City, NY-NJ-PA Metro Area"

# metro area outside New York with highest migration into NYC metro
top_inflow_to_nyc <- METRO_TO_METRO_MIGRATION |>
  filter(
    full1_name == nyc_metro,
    full2_name != nyc_metro,
    state2 != "NY",
    full2_name != "Outside Metro Area within U.S. or Puerto Rico"
  ) |>
  slice_max(moved_in, n = 1)

# most popular destination metro outside New York for NYC metro residents
top_outflow_from_nyc <- METRO_TO_METRO_MIGRATION |>
  filter(
    full1_name == nyc_metro,
    full2_name != nyc_metro,
    state2 != "NY",
    full2_name != "Outside Metro Area within U.S. or Puerto Rico"
  ) |>
  slice_max(moved_out, n = 1)

```

The largest metro area is NYC, which is contained in the New York-Newark-Jersey City, NY-NJ-PA Metro Area in our Census data.

The metro area outside of New York with the highest migration into the NYC metro was `r top_inflow_to_nyc$full2_name`, with `r scales::comma(top_inflow_to_nyc$moved_in)` people moving in.

The most popular destination metro outside of New York for NYC metro residents was `r top_outflow_from_nyc$full2_name`, with `r scales::comma(top_outflow_from_nyc$moved_out)` people moving there.


**10. Are there any metro areas that have a particularly high connection to your state’s largest metro?**

```{r nyc-metro-area-connections}

nyc_connection <- METRO_TO_METRO_MIGRATION |>
  filter(
    full2_name == nyc_metro,
    full1_name != nyc_metro,
    full1_name != "Outside Metro Area within U.S. or Puerto Rico"
  ) |>
  left_join(
    # get total in-migration for each metro area
    METRO_TO_METRO_MIGRATION |>
      filter(full1_name != full2_name) |>
      group_by(full1_name) |>
      summarize(total_in = sum(moved_in, na.rm = TRUE)),
    join_by(full1_name)
  ) |>
  mutate(
    nyc_share = moved_in / total_in * 100
  ) |>
  filter(nyc_share >= 15) |>
  arrange(desc(nyc_share)) |>
  select(full1_name, moved_in, total_in, nyc_share)

```

Outside of NYC and New York State, NYCers account for over half (`r round(nyc_connection |> slice_max(nyc_share, n=1) |> pull(nyc_share), 1)`%) of migration into the `r nyc_connection |> slice_max(nyc_share, n=1) |> pull(full1_name)`.


## Final Insights and Deliverable

### Population Projections

```{r population-projections-2030}
## TASK 7 - POPULATION PROJECTION
#
#
## National populations and migratoin excluding DC and Puerto Rico for forecast model
national_pop <- STATE_POP |>
  filter(!state %in% c("District of Columbia", "Puerto Rico"))

national_migration <- STATE_TO_STATE_MIGRATION |>
  filter(
    !state_current %in% c("District of Columbia", "Puerto Rico"),
    !state_1y      %in% c("District of Columbia", "Puerto Rico")
  )

# Step 1: Compute M(t) - total international in-migration for each year
# M(2023): people living abroad in 2023 who moved to a US state by 2024
M_2023 <- INTERNATIONAL_MIGRATION |>
  filter(year == 2024) |>
  summarize(total = sum(international_migration, na.rm = TRUE)) |>
  pull(total)

M_2022 <- INTERNATIONAL_MIGRATION |>
  filter(year == 2023) |>
  summarize(total = sum(international_migration, na.rm = TRUE)) |>
  pull(total)

# Step 2: Compute g for each year
# g(2023) = (P(2024) - P(2023) - M(2023)) / P(2023)
P_2023 <- national_pop |> filter(year == 2023) |> summarize(total = sum(population)) |> pull(total)
P_2024 <- national_pop |> filter(year == 2024) |> summarize(total = sum(population)) |> pull(total)

g_2023 <- (P_2024 - P_2023 - M_2023) / P_2023

# g(2022) = (P(2023) - P(2022) - M(2022)) / P(2022)
P_2022 <- national_pop |> filter(year == 2022) |> summarize(total = sum(population)) |> pull(total)

g_2022 <- (P_2023 - P_2022 - M_2022) / P_2022

# Step 3: Compute migration rates m_j->i for each state pair
# m_j->i = M_j->i / sqrt(P_i * P_j)
# j = origin state, i = destination state

# State populations for 2022, 2023 and 2024
state_pop_2022 <- national_pop |> filter(year == 2022) |> select(state, population)
state_pop_2023 <- national_pop |> filter(year == 2023) |> select(state, population)
state_pop_2024 <- national_pop |> filter(year == 2024) |> select(state, population)

# m_j->i for 2023 (using 2023 and 2024 populations)
m_2023 <- national_migration |>
  filter(year_current == 2024, state_current != state_1y) |>
  select(state_j = state_1y, state_i = state_current, M_ji = population) |>
  left_join(state_pop_2024, join_by(state_i == state)) |>
  rename(pop_i = population) |>
  left_join(state_pop_2023, join_by(state_j == state)) |>
  rename(pop_j = population) |>
  mutate(m_ji = M_ji / sqrt(pop_i * pop_j))

# m_j->i for 2022 (using 2022 and 2023 populations)
m_2022 <- national_migration |>
  filter(year_current == 2023, state_current != state_1y) |>
  select(state_j = state_1y, state_i = state_current, M_ji = population) |>
  left_join(state_pop_2023, join_by(state_i == state)) |>
  rename(pop_i = population) |>
  left_join(state_pop_2022, join_by(state_j == state)) |>
  rename(pop_j = population) |>
  mutate(m_ji = M_ji / sqrt(pop_i * pop_j))

# Step 5: Average g and m_ji over both years
g_bar <- (g_2023 + g_2022) / 2

m_bar <- m_2023 |>
# full join retains all state pairs even if a pair only appears in one year  
  full_join(m_2022, join_by(state_i, state_j)) |>
#replace NA values with 0 for pairs that appear in one YoY migration and not the other, then take average. Otherwise, m_ji_bar could be NA for some state pairs
  mutate(
    m_ji.x   = replace_na(m_ji.x, 0),
    m_ji.y   = replace_na(m_ji.y, 0),
    m_ji_bar = (m_ji.x + m_ji.y) / 2
  ) |>
  select(state_i, state_j, m_ji_bar)

# Step 6: Project populations forward to 2030
# Start with 2024 populations as the base year
current_pop <- state_pop_2024 |>
  rename(pop = population)

# Project forward one year using the forecast model
# P_i(t+1) = P_i(t) * (1 + g_bar) + sum over j of m_ji_bar * sqrt(P_i(t) * P_j(t))
project_one_year <- function(pop_df, m_bar, g_bar) {
  
  # join current populations onto m_bar for both state_i and state_j
  m_with_pop <- m_bar |>
    left_join(pop_df, join_by(state_i == state)) |>
    rename(pop_i = pop) |>
    left_join(pop_df, join_by(state_j == state)) |>
    rename(pop_j = pop) |>
    mutate(migration_contribution = m_ji_bar * sqrt(pop_i * pop_j))
  
  # sum migration contributions for each destination state
  migration_in <- m_with_pop |>
    group_by(state_i) |>
    summarize(total_migration_in = sum(migration_contribution, na.rm = TRUE))
  
  # apply natural growth and add migration
  pop_df |>
    left_join(migration_in, join_by(state == state_i)) |>
    mutate(
      pop = pop * (1 + g_bar) + replace_na(total_migration_in, 0)
    ) |>
    select(state, pop)
}

# iterate year by year from 2025 to 2030 storing all years
pop_projections <- accumulate(
  2025:2030,
  function(pop_df, yr) project_one_year(pop_df, m_bar, g_bar),
  .init = current_pop
) |>
  set_names(c(2024, 2025:2030)) |>
  map2(c(2024, 2025:2030), ~ mutate(.x, year = .y)) |>
  bind_rows()

nys_population_historic_and_projected_plot <- bind_rows(
  STATE_POP |>
    filter(state == "New York") |>
    mutate(type = "Actual"),
  pop_projections |>
    filter(state == "New York", year > 2024) |>
    rename(population = pop) |>
    mutate(type = "Projected")
) |>
  ggplot(aes(x = year, y = population, color = type, linetype = type)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = c(2015:2019, 2021:2030)) +
  scale_color_manual(values = c("Actual" = "#1565C0", "Projected" = "#C62828")) +
  scale_linetype_manual(values = c("Actual" = "solid", "Projected" = "dashed")) +
  labs(
    title    = "New York State Population: Historical and Projected (2015-2030)",
    subtitle = "Based on ACS-1 estimates and Huntington-Hill forecast model",
    x        = NULL,
    y        = "Population",
    color    = NULL,
    linetype = NULL,
    caption  = "Source: ACS-1, US Census Bureau. Note: 2020 excluded due to COVID-19 data gap."
  ) +
  theme_bw() +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 6)
  )

nys_population_historic_and_projected_plot

```


### Reapportionment

```{r nys-reapportionment}
##TASK 8 - REAPPORTIONMENT 
#
# Apply Huntington-Hill method to 2030 population projections
MAX_DISTRICTS <- 100 # max gives safe buffer zone for algorithm to do its thing
N_DISTRICTS   <- 435

ny_districts_2030 <- pop_projections |>
  filter(year == 2030) |>
  expand_grid(cd = seq(1, MAX_DISTRICTS)) |>
  mutate(
    hh_den    = sqrt(cd * (cd + 1)),
    hh_weight = pop / hh_den
  ) |>
  slice_max(hh_weight, n = N_DISTRICTS) |>
  group_by(state) |>
  summarize(
    population   = first(pop),
    n_districts  = n()
  ) |>
  filter(state == "New York") |>
  pull(n_districts)


all_states_apportionment_table <- pop_projections |>
  filter(year == 2030) |>
  expand_grid(cd = seq(1, MAX_DISTRICTS)) |>
  mutate(
    hh_den    = sqrt(cd * (cd + 1)),
    hh_weight = pop / hh_den
  ) |>
  slice_max(hh_weight, n = N_DISTRICTS) |>
  group_by(state) |>
  summarize(
    population  = first(pop),
    n_districts = n()
  ) |>
  mutate(pop_per_district = population / n_districts) |>
  arrange(desc(n_districts), desc(population), state) |>
  gt(rowname_col = "state") |>
  cols_label(
    population      = "Projected Population (2030)",
    n_districts     = "Number of Congressional Districts",
    pop_per_district = "Approximate Population per District"
  ) |>
  fmt_integer(c(population, pop_per_district)) |>
  data_color(
    columns = pop_per_district,
    palette = "PuOr"
  ) |>
  grand_summary_rows(
    columns = c(population, n_districts),
    fns     = list("Total" ~ sum(.)),
    fmt     = ~ fmt_integer(.)
  ) |>
  tab_header("Projected Congressional Apportionments for 2030") |>
  tab_source_note(
    md("Apportionments computed using the Huntington-Hill Method.
    Population projections based on ACS-1 migration flows (2023-2024).")
  )

```

New York is projected to have `r ny_districts_2030` congressional districts as of the 2030 redistricting cycle. 

### I'm Walking Here! But Where To? 

##### *New Yorkers are Walking Out on New York but are Enough Moving In*

```{r nys-advertising-strategy}

# Find states near the low end of population per district (vulnerable to losing a seat)
vulnerable_states <- pop_projections |>
  filter(year == 2030) |>
  expand_grid(cd = seq(1, MAX_DISTRICTS)) |>
  mutate(
    hh_den    = sqrt(cd * (cd + 1)),
    hh_weight = pop / hh_den
  ) |>
  slice_max(hh_weight, n = N_DISTRICTS) |>
  group_by(state) |>
  summarize(
    population  = first(pop),
    n_districts = n()
  ) |>
  mutate(pop_per_district = population / n_districts) |>
  arrange(pop_per_district)

# How many people would NY need to gain to get an extra seat?
# Find the priority value of the next seat NY would get vs
# the lowest priority value currently allocated
apportionment_2030 <- pop_projections |>
  filter(year == 2030) |>
  expand_grid(cd = seq(1, MAX_DISTRICTS)) |>
  mutate(
    hh_den    = sqrt(cd * (cd + 1)),
    hh_weight = pop / hh_den
  )


# Priority value of NY's next potential seat
ny_next_seat <- apportionment_2030 |>
  filter(state == "New York") |>
  slice_max(hh_weight, n = ny_districts_2030 + 1) |>
  slice_min(hh_weight, n = 1) |>
  pull(hh_weight)

#Lower priority value states currently allocated
lower_allocation <- apportionment_2030 |> 
  slice_max(hh_weight, n = N_DISTRICTS) |>
  slice_min(hh_weight, n = 10)


# Lowest priority value currently allocated (435th seat)
lowest_allocated <- apportionment_2030 |>
  slice_max(hh_weight, n = N_DISTRICTS) |>
  slice_min(hh_weight, n = 1) |>
  pull(hh_weight)

# The state holding the 435th seat
seat_435_state <- apportionment_2030 |>
  slice_max(hh_weight, n = N_DISTRICTS) |>
  slice_min(hh_weight, n = 1) |>
  pull(state)


# Top sources of in-migration to New York state
top_sources_ny <- STATE_TO_STATE_MIGRATION |>
  filter(
    state_current == "New York",
    state_1y      != "New York",
    year_current  == 2024
  ) |>
  arrange(desc(population)) |>
  slice_max(population, n = 10)

# Top destinations of out-migration from New York state
top_destinations_ny <- STATE_TO_STATE_MIGRATION |>
  filter(
    state_1y      == "New York",
    state_current != "New York",
    year_current  == 2024
  ) |>
  arrange(desc(population)) |>
  slice_max(population, n = 5)

#Identify NY metros
ny_metros <- METRO_TO_METRO_MIGRATION |>
  distinct(full1_name) |>
  filter(str_detect(full1_name, "NY")) |>
  pull(full1_name)

#NY's top 3 metro areas
ny_top3_metros <- c(
  "New York-Newark-Jersey City, NY-NJ-PA Metro Area",
  "Buffalo-Cheektowaga, NY Metro Area",
  "Rochester, NY Metro Area"
)

# Top sources of in-migration into NY's 3 largest metro areas (excluding other NY metros)
top_sources_ny_metro <- METRO_TO_METRO_MIGRATION |>
  filter(
    full1_name %in% ny_top3_metros,
    !full2_name %in% ny_metros,
    full2_name != "Outside Metro Area within U.S. or Puerto Rico"
  ) |>
  group_by(full1_name) |>
  mutate(pct = moved_in / sum(moved_in) * 100) |>
  slice_max(moved_in, n = 5) |>
  mutate(full2_name = str_replace(full2_name, "(.*),\\s*(\\S+)\\s*Metro Area", "\\1, \\2")) |>
  ungroup()

# Top destinations of out-migration from NY's 3 largest metro areas (excluding other NY metros)
top_destinations_ny_metro <- METRO_TO_METRO_MIGRATION |>
  filter(
    full1_name %in% ny_top3_metros,
    !full2_name %in% ny_metros,
    full2_name != "Outside Metro Area within U.S. or Puerto Rico"
  ) |>
  group_by(full1_name) |>
  mutate(pct = moved_out / sum(moved_out) * 100) |>
  slice_max(moved_out, n = 10) |>
  mutate(full2_name = str_replace(full2_name, "(.*),\\s*(\\S+)\\s*Metro Area", "\\1, \\2")) |>
  ungroup()

# print(top_sources_ny)
# print(top_destinations_ny)
# print(top_sources_ny_metro)
# print(top_destinations_ny_metro)


#Net Migration for the top destination states NYers move to
  top_destination_states_net <- STATE_TO_STATE_MIGRATION |>
    filter(
      year_current == 2024,
      state_current != state_1y
    ) |>
    filter(
      (state_1y == "New York" & state_current %in% top_destinations_ny$state_current) |
      (state_current == "New York" & state_1y %in% top_destinations_ny$state_current)
    ) |>
    group_by(state = if_else(state_1y == "New York", state_current, state_1y)) |>
    summarize(
      inflow  = sum(population[state_current == "New York"]),
      outflow = sum(population[state_1y == "New York"])
    ) |>
    mutate(
      net          = inflow - outflow,
      state        = fct_reorder(state, inflow),
      outflow_neg  = -outflow
    ) |>
    pivot_longer(
      cols      = c(inflow, outflow_neg),
      names_to  = "type",
      values_to = "population"
    )

#plot for net migration for top destination states NYers move to
  top_destination_states_net_plot <- top_destination_states_net |>
    ggplot(aes(x = population, y = state, fill = type)) +
    geom_col() +
    geom_vline(xintercept = 0, linewidth = 0.8, color = "black") +
    geom_text(
      aes(x = -outflow, label = paste0("Net: ", scales::comma(net))),
      hjust  = -0.1,
      size   = 3,
      color  = "white",
      data   = . %>% filter(type == "outflow_neg")
    ) +
    scale_fill_manual(
      values = c("inflow" = "#1565C0", "outflow_neg" = "#C62828"),
      labels = c("Inflow to NY", "Outflow from NY")
    ) +
    scale_x_continuous(
      labels = function(x) scales::comma(abs(x)),
      expand = expansion(mult = c(0.15, 0.25))
    ) +
    labs(
      title    = "New York State Migration Flows in Top 5 Destination States (2024)",
      subtitle = "Inflow vs Outflow",
      x        = "Number of People",
      y        = NULL,
      fill     = NULL,
      caption  = "Source: ACS-1, US Census Bureau"
    ) +
    theme_bw() +
    theme(
      legend.position    = "bottom",
      panel.grid.major.y = element_blank()
    )
  
  
# function to create diverging bar chart for a single NY metro area
plot_metro_sources <- function(metro_name) {
  top_sources_ny_metro |>
    filter(full1_name == metro_name) |>
    mutate(
      net         = moved_in - moved_out,
      full2_name  = fct_reorder(full2_name, net),
      outflow_neg = -moved_out
    ) |>
    pivot_longer(
      cols      = c(moved_in, outflow_neg),
      names_to  = "type",
      values_to = "population"
    ) |>
    ggplot(aes(x = population, y = full2_name, fill = type)) +
    geom_col() +
    geom_vline(xintercept = 0, linewidth = 0.8, color = "black") +
    geom_text(
      aes(x = -moved_out, label = paste0("Net: ", scales::comma(net))),
      hjust       = -0.1,
      vjust       = 0.5,
      size        = 3,
      color       = "white",
      data        = . %>% filter(type == "moved_in")
    ) +
    scale_fill_manual(
      values = c("moved_in" = "#1565C0", "outflow_neg" = "#C62828"),
      labels = c("Inflow to NY Metro", "Outflow from NY Metro")
    ) +
    scale_x_continuous(
      labels = function(x) scales::comma(abs(x)),
      expand = expansion(mult = c(0.15, 0.25))
    ) +
    labs(
      title    = glue("Migration Flows: {metro_name}"),
      subtitle = "Inflow (right) vs Outflow (left)",
      x        = "Number of People",
      y        = NULL,
      fill     = NULL,
      caption  = "Source: ACS-5, US Census Bureau"
    ) +
    theme_bw() +
    theme(
      legend.position    = "bottom",
      panel.grid.major.y = element_blank()
    )
}


```

New York is projected to have `r ny_districts_2030` congressional seats in 2030 according to our population projections for the state (~`r comma(round(pop_projections |> filter(year == 2030, state == "New York") |> pull(pop)))`). It is tantalizingly close to a 27th seat with its next seat priority value (`r comma(round(ny_next_seat))`) just 2,293 points below `r seat_435_state`'s 435th seat priority value (`r comma(lowest_allocated)`). This means New York needs only a relatively small net population gain to flip that seat. A rough estimate suggests attracting and retaining around 10,000-15,000 additional residents above the current projection would likely be sufficient to achieve this congressional win for the state.

Below are my proposal on where we can create a promotional campaign that lures potential constituents our way and address the deficit in new New Yorkers. In this I analyze migration flows pertaining to New York state and its three major metro areas, NYC, Buffalo and Rochester.

#### Focus 1: Stem The Flow of Out-Migration

The biggest issue New York faces is population retention. Cost of living in our major metro areas such as NYC has sky rocketed in recent years producing one of the highest median rents in the country. People are possibly leaving for cheaper pastures that stretch the dollar. 

The top states where New Yorkers are escaping to are New Jersey, Florida, Pennsylvania, Texas and Connecticut. Out of all these destination states, New York is a migration net loser meaning more people are leaving New York to that state than people from those states are moving into New York. The biggest hemorrhages are with New Jersey and Florida so we should make these states and the metro areas within them where a significant portion of New Yorkers move to our prime targets.

```{r nys-net-migration-destination-state-plot, echo=FALSE}

top_destination_states_net_plot

```

* **New Jersey**: `r comma(round(top_destinations_ny |> filter(state_current == "New Jersey") |> pull(population)))` leave and `r comma(top_sources_ny |> filter(state_1y == "New Jersey") |> pull(population))` return (net loss of ~`r comma(round(top_destination_states_net |> filter(state == "New Jersey", type == "outflow_neg") |> pull(net)))`). 
  + NYCers account for over half (`r round(nyc_connection |> slice_max(nyc_share, n=1) |> pull(nyc_share), 1)`%) of migration into the `r nyc_connection |> slice_max(nyc_share, n=1) |> pull(full1_name)`.
  + `r top_destinations_ny_metro |> filter(str_detect(full2_name, "NJ"), str_detect(full1_name, "Buffalo")) |> pull(full2_name)` Metro Area is a top moving destination for people from NYC, Rochester and Buffalo.
  
* **Florida**: `r comma(round(top_destinations_ny |> filter(state_current == "Florida") |> pull(population)))` leave and only `r comma(top_sources_ny |> filter(state_1y == "Florida") |> pull(population))` return (net loss of ~`r comma(round(top_destination_states_net |> filter(state == "Florida", type == "outflow_neg") |> pull(net)))`). 
  + Miami is the top moving destination for New Yorkers in the NYC metro heading to Florida. NYCers account for `r round(nyc_connection |> filter(str_detect(full1_name, "Miami")) |> pull(nyc_share), 1)`% of Miami's in-migration. 
  + Tampa is the top moving destination for Buffalo and Rochester residents moving out.

#### Focus 2: Increasing Metro In-Migration

Generally, in-migration to New York is fueled by neighboring northeastern states with significant contributions from California and Florida. Appealing to the distinct cultures of smaller urban metro areas that already provide a large source of in-migration may prove an effective strategy to win over more outsiders. 

```{r ny-metro-net-migration-destination-state-plot, echo=FALSE}

plot_metro_sources("Buffalo-Cheektowaga, NY Metro Area")
plot_metro_sources("Rochester, NY Metro Area")

```

* Philadelphia is the single largest source of in-migration to NYC metro (`r comma(top_sources_ny_metro |> filter(geoid1 == 35620, geoid2 == 37980) |> pull(moved_in))` people) making it a natural target since it's close and culturally similar to New York.
  + **Advertising Opportunity**: Target Philadelphia with messaging about NYC's larger job market and similar, yet more expansive, cultural scene.
* Washington DC sends significant flows to all three major NY metros: #2 in NYC (`r comma(top_sources_ny_metro |> filter(geoid1 == 35620, geoid2 == 47900) |> pull(moved_in))`), #1 in Buffalo (`r comma(top_sources_ny_metro |> filter(geoid1 == 15380, geoid2 == 47900) |> pull(moved_in))`), and #4 in Rochester (`r comma(top_sources_ny_metro |> filter(geoid1 == 40380, geoid2 == 47900) |> pull(moved_in))`) . 
* Boston is a top source for both NYC and Rochester and is home to some of the best universities in America.
  + **Advertising Opportunity**: Target DC with messaging about New York's unique political position and energy. DC and Boston both may also provide a highly educated professional urban population that would be attracted to New York's abundant job opportunities and growth for young workers. 
* San Juan, PR is a surprisingly strong source for both Buffalo and Rochester as the #2 source of in-population in both metro areas in addition to a large population presence already in NYC.
  + **Advertising Opportunity**: Target PR by highlighting the Puerto Rican diaspora and established community in Buffalo and Rochester. Major out-migration from Puerto Rico has been spurred on by lack of job and education opportunities. If upstate metros can establish larger migration patterns of just a few thousand, it would be a game changer for reapportionment.

#### Focus 3: Vulnerable States

Alabama holds the 435th seat with a priority value of 822,578, which is just barely above New York's 27th seat priority value. If New York gains ~10,000-15,000 residents while Alabama loses a similar number, New York can pick up the 27th seat. Other vulnerable single-seat states (where any meaningful out-migration could put their seat at risk) worth monitoring  are:

* Ohio (Priority Value: `r comma(lower_allocation |> filter(state == "Ohio") |> pull(hh_weight))`) 
* North Carolina (Priority Value: `r comma(lower_allocation |> filter(state == "North Carolina") |> pull(hh_weight))`)

These are states where a targeted advertising campaign pulling even a few thousand residents to New York could have an outsized political impact.


#### Proposed State Slogans
Two different slogans were chosen to appeal to both former New Yorkers we hope will come back home and new New Yorkers looking for fresh opportunity. 

**For Florida/Sun Belt leavers:** *"Done with the heat? Come home to New York."*

Juxtaposed with different pictures of heat damage (sunburn, melted street tar, etc) or other extreme weather damage from tropical storms/hurricanes will remind them New York provides safety and comfort all year round. 

**For Philadelphia/DC/Boston upward bound job seekers:** *"The next step up is always New York."*




::: {.callout-note title="AI Usage Statement" collapse="true"}

Claude was used to help write the code in this project and brainstorm final deliverable ideas, but all non-code text was generated without the use of any Generative AI tools.

:::

------------------------------------------------------------------------

This work ©2025 by Evelyn Rodriguez was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #03](https://michael-weylandt.com/STA9750/mini/mini03.html)
