---
title: "Mini-Project 03 - Who Goes There? US Internal Migration and Implications for Congressional Reapportionment"
author: "Evelyn Rodriguez"
editor:
    mode: source
format:
    html:
        toc: true
        code-fold: true
        code-summary: "Display Code"
        code-overflow: wrap
        code-block-bg: true
        code-block-border-left: "#6610f2"
        highlight-style: "a11y"
---

```{r setup, include=FALSE}

library(tidyverse)
library(tidycensus)
library(glue)
library(readxl)
library(httr2)
library(skimr)
library(dplyr)
library(scales)
library(gt)
library(jsonlite)
library(patchwork)
library(shadowtext)

```

## Introduction

What will your state look like in 2030? Many inquiring politicos are hard at work figuring this out and in this mini-project, we will examine [migration flows](https://www.census.gov/topics/population/migration/guidance/migration-flows.html) within the US to forecast the size of state congressional delegations for the 2032 midterm elections. 

paint a better picture of what this could possible look like. Using state-level and metro-level data from the US Census Bureau's [American Community Survey (ACS)](https://en.wikipedia.org/wiki/American_Community_Survey), I'll create demographic projections to predict state populations in 2030.

A suitable introduction goes here. In this mini-project, I intend to demonstrate
my skills at Data Import/API Usage/File Parsing in an analysis of Census Data/Migration Flows data.

## Data Acquisition and Preparation

Code related to Data Acquisition and Preparation goes here.

### Task 1-4

Acquire data from the `tidycensus` package in R to find total population by state from 2015 - 2024.

Note: Data for 2020 is skipped since regular 1-year ACS for 2020 was not released and is not available in `tidycensus`. 

```{r acs-data-acquisition}
#| warning: false 


acquire_acs_data <- function(dataset = c("population", "state_to_state_migration", "metro_to_metro_migration")) {
  
  # Validate dataset argument
  dataset <- match.arg(dataset)
  
  # Define and create the project cache directory if it doesn't already exist
  data_dir <- file.path("data", "mp03")
  if (!dir.exists(file.path("data", "mp03"))) {
    dir.create(file.path("data", "mp03"), showWarnings = FALSE, recursive = TRUE)
  }
  
  # Helper to convert state abbreviations to full names
  # Extends R's built-in state.abb/state.name with DC and Puerto Rico
  abb_to_name <- c(setNames(state.name, state.abb),
                   "DC" = "District of Columbia",
                   "PR" = "Puerto Rico")
  
  expand_state_name <- function(x) {
    ifelse(x %in% names(abb_to_name), abb_to_name[x], x)
  }
  
  switch(dataset,
    
    ## TASK 1 - STATE POPULATION AND BIRTH INFORMATION     
    # --- Population: ACS-1 state-level total population, 2015-2024 (no 2020) ---
    population = {
      
      acs_years <- c(2015:2019, 2021:2024)
      
      POP_ALL <- map(acs_years, function(yy) {
        year_cache_file <- file.path(data_dir, glue("state_population_{yy}.csv"))
        
        if (file.exists(year_cache_file)) {
          message(glue("Loading cached population data for {yy}"))
          return(read_csv(year_cache_file, show_col_types = FALSE))
        }
        
        message(glue("Downloading ACS-1 state population data for {yy}"))
        
        # B01003_001 is the ACS variable for total population
        state_pop_raw <- get_acs(
          variables   = "B01003_001",
          geography   = "state",
          year        = yy,
          survey      = "acs1",
          geometry    = FALSE,
          cache_table = TRUE
        ) |>
          mutate(year = yy) |>
          select(
            state      = NAME,
            population = estimate,
            year
          )
        
        # Cache the cleaned result to disk before returning
        write_csv(state_pop_raw, year_cache_file)
        state_pop_raw
        
      }) |> bind_rows()
      
      POP_ALL
    },
    
    ##TASK 4: COMBINED STATE TO STATE FLOWS
    # --- State-to-State Migration: combines 2024 and 2023 ACS-1 flows ---
    state_to_state_migration = {
      
      ##TASK 2: 2024 STATE TO STATE FLOWS
      # -- Internal function: 2024 flows and stationary population --
      state_to_state_migration_2024 <- function() {
        
        migration_url_2024  <- "https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2024/state-to-state-migration/State_to_State_Migration_Table_2024_T13.xlsx"
        migration_file_2024 <- file.path(data_dir, "State_to_State_Migration_Table_2024_T13.xlsx")
        
        if (!file.exists(migration_file_2024)) {
          message("Downloading 2024 state-to-state migration file...")
          download.file(migration_url_2024, destfile = migration_file_2024, quiet = TRUE, mode = "wb")
        } else {
          message("Loading cached 2024 state-to-state migration file...")
        }
        
        # Table sheet: long format, col A = current state, col B = state 1yr ago, col C = estimate
        # "X" = same-state non-movers (filtered out), "N" = suppressed values (replaced with 0)
        flows_2024 <- read_excel(
          migration_file_2024,
          sheet     = "Table",
          skip      = 8,
          col_names = c("state_current", "state_1y", "population", "moe")
        ) |>
          select(state_current, state_1y, population) |>
          filter(!is.na(state_current), population != "X") |>
          mutate(
            population    = replace(population, population == "N", "0"),
            population    = as.integer(population),
            state_current = expand_state_name(state_current),
            state_1y      = expand_state_name(state_1y),
            year_current  = 2024L,
            year_1y       = 2023L
          ) |>
          filter(state_1y != "Foreign country") |>
          select(state_current, state_1y, population, year_current, year_1y)
        
        # Supplemental sheet: stationary population (did not move)
        # skip = 10 to bypass headers, empty row, and US aggregate row
        # col A = state name, col B = total population; sheet has 11 columns total
        stationary_2024 <- read_excel(
          migration_file_2024,
          sheet     = "Supplemental - Current Res",
          skip      = 10,
          col_names = c("state_current", "population", "moe_1", "est_2", "moe_2",
                        "est_3", "moe_3", "est_4", "moe_4", "est_5", "moe_5")
        ) |>
          select(state_current, population) |>
          filter(!is.na(state_current), !is.na(population)) |>
          mutate(
            population   = as.integer(population),
            state_1y     = state_current,
            year_current = 2024L,
            year_1y      = 2023L
          ) |>
          select(state_current, state_1y, population, year_current, year_1y)
        
        bind_rows(flows_2024, stationary_2024)
      }
      
      ##TASK 3: 2023 STATE TO STATE MIGRATION
      # -- Internal function: 2023 flows and stationary population --
      state_to_state_migration_2023 <- function() {
        
        migration_url_2023  <- "https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2023/state-to-state-migration/State_to_State_Migration_Table_2023_T13.xlsx"
        migration_file_2023 <- file.path(data_dir, "State_to_State_Migration_Table_2023_T13.xlsx")
        
        if (!file.exists(migration_file_2023)) {
          message("Downloading 2023 state-to-state migration file...")
          download.file(migration_url_2023, destfile = migration_file_2023, quiet = TRUE, mode = "wb")
        } else {
          message("Loading cached 2023 state-to-state migration file...")
        }
        
        # tool_input sheet: hidden sheet, long format, col A = state_1y, col B = state_current
        # skip = 1 to skip header row; "N/A" = same-state non-movers (filtered out)
        # "N" = suppressed values (replaced with 0)
        flows_2023 <- read_excel(
          migration_file_2023,
          sheet     = "tool_input",
          skip      = 1,
          col_names = c("state_1y", "state_current", "population", "moe",
                        "blank", "state", "in_est", "in_moe", "out_est", "out_moe",
                        "blank2", "state_names")
        ) |>
          select(state_1y, state_current, population) |>
          filter(!is.na(state_1y), population != "N/A") |>
          mutate(
            population    = replace(population, population == "N", "0"),
            population    = as.integer(population),
            # trim whitespace - DC has trailing space in source data
            state_current = str_trim(state_current),
            state_1y      = str_trim(state_1y),
            year_current  = 2023L,
            year_1y       = 2022L
          ) |>
          filter(state_1y != "Foreign country") |>
          select(state_current, state_1y, population, year_current, year_1y)
        
        # Stationary population from Table sheet
        # range A9:D74 reads only cols A-D starting after 8 header rows
        # col A = state name, col D = same residence estimate
        stationary_2023 <- read_excel(
          migration_file_2023,
          sheet     = "Table",
          range     = "A9:D74",
          col_names = c("state_current", "pop_est", "pop_moe", "population")
        ) |>
          select(state_current, population) |>
          filter(!is.na(state_current), !is.na(population),
                 !state_current %in% c("United States2", "Current residence in")) |>
          mutate(
            population   = as.integer(population),
            state_1y     = state_current,
            year_current = 2023L,
            year_1y      = 2022L
          ) |>
          select(state_current, state_1y, population, year_current, year_1y)
        
        bind_rows(flows_2023, stationary_2023)
      }
      
      # Combine both years into one tidy data frame
      bind_rows(state_to_state_migration_2024(), state_to_state_migration_2023())
    },
    # --- Metro-to-Metro Migration: ACS-5 2016-2020 metro-level flows ---
    metro_to_metro_migration = {
      
      # API URL for ACS-5 2016-2020 metro-to-metro migration flows
      # Requests MOVEDIN, MOVEDOUT, FULL1_NAME, FULL2_NAME, GEOID1, GEOID2
      # for all metropolitan/micropolitan statistical areas
      metro_migration_url  <- "https://api.census.gov/data/2020/acs/flows?get=MOVEDIN,MOVEDOUT,FULL1_NAME,FULL2_NAME,GEOID1,GEOID2&for=metropolitan%20statistical%20area/micropolitan%20statistical%20area:*"
      metro_migration_file <- file.path(data_dir, "metro_to_metro_migration.json")
      
      if (!file.exists(metro_migration_file)) {
        message("Downloading metro-to-metro migration data from Census API...")
        download.file(metro_migration_url, destfile = metro_migration_file, quiet = TRUE, mode = "wb")
      } else {
        message("Loading cached metro-to-metro migration data...")
      }
      
      # read_json() interprets first row as data instead of column names
      # move_row_to_colnames() fixes this by promoting first row to column headers
      move_row_to_colnames <- function(X, row = 1) {
        X <- as_tibble(X)
        Xrow <- X[row, ]
        X <- X[-row, ]
        colnames(X) <- Xrow
        readr::type_convert(X)
      }
      
      # Extracts the primary state abbreviation from a metro area name
      # For single-state metros (e.g. "Los Angeles-Long Beach, CA Metro Area") -> "CA"
      # For multi-state metros (e.g. "New York-Newark-Jersey City, NY-NJ-PA Metro Area") -> "NY"
      pull_state_from_metro <- function(metro_name) {
        str_extract(metro_name, ".*, (\\S{2})[-[:alpha:]]* Metro Area", group = 1)
      }
      
      read_json(metro_migration_file, simplifyVector = TRUE) |>
        as.data.frame() |>
        move_row_to_colnames() |>
        select(
          geoid1     = GEOID1,
          geoid2     = GEOID2,
          full1_name = FULL1_NAME,
          full2_name = FULL2_NAME,
          moved_in   = MOVEDIN,
          moved_out  = MOVEDOUT
        ) |>
        filter(!full2_name %in% c("Africa", "Asia", "Caribbean", "Central America", 
                           "Europe", "Northern America", "Oceania and At Sea", 
                           "South America", "U.S. Island Areas")) |>
        mutate(
          state1 = pull_state_from_metro(full1_name),
          state2 = pull_state_from_metro(full2_name)
        )
    }
  )
}

STATE_POP                <- acquire_acs_data("population")
STATE_TO_STATE_MIGRATION <- acquire_acs_data("state_to_state_migration")
METRO_TO_METRO_MIGRATION <- acquire_acs_data("metro_to_metro_migration")

glimpse(STATE_POP)
glimpse(STATE_TO_STATE_MIGRATION)
glimpse(METRO_TO_METRO_MIGRATION)

```

*Note:* My data frame will have more than 5000 rows because I've taken District of Columbia (DC), Puerto Rico and U.S. Island Areas into consideration and not just the 50 mainland(-ish) states. 

Migrants from Puerto Rico have an automatic right to vote and we can represent significant voting blocks in states such as New York, Connecticut and Florida. Same goes for U.S. Island Areas.  
Excluded migration flows of people from foreign countries since it is not a guarantee that they will be able to vote due to legal status. 


Checking if Data is correct for Task #1

```{r initial-data-check, eval=FALSE}

dim(STATE_POP)
glimpse(STATE_POP)
head(STATE_POP)

#should be 9 years
distinct(STATE_POP, year)

#should be 52 states
STATE_POP |> count(year)

#should be california
STATE_POP |> group_by(year) |> slice_max(population)

#should be wyoming
STATE_POP |> group_by(year) |> slice_min(population)

#should increase
STATE_POP |> group_by(year) |> summarize(total_population = sum(population))
STATE_POP |> filter(state == "California", year == 2023)
STATE_POP |> summarize(across(everything(), ~sum(is.na(.))))
STATE_POP |> filter(population < 100000 | population > 50000000)
```



Remember that code blocks look like this: 

```{r}
# Your code goes in chunks like this
library(tidyverse) # You will want this line for almost all MPs
x <- 1 + 2 + 3
```

and you can then print variables from chunks: 

```{r}
x
```

or inline, like this: $x$ is `r x`.

## Exploratory Analysis

### EDA: General

**1. Which states have had the highest net population growth rates over the past decade?**

```{r highest-net-pop-growth}

top_net_pop_growth_state <- STATE_POP |>
  filter(year %in% c(2015, 2024)) |>
  pivot_wider(
    names_from   = year,
    values_from  = population,
    names_prefix = "pop_"
  ) |>
  mutate(
    net_growth      = pop_2024 - pop_2015,
    net_growth_rate = (pop_2024 - pop_2015) / pop_2015 * 100
  ) |>
  arrange(desc(net_growth_rate)) |>
  select(state, pop_2015, pop_2024, net_growth, net_growth_rate) |>
  slice_max(net_growth_rate, n = 10)

#table
top_net_growth_table <- top_net_pop_growth_state |>
  gt() |>
  tab_header(
    title    = md("**Top 10 States by Net Population Growth Rate (2015-2024)**"),
    subtitle = "Based on ACS-1 annual estimates"
  ) |>
  cols_label(
    state           = md("**State**"),
    pop_2015        = md("**Population<br>(2015)**"),
    pop_2024        = md("**Population<br>(2024)**"),
    net_growth      = md("**Net Growth**"),
    net_growth_rate = md("**Growth Rate (%)**")
  ) |>
  cols_align(
    align   = "center",
    columns = c(pop_2015, pop_2024, net_growth, net_growth_rate)
  ) |>
  fmt_number(
    columns  = c(pop_2015, pop_2024, net_growth),
    decimals = 0
  ) |>
  fmt_number(
    columns  = net_growth_rate,
    decimals = 2
  ) |>
  data_color(
    columns = net_growth_rate,
    palette = c("#90CAF9", "#1565C0")
  ) |>
  tab_source_note(
    source_note = "Source: ACS-1 Baseline State Population Data, US Census Bureau"
  )


# bar chart
top_net_growth_bar <- top_net_pop_growth_state |>
  mutate(state = fct_reorder(state, net_growth_rate)) |>
  ggplot(aes(x = net_growth_rate, y = state, fill = net_growth_rate)) +
  geom_col() +
  geom_vline(xintercept = 0, linewidth = 0.8, color = "black") +
  geom_text(
    aes(label = paste0(round(net_growth_rate, 1), "%")),
    hjust    = -0.2,
    size     = 3
  ) +
  scale_fill_gradient(low = "#90CAF9", high = "#1565C0") +
  labs(
    title    = "Top 15 States by Net Population Growth Rate (2015-2024)",
    subtitle = "Percentage change in total population",
    x        = "Net Growth Rate (%)",
    y        = NULL,
    fill     = "Growth Rate (%)",
    caption  = "Source: ACS-1, US Census Bureau"
  ) +
  theme_minimal() +
  theme(
    legend.position    = "bottom",
    panel.grid.major.y = element_blank()
  )


top_net_growth_facet <- STATE_POP |>
  filter(state %in% top_net_pop_growth_state$state) |>
  ggplot(aes(x = year, y = population, color = state)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = c(2015, 2018, 2021, 2024)) +
  facet_wrap(~state, scales = "free_y") +
  labs(
    title    = "Population Change Over Time for Top 10 Fastest Growing States",
    subtitle = "ACS-1 annual estimates, 2015-2024 (excluding 2020)",
    x        = NULL,
    y        = "Total Population",
    color    = NULL,
    caption  = "Source: ACS-1, US Census Bureau"
  ) +
  theme_minimal() +
  theme(
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 6),
    panel.spacing    = unit(1.5, "lines")
  )


```

**2. If you meet someone who moved to New York State in the last year, what are the most likely states they moved from? If one of your friends announces they are moving out of the state, where are they most likely to move?**

```{r most-likely-states-moved-to-from-NY}

# most likely states people moved FROM into New York
moved_to_ny <- STATE_TO_STATE_MIGRATION |>
  filter(
    state_current == "New York",
    state_1y != "New York",
    year_current == 2024
  ) |>
  arrange(desc(population)) |>
  select(state_1y, population) |>
# %age out of total migration to NY
  mutate(pct = population / sum(population) * 100) |>
  slice_max(population, n = 5)

# most likely states people moved TO from New York
moved_from_ny <- STATE_TO_STATE_MIGRATION |>
  filter(
    state_1y == "New York",
    state_current != "New York",
    year_current == 2024
  ) |>
  arrange(desc(population)) |>
  select(state_current, population) |>
  mutate(pct = population / sum(population) * 100) |>
  slice_max(population, n = 5)

# top 5 states people moved from into New York
plot_to_ny <- moved_to_ny |>
  ggplot(aes(x = population, y = fct_reorder(state_1y, population), fill = population)) +
  geom_col() +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    hjust = -0.2,
    size  = 3
  ) +
  scale_fill_gradient(low = "#90CAF9", high = "#1565C0") +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title   = "Top 5 States People Moved From Into New York (2024)",
    x       = "Number of People",
    y       = NULL,
    caption = "Source: ACS-1, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position    = "none",
    panel.grid.major.y = element_blank()
  )

# top 5 states people moved to from New York
plot_from_ny <- moved_from_ny |>
  ggplot(aes(x = population, y = fct_reorder(state_current, population), fill = population)) +
  geom_col() +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    hjust = -0.2,
    size  = 3
  ) +
  scale_fill_gradient(low = "#FFCDD2", high = "#C62828") +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title   = "Top 5 States People Moved To From New York (2024)",
    x       = "Number of People",
    y       = NULL,
    caption = "Source: ACS-1, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position    = "none",
    panel.grid.major.y = element_blank()
  )

# stack vertically using patchwork
plot_to_ny / plot_from_ny

```

The top 5 states people moved from into New York in 2024 were New Jersey, California, Florida, Pennsylvania and Massachusetts.
The top 5 states New Yorkers moved to in 2024 were New Jersey, Florida, Pennsylvania, Texas and Connecticut.

**3. If you meet someone who moved to New York City in the last year, what are the most likely metro areas they moved from? If one of your friends announces they are moving out of the city, where are they most likely to move?**

```{r most-likely-states-moved-to-from-nyc}

nyc_metro <- "New York-Newark-Jersey City, NY-NJ-PA Metro Area"

# top 5 metro areas people moved FROM into NYC
moved_to_nyc <- METRO_TO_METRO_MIGRATION |>
  filter(
    full1_name == nyc_metro,
    full2_name != nyc_metro,
# excluded outside metro area since these are places not considered Metro Areas    
    full2_name != "Outside Metro Area within U.S. or Puerto Rico"
  ) |>
  mutate(pct = moved_in / sum(moved_in) * 100) |>
  arrange(desc(moved_in)) |>
  slice_max(moved_in, n = 5) |>
  mutate(full2_name = str_replace(full2_name, "(.*),\\s*(\\S+)\\s*Metro Area", "\\1, \\2"))

# top 5 metro areas people moved TO from NYC
moved_from_nyc <- METRO_TO_METRO_MIGRATION |>
  filter(
    full1_name == nyc_metro,
    full2_name != nyc_metro,
# excluded outside metro area since these are places not considered Metro Areas
    full2_name != "Outside Metro Area within U.S. or Puerto Rico"
  ) |>
  mutate(pct = moved_out / sum(moved_out) * 100) |>
  arrange(desc(moved_out)) |>
  slice_max(moved_out, n = 5) |>
  mutate(full2_name = str_replace(full2_name, "(.*),\\s*(\\S+)\\s*Metro Area", "\\1, \\2"))

plot_to_nyc <- moved_to_nyc |>
  ggplot(aes(x = moved_in, y = fct_reorder(full2_name, moved_in), fill = moved_in)) +
  geom_col() +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    hjust = -0.2,
    size  = 3
  ) +
  scale_fill_gradient(low = "#90CAF9", high = "#1565C0") +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title   = "Top 5 Metro Areas People Moved From Into NYC",
    x       = "Number of People",
    y       = NULL,
    caption = "Source: ACS-5 2016-2020 Metropolitan Flows, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position    = "none",
    panel.grid.major.y = element_blank()
  )

plot_from_nyc <- moved_from_nyc |>
  ggplot(aes(x = moved_out, y = fct_reorder(full2_name, moved_out), fill = moved_out)) +
  geom_col() +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    hjust = -0.2,
    size  = 3
  ) +
  scale_fill_gradient(low = "#FFCDD2", high = "#C62828") +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title   = "Top 5 Metro Areas People Moved To From NYC",
    x       = "Number of People",
    y       = NULL,
    caption = "Source: ACS-5 2016-2020 Metropolitan Flows, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position    = "none",
    panel.grid.major.y = element_blank()
  )

plot_to_nyc / plot_from_nyc


```

In both cases, the Philadelphia metro area is where someone most likely would have moved to NYC from and where New Yorkers are most likely to move to. 


**4. Which states have had the highest amounts of in-migration and out-migration? Furthermore, which states have had the highest amount of net migration? To get total migration in a direction, add up the total migration in that direction: e.g., the total in-migration to Alabama is the sum of people who live in Alabama in 2024 but lived elsewhere in 2023 (this is a sum over 49 other states).Net migration is simply total in-migration minus total out-migration.**

```{r in-out-net-migration-state}

# total in-migration per state (people who moved INTO that state)
in_migration <- STATE_TO_STATE_MIGRATION |>
  filter(
    state_current != state_1y,
    year_current  == 2024
  ) |>
  group_by(state_current) |>
  summarize(total_in = sum(population))



# total out-migration per state (people who moved OUT of that state)
out_migration <- STATE_TO_STATE_MIGRATION |>
  filter(
    state_current != state_1y,
    year_current  == 2024
  ) |>
  group_by(state_1y) |>
  summarize(total_out = sum(population))

# combine and compute net migration
net_migration <- in_migration |>
  inner_join(out_migration, join_by(state_current == state_1y)) |>
  mutate(net = total_in - total_out) |>
  rename(state = state_current)

# plot 1: top 10 states by in-migration
plot_in_migration <- net_migration |>
  slice_max(total_in, n = 10) |>
  mutate(state = fct_reorder(state, total_in)) |>
  ggplot(aes(x = total_in, y = state, fill = total_in)) +
  geom_col() +
  geom_text(
    aes(label = scales::comma(total_in)),
    hjust = 1.2,
    size  = 3,
    color = "white"
  ) +
  scale_fill_gradient(low = "#90CAF9", high = "#1565C0") +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title   = "States by Total In-Migration",
    subtitle = "Top 10 (2024)",
    x       = "Number of People",
    y       = NULL,
    caption = "Source: ACS-1, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position    = "none",
    panel.grid.major.y = element_blank()
  )

# plot 2: top 10 states by out-migration
plot_out_migration <- net_migration |>
  slice_max(total_out, n = 10) |>
  mutate(state = fct_reorder(state, total_out)) |>
  ggplot(aes(x = total_out, y = state, fill = total_out)) +
  geom_col() +
  geom_text(
    aes(label = scales::comma(total_in)),
    hjust = 1.2,
    size  = 3,
    color = "white"
  ) +
  scale_fill_gradient(low = "#E1BEE7", high = "#4A148C") +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title   = "States by Total Out-Migration",
    subtitle = "Top 10 (2024)",
    x       = "Number of People",
    y       = NULL,
    caption = "Source: ACS-1, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position    = "none",
    panel.grid.major.y = element_blank()
  )

net_migration_table <- net_migration |>
  slice_max(net, n = 10) |>
  arrange(desc(net)) |>
  select(state, total_in, total_out, net) |>
  gt() |>
  tab_header(
    title = md("**Top 10 States by Net Migration (2024)**")
  ) |>
  cols_label(
    state     = md("**State**"),
    total_in  = md("**Total In-Migration**"),
    total_out = md("**Total Out-Migration**"),
    net       = md("**Net Migration**")
  ) |>
  cols_align(
    align   = "center",
    columns = c(total_in, total_out, net)
  ) |>
  fmt_number(
    columns  = c(total_in, total_out, net),
    decimals = 0
  ) |>
  data_color(
    columns = net,
    palette = c("#C8E6C9", "#1B5E20")
  ) |>
  tab_source_note(
    source_note = "Source: ACS-1, US Census Bureau"
  ) |>
  tab_options(
    heading.align = "center"
  )

plot_in_migration | plot_out_migration

net_migration_table

```


**5. Which metro areas have had the highest amounts of total in-migration and out-migration? Furthermore, which metro areas have had the highest amount of net migration?**

```{r in-out-net-migration-metro}

# total in-migration per metro area
metro_in_migration <- METRO_TO_METRO_MIGRATION |>
  filter(full1_name != full2_name) |>
  group_by(full1_name) |>
  summarize(total_in = sum(moved_in, na.rm = TRUE))

# total out-migration per metro area
metro_out_migration <- METRO_TO_METRO_MIGRATION |>
  filter(full1_name != full2_name) |>
  group_by(full1_name) |>
  summarize(total_out = sum(moved_out, na.rm = TRUE))

# combine and compute net migration
metro_net_migration <- metro_in_migration |>
  inner_join(metro_out_migration, join_by(full1_name)) |>
  mutate(
    net       = total_in - total_out,
    full1_name = str_replace(full1_name, "(.*),\\s*(\\S+)\\s*Metro Area", "\\1, \\2")
  ) |>
  rename(metro = full1_name)

# # top 10 metro areas by in-migration
# metro_net_migration |>
#   slice_max(total_in, n = 10)
# 
# # top 10 metro areas by out-migration
# metro_net_migration |>
#   slice_max(total_out, n = 10)
# 
# # top 10 metro areas by net migration
# metro_net_migration |>
#   slice_max(net, n = 10)

# plot 1: top 10 metro areas by in-migration
plot_metro_in <- metro_net_migration |>
  slice_max(total_in, n = 10) |>
  mutate(
    metro = paste0(str_extract(metro, "^[^-]+"), ", ", str_extract(metro, ",\\s*(\\w{2})", group = 1)),
    metro = fct_reorder(metro, total_in)
  ) |>
  ggplot(aes(x = total_in, y = metro, fill = total_in)) +
  geom_col() +
  geom_text(
    aes(label = scales::comma(total_in)),
    hjust = 1.2,
    size = 3,
    color = "white"
  ) +
  scale_fill_gradient(low = "#90CAF9", high = "#1565C0") +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Metro Areas by In-Migration",
    subtitle = "Top 10 (2016-2020)",
    x = "Number of People",
    y = NULL,
    caption = "Source: ACS-5, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank()
  )

# plot 2: top 10 metro areas by out-migration
plot_metro_out <- metro_net_migration |>
  slice_max(total_out, n = 10) |>
  mutate(
    metro = paste0(str_extract(metro, "^[^-]+"), ", ", str_extract(metro, ",\\s*(\\w{2})", group = 1)),
    metro = fct_reorder(metro, total_out)
  ) |>
  ggplot(aes(x = total_out, y = metro, fill = total_out)) +
  geom_col() +
  geom_text(
    aes(label = scales::comma(total_out)),
    hjust = 1.2,
    size  = 3,
    color = "white"
  ) +
  scale_fill_gradient(low = "#E1BEE7", high = "#4A148C") +
  scale_x_continuous(
    labels = scales::comma,
    breaks = seq(0, max(metro_net_migration$total_out), by = 200000),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Metro Areas by Out-Migration",
    subtitle = "Top 10 (2016-2020)",
    x = "Number of People",
    y = NULL,
    caption = "Source: ACS-5, US Census Bureau"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank()
  )


# gt table for top 10 metro areas by net migration
metro_net_table <- metro_net_migration |>
  slice_max(net, n = 10) |>
  arrange(desc(net)) |>
  select(metro, total_in, total_out, net) |>
  gt() |>
  tab_header(
    title = "Top 10 Metro Areas by Net Migration (2016-2020)"
  ) |>
  cols_label(
    metro     = "Metro Area",
    total_in  = "Total In-Migration",
    total_out = "Total Out-Migration",
    net       = "Net Migration"
  ) |>
  cols_align(
    align   = "center",
    columns = c(total_in, total_out, net)
  ) |>
  fmt_number(
    columns  = c(total_in, total_out, net),
    decimals = 0
  ) |>
  data_color(
    columns = net,
    palette = c("#C8E6C9", "#1B5E20")
  ) |>
  tab_source_note(
    source_note = "Source: ACS-5, US Census Bureau"
  ) |>
  tab_options(
    heading.align = "center"
  )

# display in and out migration side by side
plot_metro_in | plot_metro_out

metro_net_table

```


**6. Which states have the highest fraction of residents who lived there last year? (That is, which states have the lowest proportional total in-migration?)**

```{r state-pop-stability}

# total state populations
state_pop_2024 <- STATE_POP |>
  filter(year == 2024) |>
  select(state, total_pop = population)

# calculate in-migration fraction per state
state_stability <- STATE_TO_STATE_MIGRATION |>
  filter(year_current == 2024) |>
  group_by(state_current) |>
  summarize(in_migration = sum(population[state_current != state_1y])) |>
  rename(state = state_current) |>
  left_join(state_pop_2024, join_by(state)) |>
  mutate(
    stability_rate   = 1 - (in_migration / total_pop),
    in_migration_pct = in_migration / total_pop * 100
  ) |>
  filter(state != "Puerto Rico")

state_stability_table <- state_stability |>
  slice_max(stability_rate, n = 10) |>
  arrange(desc(stability_rate)) |>
  select(state, total_pop, in_migration, stability_rate, in_migration_pct) |>
  gt() |>
  tab_header(
    title = md("**Top States by Resident Stability Rate**"),
    subtitle = md("States with Lowest Proportional Total In-Migration (2024)")
  ) |>
  cols_label(
    state = "State",
    total_pop = "Total Population",
    in_migration = "Total In-Migration",
    stability_rate = "Stability Rate",
    in_migration_pct = "In-Migration (%)"
  ) |>
  cols_align(
    align = "center",
    columns = c(total_pop, in_migration, stability_rate, in_migration_pct)
  ) |>
  fmt_number(
    columns = c(total_pop, in_migration),
    decimals = 0
  ) |>
  fmt_percent(
    columns = stability_rate,
    decimals = 2
  ) |>
  fmt_number(
    columns = in_migration_pct,
    decimals = 2
  ) |>
  data_color(
    columns = stability_rate,
    palette = c("#C8E6C9", "#1B5E20")
  ) |>
  tab_source_note(
    source_note = "Source: ACS-1, US Census Bureau"
  ) |>
  tab_options(
    heading.align = "center"
  )

state_stability_table

# top 10 states with highest fraction of long-term residents
# state_stability |>
#   slice_max(stability_rate, n = 10) |>
#   arrange(desc(stability_rate))

```


**7. Which state has the highest fraction of its population growth attributable to net internal migration? That is, determine the total population growth of each state and see what fraction of it corresponds to net migration (as opposed to deaths, births, or international migration).**

```{r pop-growth-net-migration}

# compute population growth from 2023 to 2024 using STATE_POP
pop_growth <- STATE_POP |>
  filter(year %in% c(2023, 2024)) |>
  pivot_wider(
    names_from = year,
    values_from = population,
    names_prefix = "pop_"
  ) |>
  mutate(pop_growth = pop_2024 - pop_2023)

# combine with net_migration (from question #4) and compute fraction of growth attributable to net migration
migration_growth_share <- pop_growth |>
  left_join(net_migration, join_by(state)) |>
  mutate(
    migration_share = net / pop_growth * 100
  ) |>
  filter(state != "Puerto Rico") |>
  select(state, pop_2023, pop_2024, pop_growth, net, migration_share)

top_migration_growth_state <- migration_growth_share |>
  slice_max(migration_share, n = 1)

```

The state with the highest fraction of population growth attributable to net internal migration is `r top_migration_growth_state$state`, with `r round(top_migration_growth_state$migration_share, 1)`% of its growth coming from internal migration.

***Note:** Migration Share can exceed 100% if net migration is larger than total population growth (meaning natural growth or international migration was negative), or be negative if net migration was negative while the state still grew overall.*

This is actually an interesting finding in itself — it means some states are growing purely because of internal migration despite losing population to deaths or international emigration. 
For ex: 
Net internal migration: +100,000 (people moving in from other states)
Natural increase: -20,000 (more deaths than births, common in older states)
Net international migration: -10,000

Then total population growth would be +100,000 - 20,000 - 10,000 = +70,000, but net migration alone was +100,000. So migration_share = 100,000 / 70,000 * 100 = 142.9% — exceeding 100% because the other components partially offset the migration gains.


### EDA: New York

**8. Which state had the highest amount of migration into your state? Of the people who moved out of your state, what was the most popular destination state?**

```{r nys-state-migration-flow}

# state with highest amount of migration into New York
top_inflow_to_ny <- STATE_TO_STATE_MIGRATION |>
  filter(
    state_current == "New York",
    state_1y != "New York",
    year_current == 2024
  ) |>
  slice_max(population, n = 1)

# most popular destination state for people leaving New York
top_outflow_from_ny <- STATE_TO_STATE_MIGRATION |>
  filter(
    state_1y      == "New York",
    state_current != "New York",
    year_current  == 2024
  ) |>
  slice_max(population, n = 1)

```
The state with the highest amount of migration into New York was `r top_inflow_to_ny$state_1y`, with `r scales::comma(top_inflow_to_ny$population)` people moving in.

Of the people who moved out of New York, the most popular destination was `r top_outflow_from_ny$state_current`, with `r scales::comma(top_outflow_from_ny$population)` people moving there.


**9. What is the largest metropolitan area in your state? (You can answer this just using background knowledge; you do not have to use Census data here.) Which metro area, not located in your state, has the highest amount of migration into your state’s largest metro? To which metro area outside of your state do the most residents of your largest metro area move?**

```{r nys-metro-migration-flow}

nyc_metro <- "New York-Newark-Jersey City, NY-NJ-PA Metro Area"

# metro area outside New York with highest migration into NYC metro
top_inflow_to_nyc <- METRO_TO_METRO_MIGRATION |>
  filter(
    full1_name == nyc_metro,
    full2_name != nyc_metro,
    state2 != "NY",
    full2_name != "Outside Metro Area within U.S. or Puerto Rico"
  ) |>
  slice_max(moved_in, n = 1)

# most popular destination metro outside New York for NYC metro residents
top_outflow_from_nyc <- METRO_TO_METRO_MIGRATION |>
  filter(
    full1_name == nyc_metro,
    full2_name != nyc_metro,
    state2 != "NY",
    full2_name != "Outside Metro Area within U.S. or Puerto Rico"
  ) |>
  slice_max(moved_out, n = 1)

```

The largest metro area is NYC, which is contained in the *New York-Newark-Jersey City, NY-NJ-PA Metro Area* in our Census data.

The metro area outside of New York with the highest migration into the NYC metro was `r top_inflow_to_nyc$full2_name`, with `r scales::comma(top_inflow_to_nyc$moved_in)` people moving in.

The most popular destination metro outside of New York for NYC metro residents was `r top_outflow_from_nyc$full2_name`, with `r scales::comma(top_outflow_from_nyc$moved_out)` people moving there.


**10. Are there any metro areas that have a particularly high connection to your state’s largest metro? (E.g., NYCers were over 15% of people moving into the Miami Metro area.)**

```{r nyc-metro-area-connections}

nyc_connection <- METRO_TO_METRO_MIGRATION |>
  filter(
    full2_name == nyc_metro,
    full1_name != nyc_metro,
    full1_name != "Outside Metro Area within U.S. or Puerto Rico"
  ) |>
  left_join(
    # get total in-migration for each metro area
    METRO_TO_METRO_MIGRATION |>
      filter(full1_name != full2_name) |>
      group_by(full1_name) |>
      summarize(total_in = sum(moved_in, na.rm = TRUE)),
    join_by(full1_name)
  ) |>
  mutate(
    nyc_share = moved_in / total_in * 100
  ) |>
  filter(nyc_share >= 15) |>
  arrange(desc(nyc_share)) |>
  select(full1_name, moved_in, total_in, nyc_share)


```

Outside of NYC but within New York State, NYCers account for more than 15% of people moving into these upstate metros: 


## Final Insights and Deliverable

Code related to the final deliverable of the assignment goes here. 


::: {.callout-note title="AI Usage Statement" collapse="true"}

Claude was used to help write the code in this project and brainstorm final deliverable ideas, but all non-code text was generated without the use of any Generative AI tools.

:::

------------------------------------------------------------------------

This work ©2025 by Evelyn Rodriguez was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #03](https://michael-weylandt.com/STA9750/mini/mini03.html)
