---
title: "Mini-Project 03 - Who Goes There? US Internal Migration and Implications for Congressional Reapportionment"
author: "Evelyn Rodriguez"
editor:
    mode: source
format:
    html:
        toc: true
        code-fold: true
        code-summary: "Display Code"
        code-overflow: wrap
        code-block-bg: true
        code-block-border-left: "#6610f2"
        highlight-style: "a11y"
---

```{r setup, include=FALSE}

library(tidyverse)
library(glue)
library(readxl)
library(httr2)
library(skimr)
library(dplyr)
library(scales)
library(gt)
library(tidycensus)

```

## Introduction

What will your state look like in 2030? Many inquiring politicos are hard at work figuring this out and in this mini-project, we will examine [migration flows](https://www.census.gov/topics/population/migration/guidance/migration-flows.html) within the US to forecast the size of state congressional delegations for the 2032 midterm elections. 

paint a better picture of what this could possible look like. Using state-level and metro-level data from the US Census Bureau's [American Community Survey (ACS)](https://en.wikipedia.org/wiki/American_Community_Survey), I'll create demographic projections to predict state populations in 2030.

A suitable introduction goes here. In this mini-project, I intend to demonstrate
my skills at Data Import/API Usage/File Parsing in an analysis of Census Data/Migration Flows data.

## Data Acquisition and Preparation

Code related to Data Acquisition and Preparation goes here.

### Task 1

Acquire data from the `tidycensus` package in R to find total population by state from 2015 - 2024.
Note: We need to skip data for 2020 since regular 1-year ACS for 2020 was not released and is not available in `tidycensus`. 

```{r acs-data-acquisition}
#| warning: false 

## TASK 1 - STATE POPULATION AND BIRTH INFORMATION

acquire_acs_population <- function() {
# Define and create project cache directory if it doesn't already exist
  data_dir <- file.path("data", "mp03")
  
  if (!dir.exists(file.path("data", "mp03"))) {
    dir.create(file.path("data", "mp03"), showWarnings = FALSE, recursive = TRUE)
  }

#Exclude 2020 - standard ACS-1 data not released due to COVID  
  acs_years <- c(2015:2019, 2021:2024)

#Iterate over each year, loading from cache or download as needed    
  POP_ALL <- map(acs_years, function(yy) {
    cache_file <- file.path(data_dir, glue("state_population_{yy}.csv"))

#return cached file if it exists, otherwise download and cache    
    if (file.exists(cache_file)) {
      message(glue("Loading cached population data for {yy}"))
      return(read_csv(cache_file, show_col_types = FALSE))
    }
    
    message(glue("Downloading ACS-1 population data for {yy}"))
    
    state_pop_raw <- get_acs(
  # B01003_001 is the ACS variable for total population
      variables   = "B01003_001",
      geography   = "state",
      year        = yy,
      survey      = "acs1",
      geometry    = FALSE,
      cache_table = TRUE
    ) |>
      mutate(year = yy) |>
      select(
        state      = NAME,
        population = estimate,
        year
      )
#cache the cleaned result to disk before returning    
    write_csv(state_pop_raw, cache_file)
    state_pop_raw
    
  }) |> bind_rows()
  
  POP_ALL
}

STATE_POP <- acquire_acs_population()


```

Checking if Data is correct for Task #1

```{r initial-data-check, eval=FALSE}

dim(STATE_POP)
glimpse(STATE_POP)
head(STATE_POP)

#should be 9 years
distinct(STATE_POP, year)

#should be 52 states
STATE_POP |> count(year)

#should be california
STATE_POP |> group_by(year) |> slice_max(population)
#should be wyoming
STATE_POP |> group_by(year) |> slice_min(population)
#should increase
STATE_POP |> group_by(year) |> summarize(total_population = sum(population))
STATE_POP |> filter(state == "California", year == 2023)
STATE_POP |> summarize(across(everything(), ~sum(is.na(.))))
STATE_POP |> filter(population < 100000 | population > 50000000)
```



Remember that code blocks look like this: 

```{r}
# Your code goes in chunks like this
library(tidyverse) # You will want this line for almost all MPs
x <- 1 + 2 + 3
```

and you can then print variables from chunks: 

```{r}
x
```

or inline, like this: $x$ is `r x`.

## Exploratory Analysis

Code related to Exploratory Data Analysis goes here. This may include
exploratory graphics, instructor-provided exploratory questions, or other
similar elements. 

## Final Insights and Deliverable

Code related to the final deliverable of the assignment goes here. 


::: {.callout-note title="AI Usage Statement" collapse="true"}

Claude was used to help write the code in this project and brainstorm final deliverable ideas, but all non-code text was generated without the use of any Generative AI tools.

:::

------------------------------------------------------------------------

This work ©2025 by Evelyn Rodriguez was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #03](https://michael-weylandt.com/STA9750/mini/mini03.html)
