---
title: "Mini-Project 03 - Who Goes There? US Internal Migration and Implications for Congressional Reapportionment"
author: "Evelyn Rodriguez"
editor:
    mode: source
format:
    html:
        toc: true
        code-fold: true
        code-summary: "Display Code"
        code-overflow: wrap
        code-block-bg: true
        code-block-border-left: "#6610f2"
        highlight-style: "a11y"
---

```{r setup, include=FALSE}

library(tidyverse)
library(tidycensus)
library(glue)
library(readxl)
library(httr2)
library(skimr)
library(dplyr)
library(scales)
library(gt)

```

## Introduction

What will your state look like in 2030? Many inquiring politicos are hard at work figuring this out and in this mini-project, we will examine [migration flows](https://www.census.gov/topics/population/migration/guidance/migration-flows.html) within the US to forecast the size of state congressional delegations for the 2032 midterm elections. 

paint a better picture of what this could possible look like. Using state-level and metro-level data from the US Census Bureau's [American Community Survey (ACS)](https://en.wikipedia.org/wiki/American_Community_Survey), I'll create demographic projections to predict state populations in 2030.

A suitable introduction goes here. In this mini-project, I intend to demonstrate
my skills at Data Import/API Usage/File Parsing in an analysis of Census Data/Migration Flows data.

## Data Acquisition and Preparation

Code related to Data Acquisition and Preparation goes here.

### Task 1

Acquire data from the `tidycensus` package in R to find total population by state from 2015 - 2024.
Note: We need to skip data for 2020 since regular 1-year ACS for 2020 was not released and is not available in `tidycensus`. 

```{r acs-data-acquisition}
#| warning: false 


acquire_acs_data <- function(dataset = c("population", "state_to_state_migration")) {
  
  # Validate dataset argument
  dataset <- match.arg(dataset)
  
  # Define and create the project cache directory if it doesn't already exist
  data_dir <- file.path("data", "mp03")
  if (!dir.exists(file.path("data", "mp03"))) {
    dir.create(file.path("data", "mp03"), showWarnings = FALSE, recursive = TRUE)
  }
  
  # Helper to convert state abbreviations to full names
  # Extends R's built-in state.abb/state.name with DC and Puerto Rico
  abb_to_name <- c(setNames(state.name, state.abb),
                   "DC" = "District of Columbia",
                   "PR" = "Puerto Rico")
  
  expand_state_name <- function(x) {
    ifelse(x %in% names(abb_to_name), abb_to_name[x], x)
  }
  
  switch(dataset,
    
    ## TASK 1 - STATE POPULATION AND BIRTH INFORMATION     
    # --- Population: ACS-1 state-level total population, 2015-2024 (no 2020) ---
    population = {
      
      acs_years <- c(2015:2019, 2021:2024)
      
      POP_ALL <- map(acs_years, function(yy) {
        year_cache_file <- file.path(data_dir, glue("state_population_{yy}.csv"))
        
        if (file.exists(year_cache_file)) {
          message(glue("Loading cached population data for {yy}"))
          return(read_csv(year_cache_file, show_col_types = FALSE))
        }
        
        message(glue("Downloading ACS-1 state population data for {yy}"))
        
        # B01003_001 is the ACS variable for total population
        state_pop_raw <- get_acs(
          variables   = "B01003_001",
          geography   = "state",
          year        = yy,
          survey      = "acs1",
          geometry    = FALSE,
          cache_table = TRUE
        ) |>
          mutate(year = yy) |>
          select(
            state      = NAME,
            population = estimate,
            year
          )
        
        # Cache the cleaned result to disk before returning
        write_csv(state_pop_raw, year_cache_file)
        state_pop_raw
        
      }) |> bind_rows()
      
      POP_ALL
    },
    
    ##TASK 4: COMBINED STATE TO STATE FLOWS
    # --- State-to-State Migration: combines 2024 and 2023 ACS-1 flows ---
    state_to_state_migration = {
      
      ##TASK 2: 2024 STATE TO STATE FLOWS
      # -- Internal function: 2024 flows and stationary population --
      state_to_state_migration_2024 <- function() {
        
        migration_url_2024  <- "https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2024/state-to-state-migration/State_to_State_Migration_Table_2024_T13.xlsx"
        migration_file_2024 <- file.path(data_dir, "State_to_State_Migration_Table_2024_T13.xlsx")
        
        if (!file.exists(migration_file_2024)) {
          message("Downloading 2024 state-to-state migration file...")
          download.file(migration_url_2024, destfile = migration_file_2024, quiet = TRUE, mode = "wb")
        } else {
          message("Loading cached 2024 state-to-state migration file...")
        }
        
        # Table sheet: long format, col A = current state, col B = state 1yr ago, col C = estimate
        # "X" = same-state non-movers (filtered out), "N" = suppressed values (replaced with 0)
        flows_2024 <- read_excel(
          migration_file_2024,
          sheet     = "Table",
          skip      = 8,
          col_names = c("state_current", "state_1y", "population", "moe")
        ) |>
          select(state_current, state_1y, population) |>
          filter(!is.na(state_current), population != "X") |>
          mutate(
            population    = replace(population, population == "N", "0"),
            population    = as.integer(population),
            state_current = expand_state_name(state_current),
            state_1y      = expand_state_name(state_1y),
            year_current  = 2024L,
            year_1y       = 2023L
          ) |>
          filter(state_1y != "Foreign country") |>
          select(state_current, state_1y, population, year_current, year_1y)
        
        # Supplemental sheet: stationary population (did not move)
        # skip = 10 to bypass headers, empty row, and US aggregate row
        # col A = state name, col B = total population; sheet has 11 columns total
        stationary_2024 <- read_excel(
          migration_file_2024,
          sheet     = "Supplemental - Current Res",
          skip      = 10,
          col_names = c("state_current", "population", "moe_1", "est_2", "moe_2",
                        "est_3", "moe_3", "est_4", "moe_4", "est_5", "moe_5")
        ) |>
          select(state_current, population) |>
          filter(!is.na(state_current), !is.na(population)) |>
          mutate(
            population   = as.integer(population),
            state_1y     = state_current,
            year_current = 2024L,
            year_1y      = 2023L
          ) |>
          select(state_current, state_1y, population, year_current, year_1y)
        
        bind_rows(flows_2024, stationary_2024)
      }
      
      ##TASK 3: 2023 STATE TO STATE MIGRATION
      # -- Internal function: 2023 flows and stationary population --
      state_to_state_migration_2023 <- function() {
        
        migration_url_2023  <- "https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2023/state-to-state-migration/State_to_State_Migration_Table_2023_T13.xlsx"
        migration_file_2023 <- file.path(data_dir, "State_to_State_Migration_Table_2023_T13.xlsx")
        
        if (!file.exists(migration_file_2023)) {
          message("Downloading 2023 state-to-state migration file...")
          download.file(migration_url_2023, destfile = migration_file_2023, quiet = TRUE, mode = "wb")
        } else {
          message("Loading cached 2023 state-to-state migration file...")
        }
        
        # tool_input sheet: hidden sheet, long format, col A = state_1y, col B = state_current
        # skip = 1 to skip header row; "N/A" = same-state non-movers (filtered out)
        # "N" = suppressed values (replaced with 0)
        flows_2023 <- read_excel(
          migration_file_2023,
          sheet     = "tool_input",
          skip      = 1,
          col_names = c("state_1y", "state_current", "population", "moe",
                        "blank", "state", "in_est", "in_moe", "out_est", "out_moe",
                        "blank2", "state_names")
        ) |>
          select(state_1y, state_current, population) |>
          filter(!is.na(state_1y), population != "N/A") |>
          mutate(
            population    = replace(population, population == "N", "0"),
            population    = as.integer(population),
            # trim whitespace - DC has trailing space in source data
            state_current = str_trim(state_current),
            state_1y      = str_trim(state_1y),
            year_current  = 2023L,
            year_1y       = 2022L
          ) |>
          filter(state_1y != "Foreign country") |>
          select(state_current, state_1y, population, year_current, year_1y)
        
        # Stationary population from Table sheet
        # range A9:D74 reads only cols A-D starting after 8 header rows
        # col A = state name, col D = same residence estimate
        stationary_2023 <- read_excel(
          migration_file_2023,
          sheet     = "Table",
          range     = "A9:D74",
          col_names = c("state_current", "pop_est", "pop_moe", "population")
        ) |>
          select(state_current, population) |>
          filter(!is.na(state_current), !is.na(population),
                 !state_current %in% c("United States2", "Current residence in")) |>
          mutate(
            population   = as.integer(population),
            state_1y     = state_current,
            year_current = 2023L,
            year_1y      = 2022L
          ) |>
          select(state_current, state_1y, population, year_current, year_1y)
        
        bind_rows(flows_2023, stationary_2023)
      }
      
      # Combine both years into one tidy data frame
      bind_rows(state_to_state_migration_2024(), state_to_state_migration_2023())
    }
  )
}

STATE_POP                <- acquire_acs_data("population")
STATE_TO_STATE_MIGRATION <- acquire_acs_data("state_to_state_migration")


```

*Note:* My data frame will have more than 5000 rows because I've taken District of Columbia (DC), Puerto Rico and U.S. Island Areas into consideration and not just the 50 mainland(-ish) states. 

Migrants from Puerto Rico have an automatic right to vote and we can represent significant voting blocks in states such as New York, Connecticut and Florida. Same goes for U.S. Island Areas.  
Excluded migration flows of people from foreign countries since it is not a guarantee that they will be able to vote due to legal status. 


Checking if Data is correct for Task #1

```{r initial-data-check, eval=FALSE}

dim(STATE_POP)
glimpse(STATE_POP)
head(STATE_POP)

#should be 9 years
distinct(STATE_POP, year)

#should be 52 states
STATE_POP |> count(year)

#should be california
STATE_POP |> group_by(year) |> slice_max(population)

#should be wyoming
STATE_POP |> group_by(year) |> slice_min(population)

#should increase
STATE_POP |> group_by(year) |> summarize(total_population = sum(population))
STATE_POP |> filter(state == "California", year == 2023)
STATE_POP |> summarize(across(everything(), ~sum(is.na(.))))
STATE_POP |> filter(population < 100000 | population > 50000000)
```



Remember that code blocks look like this: 

```{r}
# Your code goes in chunks like this
library(tidyverse) # You will want this line for almost all MPs
x <- 1 + 2 + 3
```

and you can then print variables from chunks: 

```{r}
x
```

or inline, like this: $x$ is `r x`.

## Exploratory Analysis

Code related to Exploratory Data Analysis goes here. This may include
exploratory graphics, instructor-provided exploratory questions, or other
similar elements. 

## Final Insights and Deliverable

Code related to the final deliverable of the assignment goes here. 


::: {.callout-note title="AI Usage Statement" collapse="true"}

Claude was used to help write the code in this project and brainstorm final deliverable ideas, but all non-code text was generated without the use of any Generative AI tools.

:::

------------------------------------------------------------------------

This work ©2025 by Evelyn Rodriguez was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #03](https://michael-weylandt.com/STA9750/mini/mini03.html)
