---
title: "Mini-Project 03 - Who Goes There? US Internal Migration and Implications for Congressional Reapportionment"
author: "Evelyn Rodriguez"
editor:
    mode: source
format:
    html:
        toc: true
        code-fold: true
        code-summary: "Display Code"
        code-overflow: wrap
        code-block-bg: true
        code-block-border-left: "#6610f2"
        highlight-style: "a11y"
---

```{r setup, include=FALSE}

library(tidyverse)
library(glue)
library(readxl)
library(httr2)
library(skimr)
library(dplyr)
library(scales)
library(gt)

```

## Introduction

A suitable introduction goes here. In this mini-project, I intend to demonstrate
my skills at Data Import/API Usage/File Parsing in an analysis of Census Data/Migration Flows data.

## Data Acquisition and Preparation

Code related to Data Acquisition and Preparation goes here.

Remember that code blocks look like this: 

```{r}
# Your code goes in chunks like this
library(tidyverse) # You will want this line for almost all MPs
x <- 1 + 2 + 3
```

and you can then print variables from chunks: 

```{r}
x
```

or inline, like this: $x$ is `r x`.

## Exploratory Analysis

Code related to Exploratory Data Analysis goes here. This may include
exploratory graphics, instructor-provided exploratory questions, or other
similar elements. 

## Final Insights and Deliverable

Code related to the final deliverable of the assignment goes here. 


::: {.callout-note title="AI Usage Statement" collapse="true"}

Claude was used to help write the code in this project and brainstorm final deliverable ideas, but all non-code text was generated without the use of any Generative AI tools.

:::

------------------------------------------------------------------------

This work ©2025 by er0driguez was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #03](https://michael-weylandt.com/STA9750/mini/mini03.html)
